---
title: "Credit Risk Data Quality Simulation"
output:
  html_document: default
  pdf_document: default
date: "2026-03-02" 
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Part 0: Helper functions and setup

```{r, warning=FALSE, message=FALSE}
source("01_functions.R")

required_packages <- c("ggplot2", "dplyr", "tidyr", "mice", "xgboost", "DiagrammeR", "ggrepel")

for (pkg in required_packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg)
  }
  library(pkg, character.only = TRUE)
}

```

# **Stage 1: Creating the "Perfect World" with the DGP**

### Overview

------------------------------------------------------------------------

##### **Phase 1. Generating features (predictors) for the business world**

Have to define both the variables and their distributions /
dependencies.

1: **Root variables**

a)  **firm age**: right skewed so using gamma distribution
b)  **sector**: three sectors with proportions and risk properties based
    off real German practitioner data.
    -   Sector 1: Services (Dienstleistungen) at 57% of the sample.
        Least risky with 2% default rate
    -   Sector 2: Hospitality (Gastro), 19% of the sample. Most risky,
        3.6% default rate
    -   Sector 3: Manufacturing (Verarbeitendes Gewerbe), 24% of the
        sample. Medium risk, 2.3% default rate

2: **Chained regression dependencies**

Each variable is determined by the one in front of it -- older firms
tend to be larger, larger firms tend to have lower leverage, etc., plus
random noise.

c)  **log_assets**: proxy for firm size, dependent on age and sector
d)  **debt to equity**: higher leverage is more dangerous -- dependent
    on how big the firm is and how old it is
e)  **interest_cov**: proxy for cash flow
f)  **Crefo score**. (see below) Dependent on most things above.
    Sometimes gets lagged (not updated every year)

------------------------------------------------------------------------

##### **2. Choose the true coefficients that define the world**

This is the "ground truth". All are standardized (change in log-odds per 1 SD increase)

- **beta0 = -3.9** | this is the intercept -- should be whatever the log odds
of the default rate is 
- **beta_log_assets = -0.2** | larger firms less likely to default 
- **beta_debt_to_equity = 0.4** | higher leverage = higher risk
- **beta_interest_cov = -0.3** | higher coverage of interest expenses = lower risk 
- **beta_sector2 = 0.4** | gastro is the riskiest sector 
- **beta_sector3 = 0.18** | manufacturing slightly riskier 
- **beta_firm_age = -0.1** | older firms slightly safer 
- **beta_crefo = 0.3** | higher Crefo score → higher default risk

------------------------------------------------------------------------

##### **3. Create panel data with time series component**

**Inner loop**: creating firms that either stay the same over time, get
worse, or improve... (drift rates); default or not during their 3-7 year
loan periods. 

Crefo will sometimes refresh itself, sometimes remaining stale

**Outer loop** creates lots of these firms. Then the rows are bound together

-   the "linpred" -- weighted sum of the predictors. this transforms to
    logistic regression
    -   computes the log-odds of default for each borrower
-   pd_true: then after: the logistic transformation so we get the true
    PDs and not log-odds

**Next: Draw observed defaults.**
For each borrower, draw a default with pd_true value from before (weighted coin 
flip) -- borrowers can have same PDs but 2 different outcomes
(default or not)

------------------------------------------------------------------------

##### **4. Reporting and returning a list**

-   Data has both raw and scaled versions of every variable
-   Scaling = the mean and SD of every predictor so you can apply
    consistently to the impaired datasets
-   Also saving the bounds for winsorization and plausibility (related
    to data preparation and FRS scores)

True betas = ground truth coefficients

------------------------------------------------------------------------

**A quick note on Crefo score (Crefo Bonitätsindex)**

Scale: 100 (excellent) to 600 (insolvent) — higher = worse
creditworthiness. Generated as a noisy composite of the same financial
signals the bureau uses: leverage, coverage, firm size, and sector.
Idiosyncratic noise (sd = 25) reflects bureau information not in our DGP
(payment history with other creditors, legal registry data, etc.).

Sector penalties: gastro (40) and manufacturing (20) reflect higher
historical default rates in those sectors per practitioner datasets.

------------------------------------------------------------------------
------------------------------------------------------------------------

### Data Generating Process DAG

**1) Simple DAG for the five indicator variables**

Color code: 

- Gray `firm_age` : root variable drawn from a distribution,
not derived from anything else 
- Green `sector` : root variable like age , but continuous 
- Blue `log_assets`, `debt_to_equity`,
`interest_cov` : financial predictors derived from root vars and each
other 
- Orange `crefo`: Creditreform separate since it's a derived
external rating that depends on all of the above (except age)

```{r}

# ── Simple DAG: Origination-level variable dependencies ──────────────

grViz("
digraph DGP_simple {

  graph [rankdir = TB, fontname = 'Helvetica', bgcolor = 'transparent',
         nodesep = 0.8, ranksep = 0.7, margin = '0.2,0.2']
  node  [fontname = 'Helvetica', fontsize = 11, style = filled,
         shape = box, penwidth = 0.6]
  edge  [color = '#888888', penwidth = 0.7, arrowsize = 0.6]

  # ── Root variables ─────────────────────────────────────────────
  firm_age [label = 'firm_age',  fillcolor = '#F1EFE8', color = '#B4B2A9']
  sector   [label = 'sector',    fillcolor = '#E1F5EE', color = '#5DCAA5']

  # ── Derived variables ──────────────────────────────────────────
  log_assets     [label = 'log_assets',     fillcolor = '#E6F1FB', color = '#85B7EB']
  debt_to_equity [label = 'debt_to_equity', fillcolor = '#E6F1FB', color = '#85B7EB']
  interest_cov   [label = 'interest_cov',   fillcolor = '#E6F1FB', color = '#85B7EB']
  crefo          [label = 'crefo',          fillcolor = '#FAECE7', color = '#F0997B']

  # ── Edges ──────────────────────────────────────────────────────
  firm_age -> log_assets
  sector   -> log_assets
  firm_age -> debt_to_equity
  log_assets -> debt_to_equity
  debt_to_equity -> interest_cov
  log_assets -> interest_cov
  log_assets -> crefo
  debt_to_equity -> crefo
  interest_cov -> crefo
  sector -> crefo

  # ── Layout ─────────────────────────────────────────────────────
  {rank = same; firm_age; sector}
  {rank = same; log_assets; debt_to_equity; interest_cov}
}
")

```

**2) Entire DGP**

- Phase one: (left) variable and dependency declaration 
- Phase two: (right) panel drift
- Phase three: linear predictor combining both of the above phases 

```{r}
# ── Detailed DAG of the Data Generating Process ─────────────────────────────

grViz("
digraph DGP {

  # ── Global settings ──────────────────────────────────────────────
  graph [rankdir = TB, fontname = 'Helvetica', bgcolor = 'transparent',
         nodesep = 0.6, ranksep = 0.7, margin = '0.2,0.2']
  node  [fontname = 'Helvetica', fontsize = 11, style = filled,
         shape = box, penwidth = 0.6]
  edge  [color = '#888888', penwidth = 0.7, arrowsize = 0.6]

  # ── Root variables (drawn once per firm) ────────────────────────
  firm_age   [label = 'firm_age',   fillcolor = '#F1EFE8', color = '#B4B2A9']
  sector     [label = 'sector',     fillcolor = '#E1F5EE', color = '#5DCAA5']
  trajectory [label = 'trajectory', fillcolor = '#EEEDFE', color = '#AFA9EC']

  # ── Derived at origination ─────────────────────────────────────
  log_assets     [label = 'log_assets',     fillcolor = '#E6F1FB', color = '#85B7EB']
  debt_to_equity [label = 'debt_to_equity', fillcolor = '#E6F1FB', color = '#85B7EB']
  interest_cov   [label = 'interest_cov',   fillcolor = '#E6F1FB', color = '#85B7EB']
  crefo          [label = 'crefo',          fillcolor = '#FAECE7', color = '#F0997B']

  # ── Panel drift (t > 1) ───────────────────────────────────────
  pd_curr      [label = 'pd_curr\\n(drift weight)',
                fillcolor = '#F1EFE8', color = '#B4B2A9', shape = ellipse]
  annual_update [label = 'Annual updates\\nlog_assets, d/e, int_cov',
                fillcolor = '#E6F1FB', color = '#85B7EB']
  crefo_refresh [label = 'crefo refresh\\n(p = 0.5)',
                fillcolor = '#FAECE7', color = '#F0997B']

  # ── Linear predictor → default ────────────────────────────────
  gastro_trend [label = 'gastro_trend\\nsector_2 × (year − 2015)',
               fillcolor = '#E1F5EE', color = '#5DCAA5']
  linpred      [label = 'linpred\\nall _z predictors + sector dummies',
               fillcolor = '#FAEEDA', color = '#FAC775']
  default      [label = 'default\\nBernoulli(pd_true)',
               fillcolor = '#FCEBEB', color = '#F09595']

  # ── Phase 1 edges: origination dependencies ────────────────────
  firm_age -> log_assets
  sector   -> log_assets     [label = '  ±0.3', fontsize = 9, fontcolor = '#999999']
  firm_age -> debt_to_equity [label = '  −0.02·z', fontsize = 9, fontcolor = '#999999']
  log_assets -> debt_to_equity [label = '  −0.1·z', fontsize = 9, fontcolor = '#999999']
  debt_to_equity -> interest_cov [label = '  −0.4·z', fontsize = 9, fontcolor = '#999999']
  log_assets -> interest_cov [label = '  +0.3·z', fontsize = 9, fontcolor = '#999999']

  # Crefo at origination: depends on all financials + sector
  log_assets -> crefo
  debt_to_equity -> crefo
  interest_cov -> crefo
  sector -> crefo

  # ── Phase 2 edges: drift mechanism ─────────────────────────────
  trajectory -> annual_update [label = '  drift direction', fontsize = 9, fontcolor = '#7F77DD']
  pd_curr -> annual_update    [label = '  drift intensity\\n  (deteriorating only)', fontsize = 9, fontcolor = '#999999']
  annual_update -> crefo_refresh

  # ── Phase 3 edges: linpred ─────────────────────────────────────
  annual_update -> linpred
  crefo -> linpred
  crefo_refresh -> linpred [style = dashed]
  sector -> linpred
  firm_age -> linpred       [label = '  ages +1/yr', fontsize = 9, fontcolor = '#999999']
  gastro_trend -> linpred
  sector -> gastro_trend

  linpred -> default

  # ── Layout hints ───────────────────────────────────────────────
  {rank = same; firm_age; sector; trajectory}
  {rank = same; log_assets; debt_to_equity; interest_cov}
  {rank = same; pd_curr; annual_update}
  {rank = same; gastro_trend; linpred}
}
")

```

## Simulation

This creates one synthetic dataset with 5000 businesses / around 20k
rows, and tests the default rate -- should be around 2.4%.

```{r}

result <- simulate_clean_data_panel(N = 50000) 
# running with high sample size as this is the showcase document
# Monte Carlo runs with less sample but more repetitions

###################################################################

dat    <- result$data

###################################################################

# For later: record what our PD formula is
pd_formula <- default ~ log_assets_z + debt_to_equity_z + interest_cov_z +
                             firm_age_z + crefo_z + sector_2 + sector_3


```

## Checkpoint: Model Fit in the Perfect World

Fitting a logistic regression on clean data: 

- Even though we know the DGP and all that influence it, basic GLM is unable 
to perfectly recover it -- even on non-impaired data!
- This is a finding in itself: though regulators love the logistic model for 
its transparency, it is a little too simple for the complications of panel 
data. We are breaking the assumption of independence and add heterogeneity 
with firm trajectories, setting off the estimated betas
- Beta recovery on the clean data does not improve with larger sample sizes,
or accounting for random effects for time series (glmnet)



```{r}
## Model fit

fitcheck_scaled <- glm(default ~ log_assets_z + debt_to_equity_z + interest_cov_z + 
                         sector_2 + sector_3 + firm_age_z + crefo_z,
                       data = dat, family = binomial())
summary(fitcheck_scaled)

# Compare estimated vs true betas -- this is your coefficient recovery check
true_betas <- result$true_betas

recovery <- data.frame(
  true      = true_betas,
  estimated = coef(fitcheck_scaled),
  diff      = coef(fitcheck_scaled) - true_betas
)
print(round(recovery, 4))

```

Sidequest:

## Introducing Feature Reliability Score on clean data

**Feature Reliability Score (FRS)** testing from Igl & Gruber's Handbuch
Datenqualität (2025). A customizable data quality measure per indicator
variable -- in our case, looking at 1) Completeness, 2) Plausibility, 3)
Outlier Score, and 4) Cross-Field Consistency for our four continuous
predictors. FRS takes the geometric mean for the listed components.

Because this is on clean data, we should expect the scores to be near 1
(perfect quality). We get marked down slightly in Plausibility and
Outliers because the FRS function is always looking to penalize data at the
tails (it does not take into account that we are in the perfect world).

Crefo has different expectations since it is purchased from an external data
provider and by default bounded at [100, 600] so it receives the only 
"perfect score".

```{r}

frs_clean <- compute_frs(dat,
                         winsor_bounds       = result$winsor_bounds,
                         plausibility_bounds = result$plausibility_bounds)
print_frs(frs_clean)

```

------------------------------------------------------------------------

# **Stage 2: Data Impairment**

### MISSINGNESS

#### **Impairment 1: MCAR**

MCAR: missingness applied independently to each cell at random.

Every cell has the same probability of being missing regardless of any
other value in the dataset — no row-level or variable-level dependence.
This is implemented via independent rbinom draws per variable, producing
a random scatter of missingness across the data

Variables included: financial ratios and firm_age. 

Sector and Crefo are
excluded: sector is a simple categorical entry unlikely to be missing;
Crefo missingness is modelled separately under the MAR condition.

Mild: ~8% missing per cell; Severe: ~28% missing per cell


Running the MCAR function to create two levels impaired data:

```{r}

cat("Mild MCAR impairment:\n")
dat_mcar_mild   <- impair_mcar(dat, severity = "mild")
cat("\nSevere MCAR impairment:\n")
dat_mcar_severe <- impair_mcar(dat, severity = "severe")


```
------------------------------------------------------------------------

#### **Impairment 2: MAR**

MAR: missingness in crefo depends on firm_age (fully observed)

Mechanism: younger firms are less likely to have a Creditreform score
because they lack sufficient credit history for the bureau to have
issued a rating.

This is MAR because the missingness predictor (firm_age) is often
observed and is a different variable from the one with missing data
(crefo) — consistent with Little & Rubin (2002) and Zhang (2023, p.
116). The logistic regression method is used for realistic gradual
probability change (Zhang, 2023, p. 116). Domain justification:
confirmed with company supervisors — new market entrants frequently lack
Creditreform coverage in German B2B lending.


Creating the impaired datasets:

```{r}

cat("Mild MAR impairment for Crefo:\n")
dat_mar_mild   <- impair_mar(dat, severity = "mild")

cat("\nSevere MAR impairment for Crefo:\n")
dat_mar_severe <- impair_mar(dat, severity = "severe")

```

------------------------------------------------------------------------

#### **Impairment 3: MNAR**

MNAR: missingness in interest_cov depends on interest_cov itself.

Mechanism: firms with poor interest coverage strategically omit
reporting. This is MNAR because missingness depends on the unobserved
value of the variable itself (Little & Rubin, 2002; Zhang, 2023, p.
117).

Only allowing interest coverage to become impaired for a cleaner causal
story that is easier to validate (if we see coefficient distortion, we
know WHICH missing variable caused it, versus having multiple missing
variables x their interactions).

In real life: plausibly the most likely to be affected by MNAR because
it requires current earnings data that troubled firms are more likely to
delay or withhold.

Persistence: once a firm's interest_cov goes missing in a given year, it
remains missing for all subsequent years of that firm's life. This
reflects real reporting behavior — a firm that stops disclosing bad
financials does not voluntarily resume disclosure.

Intercepts tuned to achieve overall missing rates comparable to MCAR/MAR
after persistence propagation: 
Mild: ~8% of firm-years, 
Severe: ~28%of firm-years

Note: because persistence amplifies the initial trigger rate, the
per-year trigger probability is set lower than the target.


Creating the impaired datasets:

```{r}
cat("Mild MNAR impairment on Interest Coverage:\n")
dat_mnar_mild   <- impair_mnar(dat, severity = "mild")

cat("\nSevere MNAR impairment on Interest Coverage:\n")
dat_mnar_severe <- impair_mnar(dat, severity = "severe")

```

------------------------------------------------------------------------

#### **Impairment 4: Measurement Noise**

Measurement noise: Gaussian noise added to continuous financial
predictors, creating a less-precise dataset.

Only continuous predictors are corrupted (except firm_age and crefo).

Severity defined as % of each variable's clean SD 

Mild: SD_noise = +/-30% of clean SD, 
Severe: SD_noise = +/-100% of clean SD

Creating the impaired datasets:

```{r}
cat("Mild noise impairment:\n")
dat_noise_mild   <- impair_noise(dat, result$scaling, severity = "mild")

cat("\nSevere noise impairment:\n")
dat_noise_severe <- impair_noise(dat, result$scaling, severity = "severe")
```

------------------------------------------------------------------------

#### **Impairment 5: Implausible Values**

Implausible values: inject a mix of domain-invalid, extreme outlier, and
subtly wrong values into financial predictors.

Unlike the missingness impairments, implausible values are designed to
be heterogeneous — mixing obvious violations with harder-to-detect
distortions. This reflects real-world data quality issues: decimal
shifts, unit confusion, stale data, and system migration artefacts do
not produce uniform error patterns.

Three error types per variable: 

1. **Domain-invalid**: clearly impossible
values (negative age, etc.) 

2. **Extreme outlier**: technically possible but
far outside normal range 

3. **Subtle**: within plausible range but wrong
(decimal shifts, swaps [numerator / denominator, or age in months not
years])

Severity controls the total % of rows corrupted per variable.

Creating the impaired datasets:

```{r}
cat("Mild implausible impairment:\n")
dat_impl_mild   <- impair_implausible(dat, result$scaling, severity = "mild")

cat("\nSevere implausible impairment:\n")
dat_impl_severe <- impair_implausible(dat, result$scaling, severity = "severe")

```



------------------------------------------------------------------------

**Sidequest FRS part 2:**

### Feature reliability score on impaired data

Here we want to look at both the overall FRS score (aggregate, geometric
mean) along with individual FRS components for mild and severe data before we 
take any steps to prepare.

Handbuch Datenqualität chapter 8: frames it as the first thing data scientists
should do when they get a dataset, then use the results to decide what to 
do next. Then can decide which varibales are prioritized for cleaning or
dropped from modeling altogether.

```{r}

# Preparation Block

impaired_list <- list(
  mcar_mild    = dat_mcar_mild,
  mcar_severe  = dat_mcar_severe,
  mar_mild     = dat_mar_mild,
  mar_severe   = dat_mar_severe,
  mnar_mild    = dat_mnar_mild,
  mnar_severe  = dat_mnar_severe,
  noise_mild   = dat_noise_mild,
  noise_severe = dat_noise_severe,
  impl_mild    = dat_impl_mild,
  impl_severe  = dat_impl_severe
)

frs_all <- compute_frs_all(dat, impaired_list,
                           winsor_bounds       = result$winsor_bounds,
                           plausibility_bounds = result$plausibility_bounds)

frs_summary <- summarize_frs(frs_all)

```

**FRS Summary Tables**

| Score       | Bewertung                 |
|-------------|---------------------------|
| > 95%       | sehr hohe Datenqualität   |
| 90–95%      | gute Datenqualität        |
| 80–90%      | mittlere Datenqualität    |
| < 80%       | kritische Datenqualität   |

```{r}
# Quick look by impairment
frs_summary$by_condition

# And by variable (more as Igl & Gruber intended)
frs_summary$by_variable
```

**FRS visualization**

```{r}
# Heatmap: FRS by type x severity
ggplot(frs_summary$by_condition, aes(x = severity, y = type, fill = frs)) +
  geom_tile() +
  geom_text(aes(label = round(frs, 3)), size = 3) +
  scale_fill_gradient(low = "red", high = "green") +
  labs(title = "Overall FRS by Impairment Condition")

# Bar chart: which variables degrade most per mechanism
ggplot(frs_summary$by_variable[frs_summary$by_variable$type != "none", ],
       aes(x = var, y = frs, fill = type)) +
  geom_col(position = "dodge") +
  labs(title = "FRS by Variable and Impairment Type") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```

**Notes on FRS results pre-preparation:** 

- Overall FRS scores hide the actual damage being done, because some fields
like cross-field consistency and plausibility in MCAR get nearly a perfect score.
That's why the worst dataset with 28% missings (MCAR severe) is listed as 
"good data quality" (0.925 FRS)
  --> likely makes more sense with real-world data instead of one-at-a-time 
  impairments
- Noise and implausible values don't show damage like missingness
  --> Why? FRS designed to capture structural data problems: inconsistent 
  relationships, domain violations, obvious extremes.



------------------------------------------------------------------------

Last step before modeling:

## **Stage 3: Data Preparation**


 Preparation function: sits between impairment and model fitting.
 Handles the following tasks:
   1. Identify extreme outliers and converts them to missings
   2. Winsorize data (two or one tailed, depending on variables)
   3. Impute NA values introduced by missingness 
   4. Recompute _z columns using clean scaling parameters

 The method argument controls the imputation strategy:

   **"ersatzwert"** — RR univariate Ersatzwert imputation (default).
              For each variable independently: fits a univariate logistic
              regression (default ~ x) on non-missing cases, computes the
              observed default rate among missing cases (AR_MV), and solves
              for the x-value that would produce that default rate:
 
              Ersatzwert = (logit(AR_MV) - β₀) / β₁

              All missing rows receive the same replacement value per variable.
              Implements the RR development-phase imputation methodology.

   **"median"** — univariate median imputation (PD-neutral baseline;
              maps to ~z=0 after standardisation, adding approximately
              zero to the linear predictor).

   **"mice"**  — multivariate imputation by chained equations (van Buuren
              & Groothuis-Oudshoorn, 2011). Uses predictive mean matching
              (pmm) as the default method, which preserves the distribution
              shape and avoids impossible values. Single imputation (m=1)
              is used because Monte Carlo replications serve the role of
              repeated draws. Requires the mice package.

 This function is applied identically regardless of the missingness
 mechanism (MCAR/MAR/MNAR), reflecting practitioner norms where the
 true mechanism is unknown and all missingness is treated the same.



Fixing the data with the practitioner default method of Ersatzwert imputation

```{r}

# ── Prepare all impaired datasets for modeling ─────────────────────────────

# Check domain boundaries, winsorize, impute NAs using the default Ersatzwert technique + recompute z-scores

dat_mcar_mild_prepped   <- prepare_for_modeling(dat_mcar_mild, result$scaling, method = "ersatzwert", model_formula = pd_formula)
dat_mcar_severe_prepped <- prepare_for_modeling(dat_mcar_severe, result$scaling, method = "ersatzwert", model_formula = pd_formula)
dat_mar_mild_prepped    <- prepare_for_modeling(dat_mar_mild, result$scaling, method = "ersatzwert", model_formula = pd_formula)
dat_mar_severe_prepped  <- prepare_for_modeling(dat_mar_severe, result$scaling, method = "ersatzwert", model_formula = pd_formula)
dat_mnar_mild_prepped   <- prepare_for_modeling(dat_mnar_mild, result$scaling, method = "ersatzwert", model_formula = pd_formula)
dat_mnar_severe_prepped <- prepare_for_modeling(dat_mnar_severe, result$scaling, method = "ersatzwert", model_formula = pd_formula)
dat_noise_mild_prepped   <- prepare_for_modeling(dat_noise_mild, result$scaling, method = "ersatzwert", model_formula = pd_formula)
dat_noise_severe_prepped <- prepare_for_modeling(dat_noise_severe, result$scaling, method = "ersatzwert", model_formula = pd_formula)
dat_impl_mild_prepped    <- prepare_for_modeling(dat_impl_mild, result$scaling, method = "ersatzwert", model_formula = pd_formula)
dat_impl_severe_prepped  <- prepare_for_modeling(dat_impl_severe, result$scaling, method = "ersatzwert", model_formula = pd_formula)



```

Moving into a list for later:

```{r}
prep_ersatz_list <- list(
  mcar_mild    = dat_mcar_mild_prepped,
  mcar_severe  = dat_mcar_severe_prepped,
  mar_mild     = dat_mar_mild_prepped,
  mar_severe   = dat_mar_severe_prepped,
  mnar_mild    = dat_mnar_mild_prepped,
  mnar_severe  = dat_mnar_severe_prepped,
  noise_mild   = dat_noise_mild_prepped,
  noise_severe = dat_noise_severe_prepped,
  impl_mild    = dat_impl_mild_prepped,
  impl_severe  = dat_impl_severe_prepped
)
```

**Sidequest: what happens if we use median or mice?**

Sidequest 1: Imputation Test


```{r}
source("02_sidequest1_imputation_x_models.R")
# Now median_full and mice_full exist with all 10 conditions
```

**Notes on Imputation Method**:

- Despite more computational power needed, not seeing a large benefit to 
using MICE over Ersatzwert or Median after looking at beta recovery
- 


**Sidequest 2: can FRS pick up on data quality issues before modeling?**

Comparing FRS across imputation methods for
missingness

```{r}
frs_compare <- compare_frs_imputation(
  dat, impaired_list, prep_ersatz_list,
  median_full, mice_full,
  result$winsor_bounds, result$plausibility_bounds
)
frs_compare


```

And **visualizations: FRS Breakdown**

```{r}


# ---- Plot 1: Overall FRS by mechanism x severity x imputation method ------
# Collapse across variables first
frs_summary <- aggregate(
  frs ~ type + severity + stage,
  data = frs_compare[frs_compare$var != "MEAN" & frs_compare$type != "none", ],
  FUN  = mean
)

ggplot(frs_summary, aes(x = stage, y = frs, fill = stage)) +
  geom_col() +
  geom_text(aes(label = round(frs, 3)), vjust = -0.5, size = 2.5) +
  facet_grid(severity ~ type) +
  coord_cartesian(ylim = c(0.90, 1.005)) +
  scale_fill_manual(values = c(clean      = "grey70",
                               impaired   = "#e74c3c",
                               ersatzwert = "#3498db",
                               median     = "#f39c12",
                               mice       = "#2ecc71")) +
  labs(title    = "FRS Recovery by Imputation Method",
       subtitle = "How well does each method restore data quality?",
       y = "Mean FRS", x = NULL) +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))

# ---- Plot 2: Component breakdown for a deeper look -----------------------
# Pivot to long format for faceting by component
frs_long <- pivot_longer(
  frs_compare[frs_compare$var != "MEAN" & frs_compare$type != "none", ],
  cols      = c(completeness, plausibility, outlier, consistency),
  names_to  = "component",
  values_to = "score"
)

frs_comp_summary <- aggregate(
  score ~ type + severity + stage + component,
  data = frs_long,
  FUN  = mean
)

ggplot(frs_comp_summary[frs_comp_summary$severity == "severe", ],
       aes(x = stage, y = score, fill = stage)) +
  geom_col() +
  facet_grid(component ~ type) +
  scale_fill_manual(values = c(clean      = "grey70",
                               impaired   = "#e74c3c",
                               ersatzwert = "#3498db",
                               median     = "#f39c12",
                               mice       = "#2ecc71")) +
  labs(title    = "FRS Components: Severe Impairment",
       subtitle = "Completeness should recover; watch plausibility and outliers",
       y = "Score", x = NULL) +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))



```

**FRS visualizations per variable:**

```{r}
# Filter to severe + drop the aggregate MEAN row
frs_var_severe <- frs_compare %>%
  filter(severity == "severe", var != "MEAN")

# Reshape: one column per stage for the heatmap facets
# frs_compare should already have columns: type, severity, var, stage, frs
# (plus the component columns)

ggplot(frs_var_severe,
       aes(x = type, y = var, fill = frs)) +
  geom_tile(color = "white", linewidth = 0.4) +
  geom_text(aes(label = sprintf("%.3f", frs)), size = 2.5) +
  facet_wrap(~ stage, nrow = 1) +
  scale_fill_gradient2(
    low = "#e74c3c", mid = "#f9e79f", high = "#2ecc71",
    midpoint = 0.95, limits = c(0.85, 1),
    oob = scales::squish,          # clamp anything outside range
    name = "FRS"
  ) +
  labs(
    title    = "Per-Variable FRS: Severe Impairment × Imputation Method",
    subtitle = "Red = unreliable | Green = healthy | Read left→right to see recovery",
    x = "Impairment type", y = NULL
  ) +
  theme_minimal() +
  theme(
    axis.text.x  = element_text(angle = 45, hjust = 1),
    strip.text    = element_text(face = "bold"),
    panel.spacing = unit(1, "lines")
  )
```

**FRS Deep Dive**

```{r}
# Pick whichever impairment type you want to drill into
focus_type <- "implausible"   # swap to "MNAR", "noise", etc.

frs_comp_var <- frs_compare %>%
  filter(severity == "severe", type == focus_type, var != "MEAN") %>%
  pivot_longer(
    cols      = c(completeness, plausibility, outlier, consistency),
    names_to  = "component",
    values_to = "score"
  )

frs_comp_var$stage <- factor(frs_comp_var$stage,
                             levels = c("impaired", "ersatzwert", "median", "mice"))

ggplot(frs_comp_var,
       aes(x = stage, y = score, fill = stage)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = sprintf("%.3f", score)), vjust = -0.3, size = 2.2) +
  facet_grid(component ~ var) +
  coord_cartesian(ylim = c(0.80, 1.02)) +
  scale_fill_manual(values = c(impaired   = "#e74c3c",
                               ersatzwert = "#3498db",
                               median     = "#f39c12",
                               mice       = "#2ecc71")) +
  labs(
    title    = paste0("FRS Component Drill-Down: ", focus_type, " (severe)"),
    subtitle = "Which quality dimension degrades, and which method repairs it?",
    x = NULL, y = "Score"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x     = element_text(angle = 45, hjust = 1),
    strip.text       = element_text(face = "bold", size = 8),
    panel.spacing    = unit(0.5, "lines")
  )
```


------------------------------------------------------------------------

# **Stage 4: Modeling**

## 4a Logistic Regression

### Overview

The logistic regression is the workhorse model for PD estimation in
regulatory credit risk (EBA GL 2017/16). RR uses `glm()` with
`family = binomial()` -- the standard maximum-likelihood logit.

The function fits the same formula across three stages:

1.  **Clean** -- the baseline, best-case coefficient recovery
2.  **Impaired** -- damage applied, no preparation. How bad does it get?
3.  **Prepared** -- after winsorizing, imputation, and rescaling. Did
    preparation recover anything?

For the missingness conditions (MCAR, MAR, MNAR), the raw impaired
datasets contain NAs. `glm()` handles this via complete-case analysis --
it silently drops any row with a missing value in a predictor. This is
informative: it shows what the model can recover when it simply discards
incomplete observations, and it provides a natural comparison point for
the prepared version where imputation fills those gaps. For noise and
implausible conditions, there are no NAs, so the raw fit runs on the
full distorted data with no cleanup.

Each fit stores:

-   **The fitted model object** (for later prediction, diagnostics,
    etc.)
-   **Estimated coefficients** alongside the true betas, with absolute
    differences -- this is the coefficient recovery check
-   **Number of observations used** -- important for interpreting the
    complete-case fits where glm drops NA rows
-   **Convergence flag** -- glm can occasionally fail to converge on
    heavily impaired data; we want to know when that happens rather than
    silently carrying a bad fit forward

Running it on all conditions -- raw impaired data (impaired_list) and
ersatzwert-imputed prepared data (prep_ersatz_list):

```{r}

glm_results <- fit_logistic_models(
  clean_data    = dat,
  impaired_list = impaired_list,
  prepared_list = prep_ersatz_list,
  formula       = pd_formula,
  true_betas    = result$true_betas
)

# Quick look: coefficient recovery across all conditions
print(glm_results$coef_recovery)

```

### Coefficient recovery: how far did the betas drift?

A first visual diagnostic: for each impairment condition, how close are
the recovered coefficients to the DGP truth? Three stages are shown
side-by-side: the clean baseline (triangle), raw impaired fits (circle),
and prepared fits (square). The question at each panel is: does the
square move back toward the dashed truth line compared to the circle?

Note on sample sizes: for MCAR/MAR/MNAR impaired fits, glm performs
complete-case analysis (dropping rows with any NA). This means the
impaired models for missingness conditions train on fewer observations
than the prepared versions where imputation restores completeness. The
n_obs column in coef_recovery tracks this.

```{r, fig.width=10, fig.height=7}

# ── Coefficient recovery dot plot: three-stage comparison ──────────────

# Readable ordering for the y-axis
condition_order <- c("clean",
                     "mcar_mild", "mcar_severe",
                     "mar_mild",  "mar_severe",
                     "mnar_mild", "mnar_severe",
                     "noise_mild", "noise_severe",
                     "impl_mild", "impl_severe")

glm_results$coef_recovery$condition <- factor(
  glm_results$coef_recovery$condition,
  levels = condition_order
)

glm_results$coef_recovery$stage <- factor(
  glm_results$coef_recovery$stage,
  levels = c("clean", "impaired", "prepared")
)

ggplot(glm_results$coef_recovery,
       aes(x = estimated, y = condition, color = severity, shape = stage)) +
  geom_point(size = 2.5, position = position_dodge(width = 0.4)) +
  geom_vline(aes(xintercept = true), linetype = "dashed", color = "grey40") +
  facet_wrap(~ term, scales = "free_x", ncol = 4) +
  scale_color_manual(values = c(none = "grey30", mild = "#f39c12", severe = "#e74c3c")) +
  scale_shape_manual(values = c(clean = 17, impaired = 16, prepared = 15)) +
  labs(title    = "Coefficient Recovery: Logistic Regression",
       subtitle = "Dashed line = DGP truth | triangle = clean, circle = impaired, square = prepared",
       x = "Estimated coefficient", y = NULL) +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"))

```

## 4b XGBoost

### Overview

XGBoost is the second model in the simulation grid. Where logistic
regression gives us interpretable coefficients to check against the DGP
truth, XGBoost tests whether a flexible, non-parametric learner is more
or less robust to data quality impairment.

Key structural differences from glm that matter here:

-   **No formula interface.** XGBoost needs a numeric matrix
    (`xgb.DMatrix`) and a separate label vector. We extract the same
    predictor columns used in `pd_formula` and convert manually.
-   **Native NA handling.** XGBoost learns optimal split directions for
    missing values during training. This means the raw impaired fits
    (stage = "impaired") use *all* rows, including those with NAs --
    unlike glm, which drops them via complete-case analysis. This is a
    structurally different response to the same damage and part of the
    model comparison story
-   **No coefficients.** XGBoost produces feature importance
    (gain-based) rather than interpretable betas. We store importance
    tables instead of coefficient recovery, and the real model
    comparison happens downstream at the evaluation stage (predicted
    PDs, discrimination, calibration, grading).
-   **Fixed hyperparameters.** Since the research question is about data
    quality, not model tuning, we fix hyperparameters across all
    conditions. Any performance difference is then attributable to the
    data, not to tuning. The defaults below are conservative and
    standard for binary credit risk: shallow trees, moderate learning
    rate, early stopping to avoid overfitting. 

Running it:

```{r}

xgb_results <- fit_xgboost_models(
  clean_data    = dat,
  impaired_list = impaired_list,
  prepared_list = prep_ersatz_list,
  formula       = pd_formula
)

# Quick look at feature importance across conditions
print(xgb_results$importance)

```

### Feature importance: does impairment reshuffle what the model relies on?

In the clean-data fit, XGBoost should allocate gain roughly in line with
the true effect sizes in the DGP. Impairment can distort this in two
ways: a damaged variable loses signal and drops in importance, or
preparation artifacts (e.g., imputed values clustering at the median)
make a variable artificially easy to split on.

The plot below shows gain-based importance by feature, comparing clean,
impaired, and prepared stages across all conditions.

```{r, fig.width=10, fig.height=7}

# ── Feature importance comparison ──────────────────────────────────────

condition_order <- c("clean",
                     "mcar_mild", "mcar_severe",
                     "mar_mild",  "mar_severe",
                     "mnar_mild", "mnar_severe",
                     "noise_mild", "noise_severe",
                     "impl_mild", "impl_severe")

xgb_results$importance$condition <- factor(
  xgb_results$importance$condition,
  levels = condition_order
)

xgb_results$importance$stage <- factor(
  xgb_results$importance$stage,
  levels = c("clean", "impaired", "prepared")
)

ggplot(xgb_results$importance,
       aes(x = Gain, y = condition, color = severity, shape = stage)) +
  geom_point(size = 2.5, position = position_dodge(width = 0.4)) +
  facet_wrap(~ Feature, scales = "free_x", ncol = 4) +
  scale_color_manual(values = c(none = "grey30", mild = "#f39c12", severe = "#e74c3c")) +
  scale_shape_manual(values = c(clean = 17, impaired = 16, prepared = 15)) +
  labs(title    = "Feature Importance (Gain): XGBoost",
       subtitle = "How does impairment and preparation shift what the model relies on?",
       x = "Gain", y = NULL) +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"))

```

## 4c PD Grading

### Overview

After fitting models, the next step in a real-world PD framework is to
assign each firm to a risk grade. This maps continuous predicted PDs
onto a discrete rating scale, like how Moody's maps default
probabilities to letter grades (Aaa through C).

We use a 10-grade scale (1 = safest, 10 = riskiest) with PD boundaries
derived from Moody's one-year idealized corporate default rates. The
mapping is:

| Grade | Moody's Equivalent | PD Upper Bound |
|-------|--------------------|----------------|
| 1     | Aaa                | 0.001%         |
| 2     | Aa                 | 0.06%          |
| 3     | A                  | 0.11%          |
| 4     | Baa                | 0.30%          |
| 5     | Ba (upper)         | 0.70%          |
| 6     | Ba (lower)         | 1.50%          |
| 7     | B (upper)          | 3.00%          |
| 8     | B (lower)          | 8.00%          |
| 9     | Caa                | 15.00%         |
| 10    | Ca-C / Default     | 100.00%        |

Given the DGP's ~2% average default rate, most firms should cluster in
grades 5-8, with relatively few in the tails. This is realistic for a
B2B/SME portfolio.

The critical design choice: **bucket boundaries are fixed and model-
independent.** The same PD thresholds are applied to all conditions, all
stages, and both model types. This means a "grade 6" carries the same PD
interpretation everywhere, and any shift in grade distributions across
conditions is attributable to the model + data, not to a moving scale.

This also enables **grade migration analysis**: comparing which grade
each firm receives under clean vs. impaired vs. prepared conditions. A
firm graded 5 on clean data but 7 on impaired data has been
misclassified by the damage -- and whether preparation pulls it back
toward 5 is the key question.

Running it for both models -- note that we need to pass matched lists of
models and datasets so that each model scores its own training data:

```{r}

# Build matched dataset lists (same structure as the model keys)
all_datasets <- c(
  list(clean = dat),
  setNames(
    lapply(names(impaired_list), function(nm) impaired_list[[nm]]),
    paste0("imp_", names(impaired_list))
  ),
  setNames(
    lapply(names(prep_ersatz_list), function(nm) prep_ersatz_list[[nm]]),
    paste0("prep_", names(prep_ersatz_list))
  )
)

# Grade all models
glm_grades <- assign_pd_grades(
  models     = glm_results$models,
  datasets   = all_datasets,
  formula    = pd_formula,
  model_type = "glm"
)

xgb_grades <- assign_pd_grades(
  models     = xgb_results$models,
  datasets   = all_datasets,
  formula    = pd_formula,
  model_type = "xgboost"
)

```

### Grade distributions: how does impairment shift the rating landscape?

The grade distribution under clean data is the "true" portfolio profile.
Impairment should distort this -- damaged data may push firms into the
wrong grade, either overstating risk (conservative bias) or understating
it (dangerous). The question at each panel is: does preparation restore
the clean-data distribution?

```{r, fig.width=12, fig.height=8}

# ── Grade distribution comparison: GLM ─────────────────────────────────

condition_order <- c("mcar_mild", "mcar_severe",
                     "mar_mild",  "mar_severe",
                     "mnar_mild", "mnar_severe",
                     "noise_mild", "noise_severe",
                     "impl_mild", "impl_severe")

# Get the clean distribution once
clean_dist <- glm_grades %>%
  filter(condition == "clean") %>%
  group_by(grade) %>%
  summarise(n = n(), .groups = "drop") %>%
  mutate(pct = n / sum(n),
         stage = "clean")

# Replicate clean distribution into every impaired condition
clean_repeated <- do.call(rbind, lapply(condition_order, function(cond) {
  clean_dist %>% mutate(condition = cond)
}))

# Compute impaired + prepared distributions (excluding the clean panel)
impaired_dist <- glm_grades %>%
  filter(condition != "clean") %>%
  group_by(condition, stage, grade) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(condition, stage) %>%
  mutate(pct = n / sum(n)) %>%
  ungroup()

# Combine
glm_grade_dist           <- bind_rows(clean_repeated, impaired_dist)
glm_grade_dist$condition <- factor(glm_grade_dist$condition, levels = condition_order)
glm_grade_dist$stage     <- factor(glm_grade_dist$stage, levels = c("clean", "impaired", "prepared"))

ggplot(glm_grade_dist,
       aes(x = factor(grade), y = pct, fill = stage)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  facet_wrap(~ condition, ncol = 2) +
  scale_fill_manual(values = c(clean = "grey40", impaired = "#e74c3c", prepared = "#2ecc71")) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title    = "PD Grade Distribution: Logistic Regression",
       subtitle = "Grey = clean baseline repeated in every panel for comparison",
       x = "PD Grade (1=safest, 10=riskiest)", y = "Share of firms") +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"),
        legend.position = "bottom")


# 

```

```{r, fig.width=12, fig.height=8}

# ── Grade distribution comparison: XGBoost ─────────────────────────────

condition_order <- c("mcar_mild", "mcar_severe",
                     "mar_mild",  "mar_severe",
                     "mnar_mild", "mnar_severe",
                     "noise_mild", "noise_severe",
                     "impl_mild", "impl_severe")

# Get the clean distribution once
clean_dist_xgb <- xgb_grades %>%
  filter(condition == "clean") %>%
  group_by(grade) %>%
  summarise(n = n(), .groups = "drop") %>%
  mutate(pct = n / sum(n),
         stage = "clean")

# Replicate clean distribution into every impaired condition
clean_repeated_xgb <- do.call(rbind, lapply(condition_order, function(cond) {
  clean_dist_xgb %>% mutate(condition = cond)
}))

# Compute impaired + prepared distributions (excluding the clean panel)
impaired_dist_xgb <- xgb_grades %>%
  filter(condition != "clean") %>%
  group_by(condition, stage, grade) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(condition, stage) %>%
  mutate(pct = n / sum(n)) %>%
  ungroup()

# Combine
xgb_grade_dist <- bind_rows(clean_repeated_xgb, impaired_dist_xgb)
xgb_grade_dist$condition <- factor(xgb_grade_dist$condition, levels = condition_order)
xgb_grade_dist$stage     <- factor(xgb_grade_dist$stage, levels = c("clean", "impaired", "prepared"))

ggplot(xgb_grade_dist,
       aes(x = factor(grade), y = pct, fill = stage)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  facet_wrap(~ condition, ncol = 2) +
  scale_fill_manual(values = c(clean = "grey40", impaired = "#e74c3c", prepared = "#2ecc71")) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title    = "PD Grade Distribution: XGBoost",
       subtitle = "Grey = clean baseline repeated in every panel | No NAs — XGBoost handles missing values natively",
       x = "PD Grade (1=safest, 10=riskiest)", y = "Share of firms") +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"),
        legend.position = "bottom")


```




# **Stage 5: Evaluation Metrics**

## Overview

The evaluation stage asks: how much does each impairment condition
degrade model performance, and does preparation recover it? Following
the RR model validation framework, we measure two of the three pillars:
**discrimination** (can the model rank-order firms by risk?) and
**calibration** (do predicted PDs match observed default frequencies?).

(The third pillar — stability (temporal drift, population stability) — 
is outside scope. Although the simulation generates a longitudinal panel 
spanning 2005–2025, the research question concerns data quality effects 
on model reliability rather than temporal model degradation. Stability 
testing would require a dedicated development/validation time split, 
which is orthogonal to the Monte Carlo design and left for future work.)

**Evaluation design.** Every model -- clean-trained, impaired-trained,
prepared-trained -- predicts on the **same clean dataset**. This is the
critical design choice: the clean data is the true risk surface (known
from the DGP), and scoring every model against it isolates the effect of
training-data damage on model performance. The clean-trained model's
metrics are the best-case ceiling; degradation from that ceiling is what
we report.

Because the research question concerns relative degradation rather than
out-of-sample prediction, all models are evaluated on the full clean
dataset. The clean-trained model's in-sample metrics serve as the upper
bound. The Monte Carlo layer (Stage 6) provides the replication and
sampling variability that a train/test split would otherwise supply.

**Metrics computed:**

-   **AUC** (Area Under the ROC Curve): probability that a randomly
    chosen defaulter receives a higher predicted PD than a randomly
    chosen non-defaulter. Computed via the Mann-Whitney U statistic --
    no external package needed.
-   **Gini coefficient**: `2 × AUC - 1`. Linearly equivalent to AUC but
    the standard reporting metric in regulatory credit risk (EBA GL
    2017/16 uses AR/Gini for discriminatory power assessment).
-   **KS statistic** (Kolmogorov-Smirnov): maximum vertical distance
    between the CDF of predicted PDs for defaulters vs. non-defaulters.
    Captures separation at the single point of maximum divergence, which
    AUC (averaging over the whole curve) can miss.
-   **Brier score**: `mean((predicted - actual)²)`. A proper scoring
    rule that jointly captures discrimination and calibration. Lower is
    better.
-   **O/E ratio** (Observed / Expected): observed default rate divided
    by mean predicted PD. A value of 1.0 indicates perfect aggregate
    calibration; below 1.0 means the model overpredicts default risk;
    above 1.0 means underprediction. This is the simplest regulatory
    calibration check.

Main focus: Gini Coefficient (discrimination), O/E (overall calibration),
secondarily Brier Score (discrimination + calibration)

Running it for both models -- every model scores on the same clean data:

```{r}

glm_metrics <- evaluate_models(
  models     = glm_results$models,
  eval_data  = dat,
  formula    = pd_formula,
  model_type = "glm"
)

xgb_metrics <- evaluate_models(
  models     = xgb_results$models,
  eval_data  = dat,
  formula    = pd_formula,
  model_type = "xgboost"
)

# Combine into one table for comparison
all_metrics <- rbind(glm_metrics, xgb_metrics)

print(all_metrics[, c("model_type", "condition", "stage",
                       "auc", "gini", "ks", "brier", "oe_ratio")])

```

### Discrimination: how much ranking power does each model lose?

The primary question: when a model is trained on damaged data, how much
worse does it rank-order firms by risk? AUC and Gini capture this
directly. The clean-trained model sets the ceiling; any drop in the
impaired or prepared rows is degradation caused by the data quality
problem.

The plot below shows Gini (more common in credit risk reporting than
AUC) by condition and stage. The clean baseline appears as a horizontal
reference line in every panel.

```{r, fig.width=10, fig.height=6}

# ── Gini degradation plot ──────────────────────────────────────────────

condition_order <- c("mcar_mild", "mcar_severe",
                     "mar_mild",  "mar_severe",
                     "mnar_mild", "mnar_severe",
                     "noise_mild", "noise_severe",
                     "impl_mild", "impl_severe")

plot_metrics <- all_metrics[all_metrics$condition != "clean", ]
plot_metrics$condition <- factor(plot_metrics$condition, levels = condition_order)
plot_metrics$stage     <- factor(plot_metrics$stage, levels = c("impaired", "prepared"))

# Clean baselines for reference lines (one per model type)
clean_gini <- all_metrics[all_metrics$condition == "clean",
                          c("model_type", "gini")]

ggplot(plot_metrics, aes(x = condition, y = gini,
                         color = stage, shape = stage)) +
  geom_point(size = 3, position = position_dodge(width = 0.4)) +
  geom_hline(data = clean_gini,
             aes(yintercept = gini),
             linetype = "dashed", color = "grey40") +
  facet_wrap(~ model_type, ncol = 1) +
  scale_color_manual(values = c(impaired = "#e74c3c", prepared = "#2ecc71")) +
  scale_shape_manual(values = c(impaired = 16, prepared = 15)) +
  labs(title    = "Discriminatory Power: Gini Coefficient",
       subtitle = "Dashed line = clean-trained baseline | closer to line = better recovery",
       x = NULL, y = "Gini") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        strip.text  = element_text(face = "bold"),
        legend.position = "bottom")

```

**Notes on Gini / Discrimination**:

- XGBoost clean baseline (~0.675) meaningfully outperforms GLM (~0.60) due to its ability to capture nonlinearities in the DGP
- GLM is surprisingly robust across missingness conditions — complete-case deletion and RR's Ersatzwert imputation preserves clean relationships on smaller but undistorted data
- Preparation does not ever seem to help discrimination -- always on the same level as the default NA handling
- Severity gradient: mild-to-severe is a bigger step down for XGBoost than GLM
- MCAR missingness is a problem for XGB but not GLM. Because the missingness carries no signal and XG is looking to split based on signal. Preparation doesn't help in either case because there is nothing to fix in the relationship
- MNAR missingness: both models are able to rank firms in this case, I guess because decaying firms have other signs other than missing interest_cov. XGB native NA handling might have even learned that having missing "interest_cov" is a risk indicator
- Noise is where GLM shows more robustness than XGBoost relative to their own baselines, but XGBoost at its worst still beats GLM’s clean ceiling. In GLM, noise washes out so the signal isn't that far away. But XGB chases some of the noise damage / integrates into the trees
- Implausible values = biggest challenge for discrimination in GLM. Mild implausible with preparation can somehow solve it (winsorizing off the extreme values), but severe implausible is too much to ignore (especially when they can't get caught by the pipeline)




### Calibration: do predicted PDs still match reality?

The O/E ratio measures aggregate calibration -- is the average predicted
PD close to the observed default rate? Since every model predicts on the
same clean data (where the true default rate is fixed), any shift in O/E
away from the clean baseline means the model's probability estimates
have been distorted by impairment.

The Brier score combines calibration and discrimination into a single
proper scoring rule. A model that ranks firms well but assigns the wrong
PD levels will have a good AUC but a poor Brier score; the gap between
the two tells you whether the problem is discrimination or calibration.

```{r, fig.width=10, fig.height=6}

# ── Brier score degradation plot ───────────────────────────────────────

clean_brier <- all_metrics[all_metrics$condition == "clean",
                           c("model_type", "brier")]

ggplot(plot_metrics, aes(x = condition, y = brier,
                         color = stage, shape = stage)) +
  geom_point(size = 3, position = position_dodge(width = 0.4)) +
  geom_hline(data = clean_brier,
             aes(yintercept = brier),
             linetype = "dashed", color = "grey40") +
  facet_wrap(~ model_type, ncol = 1) +
  scale_color_manual(values = c(impaired = "#e74c3c", prepared = "#2ecc71")) +
  scale_shape_manual(values = c(impaired = 16, prepared = 15)) +
  labs(title    = "Calibration Quality: Brier Score",
       subtitle = "Dashed line = clean-trained baseline | lower is better",
       x = NULL, y = "Brier Score") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        strip.text  = element_text(face = "bold"),
        legend.position = "bottom")

```


**Brier score** (composite, individual level discrimination + calibration)

- XGBoost again has the better clean baseline. Starts lower and stays lower.
- Preparation hurts Brier score. Prepared data has the "mass point" issue that shrinks the variance, then GLM overcompensates / higher predicted PDs than expected. For XGB: mass point downgrades its information set. Replaces a better mechanism with a less-precise one
  - Therefore it's almost always higher than the impaired data in all cases)
  - XGBoost NA handling for MCAR / MNAR severe,noise also distorts the trees. 
- GLM complete-case analysis holds calibration well until noise severe and implausible conditions, where it jumps sharply.  Impl Mild and Severe stay at the same level.
- MAR is the most stable condition for both models, consistent with the mechanism (missingness on crefo is independent of default risk conditional on observables)
- Implausible mild GLM: the only case where preparation helps (winsorizing), but we can't get rid of the distorted data not caught by preparation. 



```{r, fig.width=10, fig.height=6}

# ── O/E ratio plot ─────────────────────────────────────────────────────

clean_oe <- all_metrics[all_metrics$condition == "clean",
                        c("model_type", "oe_ratio")]

ggplot(plot_metrics, aes(x = condition, y = oe_ratio,
                         color = stage, shape = stage)) +
  geom_point(size = 3, position = position_dodge(width = 0.4)) +
  geom_hline(yintercept = 1.0, linetype = "solid", color = "grey70", linewidth = 0.3) +
  geom_hline(data = clean_oe,
             aes(yintercept = oe_ratio),
             linetype = "dashed", color = "grey40") +
  facet_wrap(~ model_type, ncol = 1) +
  scale_color_manual(values = c(impaired = "#e74c3c", prepared = "#2ecc71")) +
  scale_shape_manual(values = c(impaired = 16, prepared = 15)) +
  labs(title    = "Aggregate Calibration: Observed / Expected Ratio",
       subtitle = "Solid grey = perfect calibration (1.0) | Dashed = clean baseline",
       x = NULL, y = "O/E Ratio") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        strip.text  = element_text(face = "bold"),
        legend.position = "bottom")

```

**O/E ratio** (pure calibration on overall portfolio level):

Note that this is a ratio of totals, and errors can cancel out in aggregate (doesn't look as dramatic as Brier)

Can diverge from Brier when prediction errors are unsystematic

- Ersatzwert with XGBoost leads to a loss of precision that shows up in Brier (which is more about grading individual predictions) but is hidden in O/E (aggregate measure)
  - Ersatzwert with GLM  hurts calibration due to the mass-point effect, though at least being under the 1.0 line is safer from a regulatory perspective
- MNAR Severe: an interesting story
  - GLM MNAR without preparation underpredicts risk (dangerous). MNAR deletes interest_cov for distressed firms, and complete case analysis deletes all of these firms, so it thinks the risk isn’t there
  - Ersatzwert GLM preparation flips it: replaces the missing interest_cov with one value that represents all the distressed firms, therefore overstating the risk. Trading one miscalibration for another
    -  XGBoost lands on the higher side regardless of preparation. Can't overcome the bias either or learn from the missingness (can overcome MAR perfectly but can't figure this out)
- Implausible and Noise doesn't look too far from the 1.0, but Gini showed meaningful loss in discrimination (dangerous because it might fly under the radar)

Implausible mild vs. severe for prepared data: under mild impairment, the winsorizing thresholds are tighter and it clips more into legit variation in the data --> distorts it



### Calibration plot: predicted vs. observed by PD grade

The grade-level calibration plot uses the same Moody's-aligned PD
buckets from Stage 4c. For each grade, we compare the average predicted
PD against the observed default frequency. A well-calibrated model
produces points along the 45-degree line. Impairment can pull points off
the diagonal -- either by mispricing individual grades or by shifting
the entire curve.

We show this for the clean baseline and all prepared conditions (the
ones a practitioner would actually deploy). Impaired conditions are
omitted for readability; the prepared-vs-clean comparison is the one
that answers the research question.

```{r, fig.width=10, fig.height=7}

# ── Calibration plot helper ────────────────────────────────────────────
build_calibration_data <- function(models,
                                   eval_data,
                                   formula,
                                   model_type = c("glm", "xgboost")) {

  model_type <- match.arg(model_type)
  response   <- all.vars(formula)[1]
  predictors <- all.vars(formula)[-1]
  actual     <- eval_data[[response]]

  # Same grade boundaries as assign_pd_grades()
  grade_breaks <- c(0, 0.00001, 0.0006, 0.0011, 0.003,
                    0.007, 0.015, 0.03, 0.08, 0.15, 1.0)
  grade_labels <- as.character(1:10)

  cal_list <- vector("list", length(models))
  names(cal_list) <- names(models)

  for (key in names(models)) {

    mod <- models[[key]]

    if (model_type == "glm") {
      pd_pred <- predict(mod, newdata = eval_data, type = "response")
    } else {
      X <- as.matrix(eval_data[, predictors, drop = FALSE])
      pd_pred <- predict(mod, xgb.DMatrix(data = X))
    }

    grade <- cut(pd_pred,
                 breaks = grade_breaks,
                 labels = grade_labels,
                 include.lowest = TRUE,
                 right = TRUE)

    tmp <- data.frame(pd_pred = pd_pred, actual = actual,
                      grade = as.integer(as.character(grade)))

    agg <- aggregate(cbind(pd_pred, actual) ~ grade, data = tmp, FUN = mean)
    counts <- aggregate(actual ~ grade, data = tmp, FUN = length)
    names(counts)[2] <- "n"
    agg <- merge(agg, counts, by = "grade")

    names(agg)[names(agg) == "pd_pred"] <- "mean_pred_pd"
    names(agg)[names(agg) == "actual"]  <- "obs_default_rate"

    # Parse key
    stage <- ifelse(key == "clean", "clean",
                    ifelse(grepl("^imp_", key), "impaired", "prepared"))
    condition <- ifelse(key == "clean", "clean",
                        sub("^imp_|^prep_", "", key))

    agg$key        <- key
    agg$stage      <- stage
    agg$condition  <- condition
    agg$model_type <- model_type

    cal_list[[key]] <- agg
  }

  cal_df <- do.call(rbind, cal_list)
  rownames(cal_df) <- NULL
  cal_df
}

# ── Build calibration data for both model types ────────────────────────

glm_cal <- build_calibration_data(
  models     = glm_results$models,
  eval_data  = dat,
  formula    = pd_formula,
  model_type = "glm"
)

xgb_cal <- build_calibration_data(
  models     = xgb_results$models,
  eval_data  = dat,
  formula    = pd_formula,
  model_type = "xgboost"
)

all_cal <- rbind(glm_cal, xgb_cal)

# Filter to clean + prepared only (what a practitioner would deploy)
cal_plot_data <- all_cal[all_cal$stage %in% c("clean", "prepared"), ]

# For prepared conditions, parse type + severity for faceting
cal_plot_data$type <- ifelse(
  cal_plot_data$condition == "clean", "clean",
  sub("_mild$|_severe$", "", cal_plot_data$condition)
)
cal_plot_data$severity <- ifelse(
  cal_plot_data$condition == "clean", "none",
  ifelse(grepl("severe", cal_plot_data$condition), "severe", "mild")
)

# Label for faceting: one panel per impairment type
cal_plot_data$facet_label <- ifelse(
  cal_plot_data$condition == "clean", "clean",
  cal_plot_data$condition
)

condition_order <- c("clean",
                     "mcar_mild", "mcar_severe",
                     "mar_mild",  "mar_severe",
                     "mnar_mild", "mnar_severe",
                     "noise_mild", "noise_severe",
                     "impl_mild", "impl_severe")

cal_plot_data$facet_label <- factor(cal_plot_data$facet_label,
                                    levels = condition_order)



```


```{r}
# ── Calibration deviation heatmap ──────────────────────────────────────

# Relative deviation: positive = under-prediction, negative = over-prediction
cal_plot_data$rel_dev <- with(cal_plot_data,
  (obs_default_rate - mean_pred_pd) / pmax(mean_pred_pd, 1e-6)
)

# Cap for visual clarity
cal_plot_data$rel_dev_cap <- pmin(pmax(cal_plot_data$rel_dev, -1), 1)

for (mt in c("glm", "xgboost")) {
  p <- ggplot(cal_plot_data[cal_plot_data$model_type == mt, ],
              aes(x = facet_label, y = factor(grade), fill = rel_dev_cap)) +
    geom_tile(color = "white", linewidth = 0.5) +
    scale_fill_gradient2(
      low = "#2166ac", mid = "white", high = "#b2182b",
      midpoint = 0, limits = c(-1, 1),
      name = "Relative\ndeviation"
    ) +
    geom_text(aes(label = sprintf("%.0f%%", rel_dev * 100)),
              size = 2.5, color = "grey20") +
    labs(
      title = paste0("Calibration Deviation by Grade — ", toupper(mt)),
      subtitle = "Blue = over-prediction (conservative) | Red = under-prediction (risky)",
      x = NULL, y = "PD Grade"
    ) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
      panel.grid   = element_blank()
    )
  print(p)
}
```


**Calibration plot (grade-level calibration): **

- GLM underestimates the risk on the safe firms (low rating) but overestimates on the risky firms (low rating)
- XGBoost does the opposite (arguably worse), trees are potentially overfit?
- Noise and implausibles for GLM also create more of a dangerous pattern where risky firms are underestimated



### Summary: degradation at a glance

A compact summary showing all five metrics for each condition, with the
clean baseline at the top for reference. The `delta_gini` and
`delta_brier` columns show absolute change from the clean-trained model,
making it easy to scan for which conditions hurt most and whether
preparation recovers the damage.

```{r}

# ── Summary table with deltas from clean ───────────────────────────────

# Get clean baselines
clean_baselines <- all_metrics[all_metrics$condition == "clean",
                               c("model_type", "gini", "brier", "oe_ratio")]
names(clean_baselines) <- c("model_type", "clean_gini", "clean_brier", "clean_oe")

summary_metrics <- merge(all_metrics, clean_baselines, by = "model_type")
summary_metrics$delta_gini  <- summary_metrics$gini  - summary_metrics$clean_gini
summary_metrics$delta_brier <- summary_metrics$brier - summary_metrics$clean_brier
summary_metrics$delta_oe    <- summary_metrics$oe_ratio - summary_metrics$clean_oe

# Order for readability
summary_metrics$condition <- factor(
  summary_metrics$condition,
  levels = c("clean",
             "mcar_mild", "mcar_severe",
             "mar_mild",  "mar_severe",
             "mnar_mild", "mnar_severe",
             "noise_mild", "noise_severe",
             "impl_mild", "impl_severe")
)

summary_metrics <- summary_metrics[order(summary_metrics$model_type,
                                         summary_metrics$condition,
                                         summary_metrics$stage), ]

print(summary_metrics[, c("model_type", "condition", "stage",
                           "gini", "delta_gini",
                           "brier", "delta_brier",
                           "oe_ratio", "delta_oe",
                           "ks")])

```

# **Stage 6: Monte Carlo**

Assuming the Monte Carlo has been run and stored as an RDS file in the current
project directory.  

---

The single-iteration analysis above demonstrates *how* each impairment type and the Ersatzwert preparation pipeline affect model performance. But a single random draw can't tell us whether those effects are stable or just an artifact of one particular dataset. The Monte Carlo simulation repeats the full pipeline — DGP → impair → prepare → fit → evaluate — across X independent iterations, each with a fresh random seed and 5,000 firms, to produce distributional evidence.

The MC runner (`03_monte_carlo.R`) sources the same functions used throughout this notebook. The full set of tables and figures is produced separately by `04_mc_analysis.R`; here we show the key results.



```{r}

mc <- readRDS("mc_results_final.rds")

cat(sprintf("Monte Carlo iterations: %d\nFirms per iteration: %d\nTotal metric rows: %d\n",
            mc$config$N_MC, mc$config$N_FIRMS, nrow(mc$metrics)))

# Shared aesthetics
stage_colors <- c(clean = "#2C3E50", impaired = "#E74C3C", prepared = "#2ECC71")

theme_thesis <- theme_minimal(base_size = 11) +
  theme(
    plot.title       = element_text(face = "bold", size = 12),
    strip.text       = element_text(face = "bold"),
    legend.position  = "bottom",
    panel.grid.minor = element_blank()
  )


summary_table <- mc$metrics %>%
  group_by(model_type, type, severity, stage) %>%
  summarise(
    gini_mean = mean(gini),   gini_sd = sd(gini),
    brier_mean = mean(brier), brier_sd = sd(brier),
    oe_mean = mean(oe_ratio), oe_sd = sd(oe_ratio),
    .groups = "drop"
  )

summary_table

```


## 6.1 Summary Table

The core finding in one table: mean and standard deviation of each metric across all MC iterations, by impairment type, severity, stage, and model architecture. Low standard deviations confirm that single-iteration results were not flukes.

```{r mc-summary-table}
clean_baseline <- mc$metrics %>%
  filter(stage == "clean") %>%
  select(iteration, model_type,
         gini_clean = gini, brier_clean = brier, oe_clean = oe_ratio)

tbl <- mc$metrics %>%
  left_join(clean_baseline, by = c("iteration", "model_type")) %>%
  mutate(
    delta_gini  = gini     - gini_clean,
    delta_brier = brier    - brier_clean,
    delta_oe    = oe_ratio - oe_clean
  ) %>%
  group_by(model_type, type, severity, stage) %>%
  summarise(
    gini_mean  = round(mean(gini), 4),
    gini_sd    = round(sd(gini), 4),
    d_gini     = round(mean(delta_gini), 4),
    brier_mean = round(mean(brier), 6),
    oe_mean    = round(mean(oe_ratio), 4),
    d_oe       = round(mean(delta_oe), 4),
    .groups    = "drop"
  ) %>%
  arrange(model_type, type, severity, match(stage, c("clean", "impaired", "prepared")))

knitr::kable(tbl, caption = "Monte Carlo summary: metric means and standard deviations")
```

## 6.2 Gini Coefficient — Discrimination Stability

Boxplots across all iterations confirm which impairment types degrade discrimination (Gini drops) and whether Ersatzwert preparation recovers it.

```{r mc-gini-boxplots, fig.width=12, fig.height=8}
plot_data <- mc$metrics %>%
  mutate(
    stage    = factor(stage, levels = c("clean", "impaired", "prepared")),
    severity = factor(severity, levels = c("none", "mild", "severe")),
    panel    = ifelse(type == "clean", "clean",
                      paste0(type, " — ", severity))
  )

for (mt in c("glm", "xgboost")) {

    clean_gini_mean <- mc$metrics %>%
    filter(model_type == mt, stage == "clean") %>%
    pull(gini) %>% mean()
  
  p <- ggplot(
    plot_data %>% filter(model_type == mt),
    aes(x = stage, y = gini, fill = stage)
  ) +
    geom_hline(yintercept = clean_gini_mean, linetype = "dotted", color = "grey40") +
    geom_boxplot(outlier.size = 0.5, alpha = 0.7) +
    scale_fill_manual(values = stage_colors) +
    facet_wrap(~ panel, scales = "free_x", nrow = 2) +
    labs(
      title = sprintf("Gini across %d MC iterations (%s)", mc$config$N_MC, toupper(mt)),
      x = NULL, y = "Gini"
    ) +
    theme_thesis +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))

  print(p)
}
```

Key observations:

- **Missing-data types (MCAR, MAR, MNAR):** Gini is largely preserved even after impairment, and Ersatzwert preparation maintains this. The discrimination-preserving property of Ersatzwert is confirmed as a stable result, not a single-draw artifact.
- **Noise and implausible:** These show visible Gini degradation, especially at severe levels. Preparation does *not* fully recover discrimination here — the signal corruption cannot be undone by imputation.
- **GLM vs. XGBoost:** XGBoost achieves a higher clean-data Gini ceiling but shows wider boxes under noise/implausible, confirming its greater sensitivity to value corruption.

## 6.3 O/E Ratio — Calibration Distortion

The O/E ratio reveals whether predicted default rates match observed rates. The dashed line at 1.0 represents perfect calibration.

```{r mc-oe-boxplots, fig.width=12, fig.height=8}
for (mt in c("glm", "xgboost")) {

  p <- ggplot(
    plot_data %>% filter(model_type == mt),
    aes(x = stage, y = oe_ratio, fill = stage)
  ) +
    geom_boxplot(outlier.size = 0.5, alpha = 0.7) +
    geom_hline(yintercept = 1, linetype = "dashed", color = "grey40") +
    scale_fill_manual(values = stage_colors) +
    facet_wrap(~ panel, scales = "free_x", nrow = 2) +
    labs(
      title = sprintf("O/E ratio across %d MC iterations (%s)", mc$config$N_MC, toupper(mt)),
      x = NULL, y = "O/E ratio"
    ) +
    theme_thesis +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))

  print(p)
}
```

Key observations:

- **Ersatzwert distorts calibration under missingness:** The mass-point effect of single-value replacement systematically shifts the predicted default rate. O/E moves away from 1.0 after preparation, even though discrimination (Gini) is preserved. This is the core Ersatzwert trade-off, confirmed across all MC iterations.
- **Noise/implausible show the opposite pattern:** O/E stays relatively stable (aggregate rates are preserved when values are corrupted, not removed), but Gini degrades. A calibration-only audit would miss this degradation entirely.

## 6.4 Degradation Heatmap

A single view of mean Gini degradation from each iteration's clean baseline.

```{r mc-heatmap, fig.width=10, fig.height=6}
heatmap_data <- mc$metrics %>%
  filter(stage != "clean") %>%
  left_join(clean_baseline, by = c("iteration", "model_type")) %>%
  mutate(delta_gini = gini - gini_clean) %>%
  group_by(model_type, type, severity, stage) %>%
  summarise(delta_gini = mean(delta_gini), .groups = "drop") %>%
  mutate(
    row_label = paste(type, severity),
    col_label = paste(toupper(model_type), stage)
  )

ggplot(heatmap_data, aes(x = col_label, y = row_label, fill = delta_gini)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = sprintf("%.4f", delta_gini)), size = 3) +
  scale_fill_gradient2(
    low = "#E74C3C", mid = "white", high = "#2ECC71",
    midpoint = 0, name = "Mean\n\u0394 Gini"
  ) +
  labs(
    title = "Mean Gini degradation from clean baseline",
    subtitle = "Negative = worse discrimination. Per-iteration pairing.",
    x = NULL, y = NULL
  ) +
  theme_thesis +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

## 6.5 Coefficient Recovery

GLM coefficients across iterations, with ±1 SD bands. The dashed line marks the true DGP beta for each predictor.

```{r mc-coef-recovery, fig.width=14, fig.height=10}
coef_summary <- mc$coefs %>%
  group_by(stage, condition, type, severity, term) %>%
  summarise(
    est_mean = mean(estimated),
    est_lo   = mean(estimated) - sd(estimated),
    est_hi   = mean(estimated) + sd(estimated),
    true_val = first(true),
    .groups  = "drop"
  ) %>%
  mutate(
    stage = factor(stage, levels = c("clean", "impaired", "prepared")),
    panel = ifelse(condition == "clean", "clean",
                   paste0(type, " — ", severity))
  )

key_terms <- c("log_assets", "debt_to_equity", "interest_cov",
               "firm_age", "crefo")

ggplot(
  coef_summary %>% filter(term %in% key_terms),
  aes(x = stage, y = est_mean, color = stage)
) +
  geom_pointrange(aes(ymin = est_lo, ymax = est_hi), size = 0.4) +
  geom_hline(aes(yintercept = true_val), linetype = "dashed",
             color = "grey50", linewidth = 0.3) +
  scale_color_manual(values = stage_colors) +
  facet_grid(term ~ panel, scales = "free_y") +
  labs(
    title = "GLM coefficient recovery across MC iterations",
    subtitle = "Points = mean estimate, bars = \u00B1 1 SD. Dashed = true DGP beta.",
    x = NULL, y = "Estimated coefficient"
  ) +
  theme_thesis +
  theme(
    axis.text.x  = element_text(angle = 45, hjust = 1, size = 7),
    strip.text.y = element_text(angle = 0)
  )
```

```{r}
# ── Quadrant plot: Gini (x) vs O/E (y) ──────────────────────────

# ── 1. Get clean baselines per model ─────────────────────────────
clean_ref <- summary_table %>%
  filter(stage == "clean") %>%
  select(model_type, clean_gini = gini_mean, clean_oe = oe_mean)

# ── 2. Build plot data: impaired + prepared only ─────────────────
quad_data <- summary_table %>%
  filter(stage != "clean") %>%
  left_join(clean_ref, by = "model_type") %>%
  mutate(
    condition = paste0(type, "_", severity),
    label     = paste0(type, "_", severity, "\n(", stage, ")"),
    short_label = paste0(
      substr(type, 1, 4),
      ifelse(severity == "mild", "_m", "_s")
    )
  )

# ── 3. Shape and color mappings ─────────────────────────────────
stage_shapes  <- c(impaired = 16, prepared = 15)   # circle vs square
type_colors  <- c(
  mcar = "#3498DB",   # blue
  mar  = "#2ECC71",   # green
  mnar = "#F39C12",   # orange
  noise = "#E74C3C",  # red
  impl = "#9B59B6"    # purple
)

# ── 4. Plot ──────────────────────────────────────────────────────
for (mt in c("glm", "xgboost")) {
  
  ref <- clean_ref %>% filter(model_type == mt)
  
  p <- ggplot(
    quad_data %>% filter(model_type == mt),
    aes(x = gini_mean, y = oe_mean, color = type, shape = stage)
  ) +
    # Crosshair at clean baseline
    geom_hline(yintercept = ref$clean_oe, linetype = "dashed", color = "grey50", linewidth = 0.4) +
    geom_vline(xintercept = ref$clean_gini, linetype = "dashed", color = "grey50", linewidth = 0.4) +
    
    # Perfect calibration reference
    geom_hline(yintercept = 1.0, linetype = "dotted", color = "grey80", linewidth = 0.3) +
    
    # Points
    geom_point(size = 3.5, alpha = 0.85) +
    
    # Labels (using ggrepel to avoid overlap)
    geom_text_repel(
      aes(label = short_label),
      size = 2.8,
      max.overlaps = 20,
      segment.size = 0.3,
      segment.color = "grey70",
      box.padding = 0.4
    ) +
    
    scale_color_manual(values = type_colors, name = "Impairment") +
    scale_shape_manual(values = stage_shapes, name = "Stage") +
    
    # Quadrant annotations
   # annotate("text", x = ref$clean_gini + 0.002, y = ref$clean_oe + 0.02,
    #         label = "Healthy", fontface = "italic", size = 3, color = "grey50") +
    #annotate("text", x = ref$clean_gini - 0.015, y = ref$clean_oe + 0.02,
     #        label = "Calibrated but\ncan't rank", fontface = "italic",
      #       size = 2.5, color = "grey50") +
  #  annotate("text", x = ref$clean_gini + 0.002, y = ref$clean_oe - 0.04,
       #      label = "Ranks well but\nmiscalibrated", fontface = "italic",
        #     size = 2.5, color = "grey50") +
  #  annotate("text", x = ref$clean_gini - 0.015, y = ref$clean_oe - 0.04,
          #   label = "Both broken", fontface = "italic",
         #    size = 2.5, color = "grey50") +
    
    labs(
      title    = sprintf("Discrimination vs. Calibration — %s", toupper(mt)),
      subtitle = "Dashed crosshair = clean-trained baseline | Dotted = perfect calibration (O/E = 1.0)",
      x = "Gini (discrimination →)",
      y = "O/E ratio (calibration)"
    ) +
    theme_minimal(base_size = 11) +
    theme(
      plot.title       = element_text(face = "bold", size = 12),
      legend.position  = "bottom",
      panel.grid.minor = element_blank()
    )
  
  print(p)
}



```



