---
title: "Credit Risk Data Quality Simulation"
output:
  html_document: default
  pdf_document: default
date: "2026-03-02" 
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Part 0: Helper functions and setup

```{r, warning=FALSE, message=FALSE}
source("02_helperfunctions.R")

required_packages <- c("ggplot2", "dplyr", "tidyr", "mice", "xgboost", "DiagrammeR")

for (pkg in required_packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg)
  }
  library(pkg, character.only = TRUE)
}

```

# **Stage 1: Creating the "Perfect World" with the DGP**

### Overview

------------------------------------------------------------------------

##### **Phase 1. Generating features (predictors) for the business world**

Have to define both the variables and their distributions /
dependencies.

1: **Root variables**

a)  **firm age**: right skewed so using gamma distribution
b)  **sector**: three sectors with proportions and risk properties based
    off real German practitioner data.
    -   Sector 1: Services (Dienstleistungen) at 57% of the sample.
        Least risky with 2% default rate
    -   Sector 2: Hospitality (Gastro), 19% of the sample. Most risky,
        3.6% default rate
    -   Sector 3: Manufacturing (Verarbeitendes Gewerbe), 24% of the
        sample. Medium risk, 2.3% default rate

2: **Chained regression dependencies**

Each variable is determined by the one in front of it -- older firms
tend to be larger, larger firms tend to have lower leverage Plus --
adding randomness

c)  **log_assets**: proxy for firm size, dependent on age and sector
d)  **debt to equity**: higher leverage is more dangerous -- dependent
    on how big the firm is and how old it is
e)  **interest_cov**: proxy for cash flow
f)  **Crefo score**. (see below) Dependent on most things above.
    Sometimes gets lagged (not updated every year)

------------------------------------------------------------------------

##### **2. Choose the true coefficients that define the world**

This is the "ground truth" -- can be confirmed from RR All are
standardized (change in log-odds per 1 SD increase)

beta0 <- -3.9 this is the intercept -- should be whatever the log odds
of the default rate is 

beta_log_assets <- -0.2 larger firms less likely to default 

beta_debt_to_equity <- 0.4 higher leverage = higher risk

beta_interest_cov <- -0.3 higher coverage of interest expenses = lower risk 

beta_sector2 <- 0.4 gastro is the riskiest sector 

beta_sector3 <- 0.18 manufacturing slightly riskier 

beta_firm_age <- -0.1 older firms slightly safer 

beta_crefo <- 0.3 higher Crefo score → higher default risk

------------------------------------------------------------------------

##### **3. Create panel data with time series component**

Inner loop - creating firms that either stay the same over time, get
worse, or improve... (drift rates); default or not during their 3-7 year
loan periods. Crefo will sometimes refresh itself, sometimes remaining
stale

Outer loop creates lots of these firms. Then the rows are bound together

-   the "linpred" part -- weighted sum of the predictors. this is
    logistic regression
    -   computes the log-odds of default for each borrower
-   pd_true: then after: the logistic transformation so we get the true
    PDs and not log-odds

Draw observed defaults. For each borrower, draw a default with pd_true
value from before - borrowers can have same PDs but 2 different outcomes
(default or not)

------------------------------------------------------------------------

##### **4. Reporting and returning a list**

-   Data has both raw and scaled versions of every variable
-   Scaling = the mean and SD of every predictor so you can apply
    consistently to the impaired datasets
-   Also saving the bounds for winsorization and plausibility (related
    to data preparation and FRS scores)

True betas = ground truth coefficients

------------------------------------------------------------------------

**A quick note on Crefo score (Crefo Bonitätsindex)**

Scale: 100 (excellent) to 600 (insolvent) — higher = worse
creditworthiness. Generated as a noisy composite of the same financial
signals the bureau uses: leverage, coverage, firm size, and sector.
Idiosyncratic noise (sd = 25) reflects bureau information not in our DGP
(payment history with other creditors, legal registry data, etc.).

Sector penalties: gastro (40) and manufacturing (20) reflect higher
historical default rates in those sectors per practitioner datasets.

```{r}

# Creating panel data

##########################################

simulate_clean_data_panel <- function(N = 5000, seed = 21,
                                      min_term = 3, max_term = 7,
                                      drift_sd_log_assets     = 0.05,
                                      drift_sd_debt_to_equity = 0.3,
                                      drift_sd_interest_cov   = 0.3,
                                      drift_scale             = 4.0,
                                      seasoning_period        = 1,
                                      crefo_refresh_prob      = 0.5) {  
  set.seed(seed)
  
  # ============================================================
  # PHASE 1: FIRM ORIGINATION
  # ============================================================
  
  # Var 1: Firm Age
  firm_age_orig <- rgamma(N, shape = 3, rate = 0.15)
  
  
  # Var 2: Sector
  # Three sectors instead of binary
  # 1 = services (reference), 2 = hospitality, 3 = manufacturing
  # Proportions reflect a realistic B2B lending portfolio for a leasing company
  sector   <- sample(1:3, N, replace = TRUE, prob = c(0.57, 0.19, 0.24))
  sector_2 <- as.integer(sector == 2)   # Hospitality (gastro) dummy
  sector_3 <- as.integer(sector == 3)   # Manufacturing (Verarb. Gewerbe) dummy
  
  # Var 3: Total Assets (firm size. log transformed, as per Ohlson (1980))
  log_assets_orig <- 8 + 0.03 * firm_age_orig - 0.3 * sector_2 + 0.3 * sector_3 + rnorm(N, 0, 1)
      # Gastro - smaller firms, manufacturing - need for more assets
  
  # Var 4: Debt-to-Equity Ratio (leverage)
  debt_to_equity_orig <- pmax(3 - 0.1 * scale(log_assets_orig) - 0.02 * scale(firm_age_orig) +   rgamma(N, shape = 1, rate = 1), 0)
  
  # Var 5: Interest Coverage Ratio (cash flow)
  interest_cov_orig   <- pmax(5 - 0.4 * scale(debt_to_equity_orig) + 0.3 * scale(log_assets_orig) + rnorm(N, 0, 1), 0)
  
  log_assets_mean     <- mean(log_assets_orig);     log_assets_sd     <- sd(log_assets_orig)
  debt_to_equity_mean <- mean(debt_to_equity_orig); debt_to_equity_sd <- sd(debt_to_equity_orig)
  interest_cov_mean   <- mean(interest_cov_orig);   interest_cov_sd   <- sd(interest_cov_orig)
  firm_age_mean       <- mean(firm_age_orig);       firm_age_sd       <- sd(firm_age_orig)
  

  # Var 6: CreditReform Bonitätsindex (Crefo score)
    crefo_orig <- round(
      300
      + 60  * scale(debt_to_equity_orig)
      - 50  * scale(interest_cov_orig)
      - 30  * scale(log_assets_orig)
      + 40  * sector_2   # Gastro: worse baseline creditworthiness
      + 20  * sector_3   # Manufactuing: slight penalty vs services
      + rnorm(N, 0, 25)
    )
  
  crefo_orig <- pmin(pmax(crefo_orig, 100), 600)     # clamp to valid Crefo range
  
  crefo_mean <- mean(crefo_orig);   crefo_sd   <- sd(crefo_orig)
  
  # Additional variables -- keeping record of firms and their loans over time
  firm_id      <- sprintf("FIRM_%05d", seq_len(N))
  vintage_year <- sample(2005:2025, N, replace = TRUE)
  loan_term    <- sample(min_term:max_term, N, replace = TRUE)
  
  # Firm trajectory type: assigned once at origination, fixed for the firm's lifetime.
  # Determines the direction of the drift mean in the annual random walk.
  # Deteriorating firms face state-dependent worsening (as before).
  # Stable firms follow a pure random walk with zero drift mean.
  # Improving firms experience a gentle drift in the opposite direction --
  # leverage falls slowly, coverage and assets improve.
  # Proportions (0.2 / 0.6 / 0.2) reflect a realistic credit portfolio
  # where most firms are broadly stable and distressed/improving firms are minorities.
  trajectory <- sample(c("deteriorating", "stable", "improving"),
                       N, replace = TRUE, prob = c(0.2, 0.6, 0.2))
  
  # ============================================================
  # PHASE 2: TRUE COEFFICIENTS
  # All betas are on the standardized scale (change in log-odds per 1 SD).
  # ============================================================
  
  beta0               <- -3.9   # the log-odds of the default rate we want to recover. currently around 2%
  beta_log_assets     <- -0.2   # larger companies - less risk
  beta_debt_to_equity <-  0.4   # companies with more leverage (debt) = more risk
  beta_interest_cov   <- -0.3   # more interest coverage / higher cash flows = less risk
  beta_sector2        <-  0.4   # gastro = more risk, higher default rate
  beta_sector3        <-  0.18  # manufacturing needs slight uplift to get to 2.3% default rate
  beta_firm_age       <- -0.2   # older companies - less risk
  beta_crefo          <-  0.3   # higher Crefo score → higher default risk
  
  
  # Accounting for Gastro default risk increases in later vintages
    # Centered on midpoint of 2005-2025 range so effect is ~zero at portfolio average
  gastro_trend        <- 0.04  # log-odds increase per year for Gastro
  vintage_center      <- 2015
         # (by 2025, gastro firms get an additional +.4 on log-odds scale relative to 2025, so a meaningful increase in PD)
      
  # ============================================================
  # PHASE 3: PANEL EXPANSION
  # ============================================================
  
  rows <- vector("list", N)
  
  # Creating the origination values for firm [i]
  for (i in seq_len(N)) {
    
    la    <- log_assets_orig[i]
    dte   <- debt_to_equity_orig[i]
    ic    <- interest_cov_orig[i]
    fa    <- firm_age_orig[i]
    crefo <- crefo_orig[i]           
    
    firm_rows <- vector("list", loan_term[i])
    
    for (t in seq_len(loan_term[i])) {
      
      obs_year <- vintage_year[i] + t - 1
      
      # --- 3a. State-dependent drift from year 2 onward ---
      if (t > 1) {
        
        # Standardize current values to compute pd_curr as deterioration weight
        la_z_curr  <- (la  - log_assets_mean)     / log_assets_sd
        dte_z_curr <- (dte - debt_to_equity_mean) / debt_to_equity_sd
        ic_z_curr  <- (ic  - interest_cov_mean)   / interest_cov_sd
        fa_z_curr  <- (fa  - firm_age_mean)        / firm_age_sd
        
        # pd_curr probability creation: how intensely are you deteriorating (if at all)?
        pd_curr <- plogis(beta0 +
                            beta_log_assets     * la_z_curr  +
                            beta_debt_to_equity * dte_z_curr +
                            beta_interest_cov   * ic_z_curr  +
                            beta_sector2        * sector_2[i] +
                            beta_sector3        * sector_3[i] +
                            beta_firm_age       * fa_z_curr)
        # Note: Crefo intentionally excluded from pd_curr used for drift weighting.
        # The drift mechanism operates on the underlying financial state; Crefo
        # is a derived / external signal and including it here would create circular feedback.
        
        # Below: the drift mechanism depending on state (deteriorating, stable, improving).
        # For deteriorating: the worse your current PD is, the stronger it pulls you down.
        if (trajectory[i] == "deteriorating") {
          drift_mean_dte <- +drift_scale * pd_curr
          drift_mean_ic  <- -drift_scale * pd_curr
          drift_mean_la  <- -drift_scale * 0.5 * pd_curr
        } else if (trajectory[i] == "stable") {
          drift_mean_dte <- 0
          drift_mean_ic  <- 0
          drift_mean_la  <- 0
        } else {  # improving
          drift_mean_dte <- -drift_scale * 0.5
          drift_mean_ic  <-  drift_scale * 0.5
          drift_mean_la  <-  drift_scale * 0.25
        }
        
        # Here are the yearly updates for the financials: drifting plus random noise
        dte <- pmax(dte + rnorm(1, mean = drift_mean_dte, sd = drift_sd_debt_to_equity), 0)
        ic  <- pmax(ic  + rnorm(1, mean = drift_mean_ic,  sd = drift_sd_interest_cov),   0)
        la  <-      la  + rnorm(1, mean = drift_mean_la,  sd = drift_sd_log_assets)
        fa  <- fa + 1
        
        # Added: Crefo staleness mechanism
        # Each year the score is refreshed with probability crefo_refresh_prob.
        # A refreshed score reflects the firm's *current* financial state (post-drift).
        # Otherwise the prior year's value carries forward unchanged.
        # This is a baseline data generation feature, not an experimental impairment —
        # it reflects realistic bureau update cycles (Creditreform typically updates
        # scores annually or on event triggers, but not all firms are updated each year).
        if (runif(1) < crefo_refresh_prob) {
          crefo <- round(
            340
            + 30  * ((dte - debt_to_equity_mean) / debt_to_equity_sd)
            - 20  * ((ic  - interest_cov_mean)   / interest_cov_sd)
            - 20  * ((la  - log_assets_mean)      / log_assets_sd)
            + 40  * sector_2[i]
            + 20  * sector_3[i]
            + rnorm(1, 0, 20)
          )
          crefo <- min(max(crefo, 100), 600)
        }
        # else: crefo carries forward from previous year — staleness preserved
      }
      
      # --- 3b. Standardize using origination-year scaling parameters ---
      la_z    <- (la    - log_assets_mean)     / log_assets_sd
      dte_z   <- (dte   - debt_to_equity_mean) / debt_to_equity_sd
      ic_z    <- (ic    - interest_cov_mean)   / interest_cov_sd
      fa_z    <- (fa    - firm_age_mean)        / firm_age_sd
      crefo_z <- (crefo - crefo_mean)           / crefo_sd   # NEW
      
      # --- 3c. Compute default probability and draw outcome ---
      
      # The linear predictor: computing log-odds of default for each firm-year
      linpred <- beta0 +
        beta_log_assets     * la_z +
        beta_debt_to_equity * dte_z +
        beta_interest_cov   * ic_z +
        beta_sector2        * sector_2[i] +
        beta_sector3        * sector_3[i] +
        beta_firm_age       * fa_z +
        beta_crefo          * crefo_z +
        gastro_trend * sector_2[i] * (obs_year - vintage_center)
      
      #remove gastro_trend to see if recovery improves (or include in the model)

      
      pd_true <- plogis(linpred)
      
      # --- 3d. Seasoning period: no defaults allowed in year 1 on the books ---
      if (t <= seasoning_period) {
        default <- 0L
      } else {
        default <- rbinom(1, size = 1, prob = pd_true)
      }
      
      # Some firms will pay back their loans when due.
      repaid <- as.integer(!default & t == loan_term[i])
      
      # --- 3e. Store this firm-year row ---
      firm_rows[[t]] <- data.frame(
        firm_id          = firm_id[i],
        vintage_year     = vintage_year[i],
        obs_year         = vintage_year[i] + t - 1,
        years_on_book    = t,
        sector           = sector[i],      # raw label (1/2/3) — for diagnostics
        sector_2         = sector_2[i],
        sector_3         = sector_3[i],
        loan_term        = loan_term[i],
        trajectory       = trajectory[i],
        log_assets       = la,
        debt_to_equity   = dte,
        interest_cov     = ic,
        firm_age         = fa,
        crefo            = crefo,          
        log_assets_z     = la_z,
        debt_to_equity_z = dte_z,
        interest_cov_z   = ic_z,
        firm_age_z       = fa_z,
        crefo_z          = crefo_z,        
        pd_true          = pd_true,
        default          = default,
        repaid           = repaid
      )
      
      # Stop following the firm after default.
      if (default == 1L) {
        firm_rows <- firm_rows[seq_len(t)]
        break
      }
    }
    
    # Finish the firm, then store it in its "sub-list".
    rows[[i]] <- do.call(rbind, firm_rows)
  }
  
  # Then combine all firms to create the final panel
  panel <- do.call(rbind, rows)
  
  # ============================================================
  # PHASE 4: REPORTING
  # ============================================================
  
  n_rows      <- nrow(panel)
  n_defaulted <- sum(panel$default)
  n_repaid    <- length(unique(panel$firm_id[panel$repaid == 1]))
  
  cat("Firms simulated:       ", N, "\n")
  cat("Total firm-year rows:  ", n_rows, "\n")
  cat("Default events:        ", n_defaulted,
      sprintf("(%.1f%% of firm-years)\n", 100 * n_defaulted / n_rows))
  cat("Firms repaid:          ", n_repaid, "\n")
  cat("Obs year range:        ", min(panel$obs_year), "to", max(panel$obs_year), "\n")
  
  # --- Deterioration diagnostic ---
  defaulted_firms <- panel[panel$default == 1, "firm_id"]
  defaulter_panel <- panel[panel$firm_id %in% defaulted_firms, ]
  
  pd_trajectory <- do.call(rbind, lapply(
    split(defaulter_panel, defaulter_panel$firm_id),
    function(df) {
      df_sorted <- df[order(df$years_on_book), ]
      data.frame(pd_year1 = df_sorted$pd_true[1],
                 pd_final = df_sorted$pd_true[nrow(df_sorted)])
    }
  ))
  
  cat("\n--- Deterioration diagnostic (defaulting firms only) ---\n")
  cat("Mean pd_true year 1:  ", round(mean(pd_trajectory$pd_year1), 4), "\n")
  cat("Mean pd_true final:   ", round(mean(pd_trajectory$pd_final), 4), "\n")
  cat("Mean years observed:  ",
      round(mean(tapply(defaulter_panel$firm_id, defaulter_panel$firm_id, length)), 2), "\n")
  
  # ============================================================
  # PHASE 5: RETURN
  # ============================================================
  
  # Winsorizing bounds from clean data — used by FRS diagnostics (compute_frs)
  # to measure data quality against the clean-world reference distribution.
  # NOT used by prepare_for_modeling(), which computes its own percentile
  # cutoffs from the data it receives (practitioner approach).
  winsor_bounds <- list(
    log_assets     = c(lower = as.numeric(quantile(panel$log_assets, 0.025)),
                       upper = as.numeric(quantile(panel$log_assets, 0.975))),
    debt_to_equity = c(lower = 0,
                       upper = as.numeric(quantile(panel$debt_to_equity, 0.95))),
    interest_cov   = c(lower = 0,
                       upper = as.numeric(quantile(panel$interest_cov, 0.95))),
    firm_age       = c(lower = 0,
                       upper = as.numeric(quantile(panel$firm_age, 0.95)))
      # Note: not winsorizing Crefo as it already has hard domain bounds 
  )
  
  
  # Plausibility bounds from clean data — wider than winsorizing cutoffs.
  # These represent "can this value exist in this domain?" rather than
  # "is this value typical?" (which is what winsor_bounds/outlier captures).
  # Using 1st/99th percentiles for two-sided variables, 99th for one-sided.
  # Igl (Sec. 8.6): plausibility requires domain knowledge; these empirical
  # bounds operationalize a practitioner's sense of "what's possible."
  plausibility_bounds_empirical <- list(
    log_assets     = c(as.numeric(quantile(panel$log_assets, 0.01)),
                       as.numeric(quantile(panel$log_assets, 0.99))),
    debt_to_equity = c(0,
                       as.numeric(quantile(panel$debt_to_equity, 0.99))),
    interest_cov   = c(0,
                       as.numeric(quantile(panel$interest_cov, 0.99))),
    firm_age       = c(0,
                       as.numeric(quantile(panel$firm_age, 0.99))),
    crefo          = c(100, 600)   # hard domain bounds from Creditreform index
  )
  
  list(
    data = panel,
    scaling = list(
      log_assets     = c(mean = log_assets_mean,     sd = log_assets_sd),
      debt_to_equity = c(mean = debt_to_equity_mean, sd = debt_to_equity_sd),
      interest_cov   = c(mean = interest_cov_mean,   sd = interest_cov_sd),
      firm_age       = c(mean = firm_age_mean,        sd = firm_age_sd),
      crefo          = c(mean = crefo_mean,           sd = crefo_sd)    # NEW
    ),
    winsor_bounds = winsor_bounds,
    plausibility_bounds = plausibility_bounds_empirical,  
    true_betas = c(
      intercept       = beta0,
      log_assets      = beta_log_assets,
      debt_to_equity  = beta_debt_to_equity,
      interest_cov    = beta_interest_cov,
      sector_2        = beta_sector2,        
      sector_3        = beta_sector3,        
      firm_age        = beta_firm_age,
      crefo           = beta_crefo           
    )
  )
}


```

#### Data Generating Process DAG

*1) Simple DAG for the five indicator variables)*

Color code: 

- Gray `firm_age` : root variable drawn from a distribution,
not derived from anything else 

- Green `sector` : root variable like age , but continuous 

- Blue `log_assets`, `debt_to_equity`,
`interest_cov` : financial predictors derived from root vars and each
other 

- Orange `crefo`: Creditreform separate since it's a derived
external rating that depends on all of the above (except age)

```{r}

# ── Simple DAG: Origination-level variable dependencies ──────────────

grViz("
digraph DGP_simple {

  graph [rankdir = TB, fontname = 'Helvetica', bgcolor = 'transparent',
         nodesep = 0.8, ranksep = 0.7, margin = '0.2,0.2']
  node  [fontname = 'Helvetica', fontsize = 11, style = filled,
         shape = box, penwidth = 0.6]
  edge  [color = '#888888', penwidth = 0.7, arrowsize = 0.6]

  # ── Root variables ─────────────────────────────────────────────
  firm_age [label = 'firm_age',  fillcolor = '#F1EFE8', color = '#B4B2A9']
  sector   [label = 'sector',    fillcolor = '#E1F5EE', color = '#5DCAA5']

  # ── Derived variables ──────────────────────────────────────────
  log_assets     [label = 'log_assets',     fillcolor = '#E6F1FB', color = '#85B7EB']
  debt_to_equity [label = 'debt_to_equity', fillcolor = '#E6F1FB', color = '#85B7EB']
  interest_cov   [label = 'interest_cov',   fillcolor = '#E6F1FB', color = '#85B7EB']
  crefo          [label = 'crefo',          fillcolor = '#FAECE7', color = '#F0997B']

  # ── Edges ──────────────────────────────────────────────────────
  firm_age -> log_assets
  sector   -> log_assets
  firm_age -> debt_to_equity
  log_assets -> debt_to_equity
  debt_to_equity -> interest_cov
  log_assets -> interest_cov
  log_assets -> crefo
  debt_to_equity -> crefo
  interest_cov -> crefo
  sector -> crefo

  # ── Layout ─────────────────────────────────────────────────────
  {rank = same; firm_age; sector}
  {rank = same; log_assets; debt_to_equity; interest_cov}
}
")

```

*2) Entire DGP*

Phase one: (left) variable and dependency declaration 

Phase two: (right) panel drift

Phase three: linear predictor combining both of the above phases 

```{r}
# ── Detailed DAG of the Data Generating Process ─────────────────────────────

grViz("
digraph DGP {

  # ── Global settings ──────────────────────────────────────────────
  graph [rankdir = TB, fontname = 'Helvetica', bgcolor = 'transparent',
         nodesep = 0.6, ranksep = 0.7, margin = '0.2,0.2']
  node  [fontname = 'Helvetica', fontsize = 11, style = filled,
         shape = box, penwidth = 0.6]
  edge  [color = '#888888', penwidth = 0.7, arrowsize = 0.6]

  # ── Root variables (drawn once per firm) ────────────────────────
  firm_age   [label = 'firm_age',   fillcolor = '#F1EFE8', color = '#B4B2A9']
  sector     [label = 'sector',     fillcolor = '#E1F5EE', color = '#5DCAA5']
  trajectory [label = 'trajectory', fillcolor = '#EEEDFE', color = '#AFA9EC']

  # ── Derived at origination ─────────────────────────────────────
  log_assets     [label = 'log_assets',     fillcolor = '#E6F1FB', color = '#85B7EB']
  debt_to_equity [label = 'debt_to_equity', fillcolor = '#E6F1FB', color = '#85B7EB']
  interest_cov   [label = 'interest_cov',   fillcolor = '#E6F1FB', color = '#85B7EB']
  crefo          [label = 'crefo',          fillcolor = '#FAECE7', color = '#F0997B']

  # ── Panel drift (t > 1) ───────────────────────────────────────
  pd_curr      [label = 'pd_curr\\n(drift weight)',
                fillcolor = '#F1EFE8', color = '#B4B2A9', shape = ellipse]
  annual_update [label = 'Annual updates\\nlog_assets, d/e, int_cov',
                fillcolor = '#E6F1FB', color = '#85B7EB']
  crefo_refresh [label = 'crefo refresh\\n(p = 0.5)',
                fillcolor = '#FAECE7', color = '#F0997B']

  # ── Linear predictor → default ────────────────────────────────
  gastro_trend [label = 'gastro_trend\\nsector_2 × (year − 2015)',
               fillcolor = '#E1F5EE', color = '#5DCAA5']
  linpred      [label = 'linpred\\nall _z predictors + sector dummies',
               fillcolor = '#FAEEDA', color = '#FAC775']
  default      [label = 'default\\nBernoulli(pd_true)',
               fillcolor = '#FCEBEB', color = '#F09595']

  # ── Phase 1 edges: origination dependencies ────────────────────
  firm_age -> log_assets
  sector   -> log_assets     [label = '  ±0.3', fontsize = 9, fontcolor = '#999999']
  firm_age -> debt_to_equity [label = '  −0.02·z', fontsize = 9, fontcolor = '#999999']
  log_assets -> debt_to_equity [label = '  −0.1·z', fontsize = 9, fontcolor = '#999999']
  debt_to_equity -> interest_cov [label = '  −0.4·z', fontsize = 9, fontcolor = '#999999']
  log_assets -> interest_cov [label = '  +0.3·z', fontsize = 9, fontcolor = '#999999']

  # Crefo at origination: depends on all financials + sector
  log_assets -> crefo
  debt_to_equity -> crefo
  interest_cov -> crefo
  sector -> crefo

  # ── Phase 2 edges: drift mechanism ─────────────────────────────
  trajectory -> annual_update [label = '  drift direction', fontsize = 9, fontcolor = '#7F77DD']
  pd_curr -> annual_update    [label = '  drift intensity\\n  (deteriorating only)', fontsize = 9, fontcolor = '#999999']
  annual_update -> crefo_refresh

  # ── Phase 3 edges: linpred ─────────────────────────────────────
  annual_update -> linpred
  crefo -> linpred
  crefo_refresh -> linpred [style = dashed]
  sector -> linpred
  firm_age -> linpred       [label = '  ages +1/yr', fontsize = 9, fontcolor = '#999999']
  gastro_trend -> linpred
  sector -> gastro_trend

  linpred -> default

  # ── Layout hints ───────────────────────────────────────────────
  {rank = same; firm_age; sector; trajectory}
  {rank = same; log_assets; debt_to_equity; interest_cov}
  {rank = same; pd_curr; annual_update}
  {rank = same; gastro_trend; linpred}
}
")

```

## Simulation

This creates one synthetic dataset with 5000 businesses / around 20k
rows, and tests the default rate -- should be around 2.4%.

```{r}
######################################################################

result <- simulate_clean_data_panel(N = 50000) # running with high sample size as this is the showcase document
# sorry in advance as this is slow

###################################################################

dat    <- result$data

###################################################################

# For later: record what our PD formula is
pd_formula <- default ~ log_assets_z + debt_to_equity_z + interest_cov_z +
                             firm_age_z + crefo_z + sector_2 + sector_3


```

## Checkpoint: Model fit - perfect world

Fitting a logistic regression: this is the baseline model and should
recover the betas very well. If anything, a little worse than purely
cross-sectional data -- observations aren't independent and we add
heterogeneity with the firm trajectories

```{r}
## Model fit

fitcheck_scaled <- glm(default ~ log_assets_z + debt_to_equity_z + interest_cov_z + 
                         sector_2 + sector_3 + firm_age_z + crefo_z,
                       data = dat, family = binomial())
summary(fitcheck_scaled)

# Compare estimated vs true betas -- this is your coefficient recovery check
true_betas <- result$true_betas

recovery <- data.frame(
  true      = true_betas,
  estimated = coef(fitcheck_scaled),
  diff      = coef(fitcheck_scaled) - true_betas
)
print(round(recovery, 4))

```

## Feature reliability score on clean data

**Feature Reliability Score (FRS)** testing from Igl & Gruber's Handbuch
Datenqualität (2025). A customizable data quality measure per indicator
variable -- in our case, looking at 1) Completeness, 2) Plausibility, 3)
Outlier Score, and 4) Cross-Field Consistency for our four continuous
predictors. FRS takes the geometric mean for the listed components.

Because this is on clean data, we should expect the scores to be near 1
(perfect quality). We get marked down slightly in Plausibility and
Outliers because the FRS function is always looking to winsorize the
tails (it does not take into account that we are in the perfect world).
Crefo does not get winsorized and is by default bounded at [100, 600] so
it receives the only "perfect score".

```{r}
source("03_compute_frs.R")

frs_clean <- compute_frs(dat,
                         winsor_bounds       = result$winsor_bounds,
                         plausibility_bounds = result$plausibility_bounds)
print_frs(frs_clean)

```

------------------------------------------------------------------------

# **Stage 2: Data Impairment**

### MISSINGNESS

#### **Impairment 1: MCAR**

MCAR: missingness applied independently to each cell at random.

Every cell has the same probability of being missing regardless of any
other value in the dataset — no row-level or variable-level dependence.
This is implemented via independent rbinom draws per variable, producing
a random scatter of missingness across the data

Variables included: financial ratios and firm_age. Sector and Crefo are
excluded: sector is a simple categorical entry unlikely to be missing;
Crefo missingness is modelled separately under the MAR condition.

Mild: ~8% missing per cell, Severe: ~28% missing per cell

------------------------------------------------------------------------

```{r}
impair_mcar <- function(dat, severity = "mild") {
  
  # Mild: ~8% average missing, Severe: ~28% average missing 
  miss_prob <- switch(severity,
    mild   = 0.08,
    severe = 0.28
  )
  
  dat_impaired <- dat
  n <- nrow(dat)
  
  financial_vars <- c("log_assets", "debt_to_equity", "interest_cov", "firm_age")
  
  for (var in financial_vars) {
    
    # Independent per-cell missingness — this is what makes it MCAR
    missing_idx <- rbinom(n, 1, prob = miss_prob) == 1
    dat_impaired[[var]][missing_idx] <- NA
    dat_impaired[[paste0(var, "_z")]][missing_idx] <- NA
    
    cat(sprintf("MCAR missing rate for %s: %.3f\n", var, mean(missing_idx)))
  }
  
  attr(dat_impaired, "impairment") <- list(
    type      = "MCAR",
    severity  = severity,
    miss_prob = miss_prob,
    vars      = financial_vars
  )
  
  dat_impaired
}

```

Running the MCAR function to create "mild" and "severe" impaired data:

```{r}

cat("Mild MCAR impairment:\n")
dat_mcar_mild   <- impair_mcar(dat, severity = "mild")
cat("\nSevere MCAR impairment:\n")
dat_mcar_severe <- impair_mcar(dat, severity = "severe")


```

#### **Impairment 2: MAR**

MAR: missingness in crefo depends on firm_age (fully observed)

Mechanism: younger firms are less likely to have a Creditreform score
because they lack sufficient credit history for the bureau to have
issued a rating.

This is MAR because the missingness predictor (firm_age) is often
observed and is a different variable from the one with missing data
(crefo) — consistent with Little & Rubin (2002) and Zhang (2023, p.
116). The logistic regression method is used for realistic gradual
probability change (Zhang, 2023, p. 116). Domain justification:
confirmed with company supervisors — new market entrants frequently lack
Creditreform coverage in German B2B lending.

```{r}

impair_mar <- function(dat, severity = "mild") {
  
  # Intercepts tuned to match MCAR severity levels for fair comparison:
  # Mild: ~8% average missing, Severe: ~28% average missing
  # Negative coefficient on firm_age: younger firms → higher missingness prob
  mar_intercept <- switch(severity,
    mild   = -1.5,
    severe = 0.1
  )
  
  dat_impaired <- dat
  n <- nrow(dat)
  
  # Missingness probability decreases with firm age
  # firm_age is used on the raw scale here — older firms are progressively
  # less likely to be missing a Crefo score
  prob_missing <- plogis(mar_intercept - 0.05 * dat$firm_age)
  
  # Sanity check
  cat("Average MAR missing rate:", round(mean(prob_missing), 3), "\n")
  
  # Apply missingness
  missing_idx <- rbinom(n, 1, prob = prob_missing) == 1
  dat_impaired$crefo[missing_idx] <- NA
  dat_impaired$crefo_z[missing_idx] <- NA
  
  # Tag with metadata
  attr(dat_impaired, "impairment") <- list(
    type          = "MAR",
    severity      = severity,
    mar_intercept = mar_intercept,
    mar_coef      = -0.05,
    avg_miss_rate = mean(prob_missing),
    driven_by     = "firm_age",
    vars          = "crefo"
  )
  
  dat_impaired
}
```

Creating the impaired datasets:

```{r}

cat("Mild MAR impairment for Crefo:\n")
dat_mar_mild   <- impair_mar(dat, severity = "mild")

cat("\nSevere MAR impairment for Crefo:\n")
dat_mar_severe <- impair_mar(dat, severity = "severe")

```

#### **Impairment 3: MNAR**

MNAR: missingness in interest_cov depends on interest_cov itself.

Mechanism: firms with poor interest coverage strategically omit
reporting. This is MNAR because missingness depends on the unobserved
value of the variable itself (Little & Rubin, 2002; Zhang, 2023, p.
117).

Only allowing interest coverage to become impaired for a cleaner causal
story that is easier to validate (if we see coefficient distortion, we
know WHICH missing variable caused it, versus having multiple missing
vars x their interactions).

In real life: plausibly the most likely to be affected by MNAR because
it requires current earnings data that troubled firms are more likely to
delay or withhold.

Persistence: once a firm's interest_cov goes missing in a given year, it
remains missing for all subsequent years of that firm's life. This
reflects real reporting behavior — a firm that stops disclosing bad
financials does not voluntarily resume disclosure.

Intercepts tuned to achieve overall missing rates comparable to MCAR/MAR
after persistence propagation: Mild: ~8% of firm-years, Severe: ~28%
of firm-years

Note: because persistence amplifies the initial trigger rate, the
per-year trigger probability is set lower than the target.

```{r}

impair_mnar <- function(dat, severity = "mild") {
  
  mnar_intercept <- switch(severity,
    mild   = -3.9,
    severe = -2.5
  )
  
  dat_impaired <- dat
  n <- nrow(dat)
  
  # Step 1: Compute per-observation trigger probability
  # Negative coefficient: lower interest coverage = higher missingness prob
  prob_missing <- plogis(mnar_intercept - 1.0 * dat$interest_cov_z)
  
  cat("Average MNAR trigger probability:", round(mean(prob_missing), 3), "\n")
  
  # Step 2: Draw initial missingness triggers (per firm-year)
  trigger <- rbinom(n, 1, prob = prob_missing) == 1
  
  # Step 3: Carry forward within each firm
  # Once triggered, all subsequent years for that firm are also missing
  firms <- unique(dat_impaired$firm_id)
  
  for (f in firms) {
    idx <- which(dat_impaired$firm_id == f)
    # Rows should already be in chronological order, but sort to be safe
    idx <- idx[order(dat_impaired$years_on_book[idx])]
    
    firm_trigger <- trigger[idx]
    
    # Find the first year the trigger fires
    first_miss <- which(firm_trigger)
    
    if (length(first_miss) > 0) {
      # Everything from the first trigger onward is missing
      miss_from <- first_miss[1]
      firm_trigger[miss_from:length(firm_trigger)] <- TRUE
    }
    
    trigger[idx] <- firm_trigger
  }
  
  # Step 4: Apply missingness
  dat_impaired$interest_cov[trigger] <- NA
  dat_impaired$interest_cov_z[trigger] <- NA
  
  realized_rate <- mean(trigger)
  cat("Realized MNAR missing rate (after persistence):", round(realized_rate, 3), "\n")
  
  # Tag with metadata
  attr(dat_impaired, "impairment") <- list(
    type           = "MNAR",
    severity       = severity,
    mnar_intercept = mnar_intercept,
    mnar_coef      = -1.0,
    trigger_rate   = mean(prob_missing),
    realized_rate  = realized_rate,
    persistent     = TRUE,
    driven_by      = "interest_cov_z",
    vars           = "interest_cov"
  )
  
  dat_impaired
}

```

Creating the impaired datasets:

```{r}
cat("Mild MNAR impairment on Interest Coverage:\n")
dat_mnar_mild   <- impair_mnar(dat, severity = "mild")

cat("\nSevere MNAR impairment on Interest Coverage:\n")
dat_mnar_severe <- impair_mnar(dat, severity = "severe")

```

#### **Impairment 4: Measurement Noise**

Measurement noise: Gaussian noise added to continuous financial
predictors, creating a less-precise dataset.

Only continuous predictors are corrupted (except firm_age).

Severity defined as % of each variable's clean SD Mild: SD_noise =
+/-30% of clean SD, Severe: SD_noise = +/-100% of clean SD

```{r}

impair_noise <- function(dat, scaling, severity = "mild") {
  

  noise_sd_multiplier <- switch(severity,
    mild   = 0.30,
    severe = 1.00
  )
  
  dat_impaired <- dat
  n <- nrow(dat)
  
  # Variables to corrupt — continuous financial predictors only
  # (firm_age excluded: it drives MNAR so corrupting it would confound comparisons)
  # (sector excluded: categorical)
  noise_vars <- c("log_assets", "debt_to_equity", "interest_cov")
  
  for (v in noise_vars) {
    clean_sd   <- scaling[[v]]["sd"]
    clean_mean <- scaling[[v]]["mean"]
    noise      <- rnorm(n, mean = 0, sd = noise_sd_multiplier * clean_sd)
    dat_impaired[[v]] <- dat_impaired[[v]] + noise
    dat_impaired[[paste0(v, "_z")]] <- (dat_impaired[[v]] - clean_mean) / clean_sd
  }
  
  # Print SD inflation per variable so you can verify severity
  for (v in noise_vars) {
    cat(sprintf("Noise SD for %s: %.4f (%.0f%% of clean SD = %.4f) | realized SD of corrupted var: %.4f\n",
      v,
      noise_sd_multiplier * scaling[[v]]["sd"],
      noise_sd_multiplier * 100,
      scaling[[v]]["sd"],
      sd(dat_impaired[[v]], na.rm = TRUE)
    ))
  }
  
  # Tag with metadata
  attr(dat_impaired, "impairment") <- list(
    type               = "noise",
    severity           = severity,
    noise_sd_multiplier = noise_sd_multiplier,
    vars               = noise_vars
  )
  
  dat_impaired
}

```

Creating the impaired datasets:

```{r}
cat("Mild noise impairment:\n")
dat_noise_mild   <- impair_noise(dat, result$scaling, severity = "mild")
cat("\nSevere noise impairment:\n")
dat_noise_severe <- impair_noise(dat, result$scaling, severity = "severe")
```

#### **Impairment 5: Implausible Values**

Implausible values: inject a mix of domain-invalid, extreme outlier, and
subtly wrong values into financial predictors.

Unlike the missingness impairments, implausible values are designed to
be heterogeneous — mixing obvious violations with harder-to-detect
distortions. This reflects real-world data quality issues: decimal
shifts, unit confusion, stale data, and system migration artefacts do
not produce uniform error patterns.

Three error types per variable: 

1. Domain-invalid: clearly impossible
values (negative age, etc.) 

2. Extreme outlier: technically possible but
far outside normal range 

3. Subtle: within plausible range but wrong
(decimal shifts, swaps [numerator / denominator, or age in months not
years])

Severity controls the total % of rows corrupted per variable.

```{r}

impair_implausible <- function(dat, scaling, severity = "mild") {
  
  corrupt_rate <- switch(severity,
    mild   = 0.05,
    severe = 0.15
  )
  
  dat_impaired <- dat
  n <- nrow(dat)
  
  # ---- log_assets --------------------------------------------------------
  # Real-world errors: decimal shifts, unit confusion (thousands vs millions),
  # or placeholder values from system migrations
  n_corrupt <- floor(corrupt_rate * n)
  idx <- sample(n, n_corrupt, replace = FALSE)
  error_type <- sample(1:3, n_corrupt, replace = TRUE, prob = c(0.3, 0.4, 0.3))
  
  new_vals <- dat_impaired$log_assets[idx]  # start from real values
  new_vals[error_type == 1] <- -1           # domain-invalid: negative log assets
  new_vals[error_type == 2] <- new_vals[error_type == 2] + 
    sample(c(-1, 1), sum(error_type == 2), replace = TRUE) * 
    runif(sum(error_type == 2), 4, 6) * sd(dat$log_assets)   # extreme outlier
  new_vals[error_type == 3] <- new_vals[error_type == 3] * 10  # decimal shift
  
  dat_impaired$log_assets[idx] <- new_vals
  dat_impaired$log_assets_z <- (dat_impaired$log_assets - scaling$log_assets["mean"]) / scaling$log_assets["sd"]
  cat(sprintf("Implausible in log_assets: %d rows (%.0f%%) — %d invalid, %d extreme, %d subtle\n",
    n_corrupt, corrupt_rate * 100,
    sum(error_type == 1), sum(error_type == 2), sum(error_type == 3)))
  
  # ---- debt_to_equity ----------------------------------------------------
  # Real-world errors: sign errors, ratio computed upside down,
  # or values carried from a different reporting standard
  idx <- sample(n, n_corrupt, replace = FALSE)
  error_type <- sample(1:3, n_corrupt, replace = TRUE, prob = c(0.3, 0.4, 0.3))
  
  new_vals <- dat_impaired$debt_to_equity[idx]
  new_vals[error_type == 1] <- -abs(new_vals[error_type == 1]) - 
    runif(sum(error_type == 1), 0.5, 5)                        # negative values
  new_vals[error_type == 2] <- new_vals[error_type == 2] * 
    runif(sum(error_type == 2), 5, 20)                          # ratio blown up
  new_vals[error_type == 3] <- 1 / pmax(new_vals[error_type == 3], 0.01)  # inverted ratio
  
  dat_impaired$debt_to_equity[idx] <- new_vals
 dat_impaired$debt_to_equity_z <- (dat_impaired$debt_to_equity - scaling$debt_to_equity["mean"]) / scaling$debt_to_equity["sd"]
  cat(sprintf("Implausible in debt_to_equity: %d rows (%.0f%%) — %d invalid, %d extreme, %d subtle\n",
    n_corrupt, corrupt_rate * 100,
    sum(error_type == 1), sum(error_type == 2), sum(error_type == 3)))
  
  # ---- interest_cov ------------------------------------------------------
  # Real-world errors: sign flip, decimal shift, or stale placeholder
  idx <- sample(n, n_corrupt, replace = FALSE)
  error_type <- sample(1:3, n_corrupt, replace = TRUE, prob = c(0.3, 0.4, 0.3))
  
  new_vals <- dat_impaired$interest_cov[idx]
  new_vals[error_type == 1] <- -abs(new_vals[error_type == 1])  # sign flip: negative coverage
  new_vals[error_type == 2] <- new_vals[error_type == 2] * 
    runif(sum(error_type == 2), 10, 100)                         # decimal shift
  new_vals[error_type == 3] <- 0                                 # placeholder zero
  
  dat_impaired$interest_cov[idx] <- new_vals
  dat_impaired$interest_cov_z <- (dat_impaired$interest_cov - scaling$interest_cov["mean"]) / scaling$interest_cov["sd"]
  cat(sprintf("Implausible in interest_cov: %d rows (%.0f%%) — %d invalid, %d extreme, %d subtle\n",
    n_corrupt, corrupt_rate * 100,
    sum(error_type == 1), sum(error_type == 2), sum(error_type == 3)))
  
  # ---- firm_age ----------------------------------------------------------
  # Real-world errors: negative from date arithmetic bugs, age in months
  # instead of years, or placeholder values
  idx <- sample(n, n_corrupt, replace = FALSE)
  error_type <- sample(1:3, n_corrupt, replace = TRUE, prob = c(0.3, 0.3, 0.4))
  
  new_vals <- dat_impaired$firm_age[idx]
  new_vals[error_type == 1] <- -runif(sum(error_type == 1), 1, 20)   # negative age
  new_vals[error_type == 2] <- 999                                    # placeholder
  new_vals[error_type == 3] <- new_vals[error_type == 3] * 12         # months not years
  
  dat_impaired$firm_age[idx] <- new_vals
  cat(sprintf("Implausible in firm_age: %d rows (%.0f%%) — %d invalid, %d placeholder, %d subtle\n",
    n_corrupt, corrupt_rate * 100,
    sum(error_type == 1), sum(error_type == 2), sum(error_type == 3)))
  
  # No rescaling — implausible values are intended to distort the variable's
  # distribution. Rescaling would compress the damage.
  
  attr(dat_impaired, "impairment") <- list(
    type         = "implausible",
    severity     = severity,
    corrupt_rate = corrupt_rate,
    error_types  = c("domain_invalid", "extreme_outlier", "subtle"),
    vars         = c("log_assets", "debt_to_equity", "interest_cov", "firm_age")
  )
  
  dat_impaired
}

```

Creating the impaired datasets:

```{r}
cat("Mild implausible impairment:\n")
dat_impl_mild   <- impair_implausible(dat, result$scaling, severity = "mild")

cat("\nSevere implausible impairment:\n")
dat_impl_severe <- impair_implausible(dat, result$scaling, severity = "severe")

```

------------------------------------------------------------------------

### Function Checks Before Impairment / Modeling

```{r}

# ── DGP Checks ────────────────────────────────

# Do deteriorating firms actually have rising debt-equity and falling interest coverage over time?
det_firms <- dat[dat$trajectory == "deteriorating", ]
cat("Deteriorating firms — mean dte year 1:", 
    round(mean(det_firms$debt_to_equity[det_firms$years_on_book == 1]), 3), "\n")
cat("Deteriorating firms — mean dte year 3+:", 
    round(mean(det_firms$debt_to_equity[det_firms$years_on_book >= 3]), 3), "\n")
cat("Deteriorating firms — mean ic year 1:", 
    round(mean(det_firms$interest_cov[det_firms$years_on_book == 1]), 3), "\n")
cat("Deteriorating firms — mean ic year 3+:", 
    round(mean(det_firms$interest_cov[det_firms$years_on_book >= 3]), 3), "\n")


# ── Are sector rates / trends accurate? ────────────────────────────────

# Overall sector default rates (firm-year level)
cat("Default rates per sector (1: services / 2: gastro / 3: manufacturing):\n", 
    tapply(dat$default, dat$sector, mean), "\n",
    "Target: sector 1 ≈ 0.020, sector 2 ≈ 0.036, sector 3 ≈ 0.023\n")

# Gastro trend over time
cat("Over-time default rate increases for Gastro per year (note low base on the tails):\n", 
    round(tapply(dat$default[dat$sector == 2], dat$obs_year[dat$sector == 2], mean), 3), "\n")




```

```{r}
# ── All Impairment Functions ────────────────────────────────

# Assumes the following objects exist in your environment:
# dat                                        (clean data)
# dat_mcar_mild,    dat_mcar_severe
# dat_mnar_mild,    dat_mnar_severe
# dat_noise_mild,   dat_noise_severe
# dat_impl_mild,    dat_impl_severe


# ── 1. Missing rates (MCAR MAR & MNAR) ────────────────────────────────────────

# Within the expected column! E.g. MAR only for CREFO, MNAR only for Interest Coverage

cat("=== Missing Rates ===\n")
cat("MCAR mild:  ", round(mean(is.na(dat_mcar_mild$interest_cov)), 3),  "\n")
cat("MCAR severe:", round(mean(is.na(dat_mcar_severe$interest_cov)), 3), "\n")
cat("MAR mild:  ", round(mean(is.na(dat_mar_mild$crefo)), 3),  "\n")
cat("MAR severe:", round(mean(is.na(dat_mar_severe$crefo)), 3), "\n")
cat("MNAR mild:  ", round(mean(is.na(dat_mnar_mild$interest_cov)), 3),  "\n")
cat("MNAR severe:", round(mean(is.na(dat_mnar_severe$interest_cov)), 3), "\n")


```

```{r}

# ──  2. MNAR getting worse by int_cov quartile ─────────────────────────
cat("\n=== MNAR: Missing Rate by Interest Coverage Quartile ===\n")
ic_bins <- cut(dat$interest_cov, breaks = 4)

cat("Mild:\n")
print(round(tapply(is.na(dat_mnar_mild$interest_cov), ic_bins, mean), 3))

cat("Severe:\n")
print(round(tapply(is.na(dat_mnar_severe$interest_cov), ic_bins, mean), 3))

```

```{r}
# ── 3. Noise inflated SD ──────────────────────────────────────────────────

cat("\n=== Noise: SD Before and After ===\n")
noise_vars <- c("log_assets", "debt_to_equity", "interest_cov")

for (v in noise_vars) {
  cat(sprintf("%-20s clean SD: %.4f | mild SD: %.4f | severe SD: %.4f\n",
    v,
    sd(dat[[v]]),
    sd(dat_noise_mild[[v]]),
    sd(dat_noise_severe[[v]])
  ))
}
```

```{r}

# ── 4. Implausible values are present ─────────────────────────────────────

cat("\n=== Implausible: Range Summary of Affected Variables ===\n")
impl_vars <- c("log_assets", "debt_to_equity", "interest_cov", "firm_age")

for (v in impl_vars) {
  cat(sprintf("\n%s:\n", v))
  cat("  mild:   min =", round(min(dat_impl_mild[[v]]),  2),
              " max =", round(max(dat_impl_mild[[v]]),  2), "\n")
  cat("  severe: min =", round(min(dat_impl_severe[[v]]), 2),
              " max =", round(max(dat_impl_severe[[v]]), 2), "\n")
}

# Domain-invalid checks
cat("\nDomain violations (proportion of rows):\n")
cat("  Negative debt_to_equity — mild:", round(mean(dat_impl_mild$debt_to_equity   < 0), 3),
                                "| severe:", round(mean(dat_impl_severe$debt_to_equity  < 0), 3), "\n")
cat("  Negative firm_age       — mild:", round(mean(dat_impl_mild$firm_age          < 0), 3),
                                "| severe:", round(mean(dat_impl_severe$firm_age         < 0), 3), "\n")


```

```{r}
# ── 5a. Visual distributions: Noise ───────────────────────────────────────

par(mfrow = c(3, 3), mar = c(4, 4, 2, 1))

for (v in c("log_assets", "debt_to_equity", "interest_cov")) {
  hist(dat[[v]],             main = paste("Clean -", v),        xlab = v, col = "grey90")
  hist(dat_noise_mild[[v]],  main = paste("Noise mild -", v),   xlab = v, col = "lightblue")
  hist(dat_noise_severe[[v]], main = paste("Noise severe -", v), xlab = v, col = "steelblue")
}

# ── 5b. Visual distributions: Implausible ─────────────────────────────────

par(mfrow = c(1, 3), mar = c(5, 4, 3, 1), oma = c(0, 0, 3, 0))
for (v in c("log_assets", "debt_to_equity", "interest_cov")) {
  boxplot(
    dat[[v]], dat_impl_mild[[v]], dat_impl_severe[[v]],
    names   = c("Clean", "Mild", "Severe"),
    main    = v,
    col     = c("grey80", "lightyellow", "orange"),
    outline = TRUE,
    las     = 2
  )
}
mtext("Implausible Values Impairment: Distribution by Severity", 
      outer = TRUE, cex = 1.1, font = 2, line = 1)
```

```{r}
# ── 6. Metadata tags ──────────────────────────────────────────────────────

cat("\n=== Metadata Tags ===\n")
for (nm in c("dat_mcar_mild", "dat_mcar_severe",
             "dat_mnar_mild", "dat_mnar_severe",
             "dat_noise_mild", "dat_noise_severe",
             "dat_impl_mild", "dat_impl_severe")) {
  cat(nm, "→ type:", attr(get(nm), "impairment")$type,
      "| severity:", attr(get(nm), "impairment")$severity, "\n")
}
```

```{r}
# ── 7. Crefo Score: Missing Rates (MAR) ───────────────────────────

# crefo variable check
cat("Mean crefo (raw):     ", round(mean(dat$crefo), 1), "\n")
cat("SD crefo (raw):       ", round(sd(dat$crefo), 1), "\n")
cat("Mean crefo_z:         ", round(mean(dat$crefo_z), 4), "\n")
cat("Crefo distribution:\n")
print(quantile(dat$crefo, probs = c(0.1, 0.25, 0.5, 0.75, 0.9)))

cat("\n=== Crefo Score: Missing Rates ===\n")
cat("MAR mild:  ", round(mean(is.na(dat_mar_mild$crefo)), 3),  "\n")
cat("MAR severe:", round(mean(is.na(dat_mar_severe$crefo)), 3), "\n")

cat("\n=== MAR: Crefo Missing Rate by Firm Age Quartile ===\n")
age_bins <- cut(dat$firm_age, breaks = 4)

cat("Mild:\n")
print(round(tapply(is.na(dat_mar_mild$crefo), age_bins, mean), 3))

cat("Severe:\n")
print(round(tapply(is.na(dat_mar_severe$crefo), age_bins, mean), 3))


```

```{r}
# ── 8. Crefo Score: Distribution by Sector ────────────────────────────────

par(mfrow = c(1, 3), mar = c(4, 4, 3, 1))

for (s in 1:3) {
  label <- c("1 (Services)", "2 (Gastro)", "3 (Manufacturing)")[s]
  subset <- dat$crefo[dat$sector == s]
  hist(subset, main = paste("Crefo - Sector", label),
       xlab = "crefo", col = c("grey90", "lightblue", "salmon")[s],
       xlim = c(100, 600), breaks = 20)
  abline(v = median(subset), lty = 2, lwd = 2)
  legend("topright", legend = paste("median:", round(median(subset))),
         bty = "n", cex = 0.9)
}

cat("\n=== Crefo Median by Sector ===\n")
for (s in 1:3) {
  label <- c("services", "gastro", "manufacturing")[s]
  cat(sprintf("Sector %d (%-15s): median = %.0f, mean = %.1f\n",
    s, label,
    median(dat$crefo[dat$sector == s]),
    mean(dat$crefo[dat$sector == s])
  ))
}
```

```{r}
# ── 10. Crefo Missing Rate by Sector ───────────────────────────────────────

cat("\n=== Crefo Missing Rate by Sector (MAR Diagnostic) ===\n")

for (nm in c("dat_mar_mild", "dat_mar_severe")) {
  d <- get(nm)
  miss_by_sector <- tapply(is.na(d$crefo), d$sector, mean)
  cat(sprintf("%-14s 1 (services): %.3f | 2 (gastro): %.3f | 3 (manufacturing): %.3f\n",
    nm, miss_by_sector["1"], miss_by_sector["2"], miss_by_sector["3"]
  ))
}

```

------------------------------------------------------------------------

### Feature reliability score on impaired data

Here we want to look at both the overall FRS score (aggregate, geometric
mean) along with individual FRS components for mild and severe data.

```{r}

# Preparation Block

impaired_list <- list(
  mcar_mild    = dat_mcar_mild,
  mcar_severe  = dat_mcar_severe,
  mar_mild     = dat_mar_mild,
  mar_severe   = dat_mar_severe,
  mnar_mild    = dat_mnar_mild,
  mnar_severe  = dat_mnar_severe,
  noise_mild   = dat_noise_mild,
  noise_severe = dat_noise_severe,
  impl_mild    = dat_impl_mild,
  impl_severe  = dat_impl_severe
)

frs_all <- compute_frs_all(dat, impaired_list,
                           winsor_bounds       = result$winsor_bounds,
                           plausibility_bounds = result$plausibility_bounds)

frs_summary <- summarize_frs(frs_all)

```

**FRS Summary Tables**

| Score       | Bewertung                 |
|-------------|---------------------------|
| > 95%       | sehr hohe Datenqualität   |
| 90–95%      | gute Datenqualität        |
| 80–90%      | mittlere Datenqualität    |
| < 80%       | kritische Datenqualität   |

```{r}
# Quick look by impairment
frs_summary$by_condition

# And by variable (more as Igl & Gruber intended)
frs_summary$by_variable
```

**FRS visualization**

```{r}
# Heatmap: FRS by type x severity
ggplot(frs_summary$by_condition, aes(x = severity, y = type, fill = frs)) +
  geom_tile() +
  geom_text(aes(label = round(frs, 3)), size = 3) +
  scale_fill_gradient(low = "red", high = "green") +
  labs(title = "Overall FRS by Impairment Condition")

# Bar chart: which variables degrade most per mechanism
ggplot(frs_summary$by_variable[frs_summary$by_variable$type != "none", ],
       aes(x = var, y = frs, fill = type)) +
  geom_col(position = "dodge") +
  labs(title = "FRS by Variable and Impairment Type") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```

------------------------------------------------------------------------

Last step before modeling:

## **Stage 3: Data Preparation**

```{r}
prepare_for_modeling <- function(dat, scaling, method = "ersatzwert",
                                 model_formula = NULL, label = NULL) {
  
    # Auto-generate label from impairment metadata if not provided
  if (is.null(label)) {
    imp_meta <- attr(dat, "impairment")
    if (!is.null(imp_meta)) {
      label <- paste(imp_meta$type, imp_meta$severity)
    } else {
      label <- "unlabeled"
    }
  }
  
  # Preparation function: sits between impairment and model fitting.
  # Handles the following tasks:
  #   1. Identify extreme outliers and converts them to missings
  #   2. Winsorize data (two or one tailed, depending on variables)
  #   3. Impute NA values introduced by missingness impairment (MCAR/MAR/MNAR)
  #   4. Recompute _z columns using clean scaling parameters
  #
  # The method argument controls the imputation strategy:
  #   "ersatzwert" — RR univariate Ersatzwert imputation (default).
  #              For each variable independently: fits a univariate logistic
  #              regression (default ~ x) on non-missing cases, computes the
  #              observed default rate among missing cases (AR_MV), and solves
  #              for the x-value that would produce that default rate:
  #              Ersatzwert = (logit(AR_MV) - β₀) / β₁
  #              All missing rows receive the same replacement value per variable.
  #              Implements the RR development-phase imputation methodology.
  #   "median" — univariate median imputation (PD-neutral baseline;
  #              maps to ~z=0 after standardisation, adding approximately
  #              zero to the linear predictor). 
  #   "mice"  — multivariate imputation by chained equations (van Buuren
  #              & Groothuis-Oudshoorn, 2011). Uses predictive mean matching
  #              (pmm) as the default method, which preserves the distribution
  #              shape and avoids impossible values. Single imputation (m=1)
  #              is used because Monte Carlo replications serve the role of
  #              repeated draws. Requires the mice package.
  #
  # This function is applied identically regardless of the missingness
  # mechanism (MCAR/MAR/MNAR), reflecting practitioner norms where the
  # true mechanism is unknown and all missingness is treated the same.
  
  dat_prepared <- dat
  
  # Variables that may have been impaired with missingness
  imputable_vars <- c("log_assets", "debt_to_equity", "interest_cov", "firm_age", "crefo")
  
  # ---- Prep Part 1 & 2: Domain check + Winsorize (RR practitioner workflow) --------
  # Applied to ALL impairment types, not just implausible values.
  # A practitioner doesn't know the true impairment mechanism, so the
  # same preparation pipeline runs uniformly on every dataset.
  cat(sprintf("[%s] Data preparation:\n", label))
  
  # This function can be found in "helper functions" -- creating NAs where implausible
  dat_prepared <- domain_to_na(dat_prepared)
  
  # Make sure NAs also live in _z columns
  for (v in imputable_vars) {
    z_col <- paste0(v, "_z")
    if (z_col %in% names(dat_prepared)) {
      dat_prepared[[z_col]][is.na(dat_prepared[[v]])] <- NA
    }
  }
  
  # Winsorizing the data — percentile cutoffs computed from the data as
  # received, matching what a practitioner would do without oracle access.
  dat_prepared <- winsorize_vars(dat_prepared)
  
  # Count NAs here for diagnostics BEFORE imputation (in _z and raw cols)
  predictor_vars_z <- paste0(imputable_vars, "_z")
  predictor_vars_z <- predictor_vars_z[predictor_vars_z %in% names(dat_prepared)]
  n_missing_cells_pre <- sum(is.na(dat_prepared[, predictor_vars_z]))
  
  # Check if there are any NAs to impute
  has_missing <- any(sapply(imputable_vars, function(v) any(is.na(dat_prepared[[v]]))))
  
  
  # ---- Prep Part 3: Imputation 3 ways --------
  # ---- Ersatzwert (default), median, MICE ----
  
  if (has_missing) {
    
    if (method == "ersatzwert") {
    
      outcome_var <- all.vars(model_formula)[1]
    
      for (v in imputable_vars) {
        z_col <- paste0(v, "_z")
        if (!(z_col %in% names(dat_prepared))) next
    
        miss_idx <- is.na(dat_prepared[[z_col]])
        if (!sum(miss_idx)) next
    
        # Default rate among firms with missing values for this variable
        ar_mv <- mean(dat_prepared[[outcome_var]][miss_idx], na.rm = TRUE)
    
        # Guard against 0 or 1 default rates (would produce -Inf/Inf)
        ar_mv <- pmax(pmin(ar_mv, 1 - 1e-6), 1e-6)
    
        # Univariate logit on non-missing cases
        uni_dat <- dat_prepared[!miss_idx, c(outcome_var, z_col)]
        uni_fit <- glm(as.formula(paste(outcome_var, "~", z_col)),
                       data = uni_dat, family = binomial())
    
        beta0 <- coef(uni_fit)[1]
        beta1 <- coef(uni_fit)[2]
    
        if (is.na(beta1) || abs(beta1) < 1e-8) {
          # Degenerate case: feature has no univariate signal → impute 0
          replacement <- 0
        } else {
          replacement <- (log(ar_mv / (1 - ar_mv)) - beta0) / beta1
        }
    
        # All missing rows get the same replacement value
        dat_prepared[[z_col]][miss_idx] <- replacement
    
        # Back-transform to raw scale for consistency
        if (v %in% names(scaling) && v %in% names(dat_prepared)) {
          dat_prepared[[v]][miss_idx] <- scaling[[v]]["mean"] +
                                          replacement * scaling[[v]]["sd"]
        }
    
        cat(sprintf("  %s: %d NAs → Ersatzwert = %.4f (AR_MV = %.4f)\n",
                    z_col, sum(miss_idx), replacement, ar_mv))
      }

  cat(sprintf("[%s] RR univariate imputation complete | %d cells imputed\n",
              label, n_missing_cells_pre))
      
    } else if (method == "median") {
      
      for (var in imputable_vars) {
        n_missing <- sum(is.na(dat_prepared[[var]]))
        if (n_missing > 0) {
          fill_value <- median(dat_prepared[[var]], na.rm = TRUE)
          dat_prepared[[var]][is.na(dat_prepared[[var]])] <- fill_value
          cat(sprintf("Median imputed %d NAs in %-20s method: %s, value: %.4f\n",
            n_missing, var, method, fill_value))
        }
      }
      
    } else if (method == "mice") {
      
      require(mice)
      
      mice_vars <- imputable_vars[imputable_vars %in% names(dat_prepared)]
      mice_cols <- c(mice_vars, "sector_2", "sector_3", "default")
      mice_cols <- mice_cols[mice_cols %in% names(dat_prepared)]
      
      mice_input <- dat_prepared[, mice_cols]
      
      for (var in imputable_vars) {
        n_missing <- sum(is.na(dat_prepared[[var]]))
        if (n_missing > 0) {
          cat(sprintf("MICE imputing %d NAs in %s\n", n_missing, var))
        }
      }
      
      imp <- mice(mice_input, m = 1, method = "pmm", maxit = 5,
                  printFlag = FALSE)
      completed <- complete(imp, 1)
      
      for (var in mice_vars) {
        dat_prepared[[var]] <- completed[[var]]
      }
      
      cat("MICE imputation complete.\n")
      
    } else {
      stop(sprintf("Unknown imputation method: '%s'", method))
    }
  }
  

  # ---- Prep Part 4: Rescaling --------
  # Recompute all _z columns using clean scaling parameters.
  # Applied identically regardless of impairment type — the practitioner
  # does not know the true mechanism, so the same standardization runs
  # on every dataset.
  for (var in names(scaling)) {
    z_col <- paste0(var, "_z")
    if (z_col %in% names(dat_prepared)) {
      dat_prepared[[z_col]] <- (dat_prepared[[var]] - scaling[[var]]["mean"]) / scaling[[var]]["sd"]
    }
  }
  
  # Tag with preparation metadata
  attr(dat_prepared, "preparation") <- list(
    method = method,
    vars_checked = imputable_vars
  )
  
  dat_prepared
}



```

Fixing the data: with the default method of Ersatzwert imputation, then median, then MICE

```{r}

# ── Prepare all impaired datasets for modeling ─────────────────────────────

# Missingness conditions: impute NAs using the default ersatzwert technique + recompute z-scores

dat_mcar_mild_prepped   <- prepare_for_modeling(dat_mcar_mild, result$scaling, method = "ersatzwert", model_formula = pd_formula)
dat_mcar_severe_prepped <- prepare_for_modeling(dat_mcar_severe, result$scaling, method = "ersatzwert", model_formula = pd_formula)
dat_mar_mild_prepped    <- prepare_for_modeling(dat_mar_mild, result$scaling, method = "ersatzwert", model_formula = pd_formula)
dat_mar_severe_prepped  <- prepare_for_modeling(dat_mar_severe, result$scaling, method = "ersatzwert", model_formula = pd_formula)
dat_mnar_mild_prepped   <- prepare_for_modeling(dat_mnar_mild, result$scaling, method = "ersatzwert", model_formula = pd_formula)
dat_mnar_severe_prepped <- prepare_for_modeling(dat_mnar_severe, result$scaling, method = "ersatzwert", model_formula = pd_formula)

# Noise and implausible: implaus values are turned to NAs, then imputation

dat_noise_mild_prepped   <- prepare_for_modeling(dat_noise_mild, result$scaling, method = "ersatzwert", model_formula = pd_formula)
dat_noise_severe_prepped <- prepare_for_modeling(dat_noise_severe, result$scaling, method = "ersatzwert", model_formula = pd_formula)
dat_impl_mild_prepped    <- prepare_for_modeling(dat_impl_mild, result$scaling, method = "ersatzwert", model_formula = pd_formula)
dat_impl_severe_prepped  <- prepare_for_modeling(dat_impl_severe, result$scaling, method = "ersatzwert", model_formula = pd_formula)



```



```{r}
# Missingness conditions fixed with MEDIAN imputation: impute NAs + recompute z-scores
dat_mcar_mild_prepped_med   <- prepare_for_modeling(dat_mcar_mild, result$scaling, method = "median")
dat_mcar_severe_prepped_med <- prepare_for_modeling(dat_mcar_severe, result$scaling,  method = "median")
dat_mar_mild_prepped_med    <- prepare_for_modeling(dat_mar_mild, result$scaling, method = "median")
dat_mar_severe_prepped_med  <- prepare_for_modeling(dat_mar_severe, result$scaling,  method = "median")
dat_mnar_mild_prepped_med   <- prepare_for_modeling(dat_mnar_mild, result$scaling,  method = "median")
dat_mnar_severe_prepped_med <- prepare_for_modeling(dat_mnar_severe, result$scaling,  method = "median")
```

```{r}
# Missingness conditions fixed with MICE imputation: impute NAs + recompute z-scores
dat_mcar_mild_prepped_mice   <- prepare_for_modeling(dat_mcar_mild, result$scaling,  method = "mice")
dat_mcar_severe_prepped_mice <- prepare_for_modeling(dat_mcar_severe, result$scaling,  method = "mice")
dat_mar_mild_prepped_mice    <- prepare_for_modeling(dat_mar_mild, result$scaling, method = "mice")
dat_mar_severe_prepped_mice  <- prepare_for_modeling(dat_mar_severe, result$scaling,  method = "mice")
dat_mnar_mild_prepped_mice   <- prepare_for_modeling(dat_mnar_mild, result$scaling,  method = "mice")
dat_mnar_severe_prepped_mice <- prepare_for_modeling(dat_mnar_severe, result$scaling,  method = "mice")

```



```{r}
prep_ersatz_list <- list(
  mcar_mild    = dat_mcar_mild_prepped,
  mcar_severe  = dat_mcar_severe_prepped,
  mar_mild     = dat_mar_mild_prepped,
  mar_severe   = dat_mar_severe_prepped,
  mnar_mild    = dat_mnar_mild_prepped,
  mnar_severe  = dat_mnar_severe_prepped,
  noise_mild   = dat_noise_mild_prepped,
  noise_severe = dat_noise_severe_prepped,
  impl_mild    = dat_impl_mild_prepped,
  impl_severe  = dat_impl_severe_prepped
)
```

**Sidequest 1: what happens if we use median or mice?**
Sidequest 1: Imputation Test


```{r}
source("05_sidequest1_imputation_x_models.R")
# Now median_full and mice_full exist with all 10 conditions
```


Last thing before modeling: Comparing FRS across imputation methods for
missingness

```{r}
frs_compare <- compare_frs_imputation(
  dat, impaired_list, prep_ersatz_list,
  median_full, mice_full,
  result$winsor_bounds, result$plausibility_bounds
)
frs_compare


```

And visualizations

```{r}


# ---- Plot 1: Overall FRS by mechanism x severity x imputation method ------
# Collapse across variables first
frs_summary <- aggregate(
  frs ~ type + severity + stage,
  data = frs_compare[frs_compare$var != "MEAN" & frs_compare$type != "none", ],
  FUN  = mean
)

ggplot(frs_summary, aes(x = stage, y = frs, fill = stage)) +
  geom_col() +
  geom_text(aes(label = round(frs, 3)), vjust = -0.5, size = 2.5) +
  facet_grid(severity ~ type) +
  coord_cartesian(ylim = c(0.90, 1.005)) +
  scale_fill_manual(values = c(clean      = "grey70",
                               impaired   = "#e74c3c",
                               ersatzwert = "#3498db",
                               median     = "#f39c12",
                               mice       = "#2ecc71")) +
  labs(title    = "FRS Recovery by Imputation Method",
       subtitle = "How well does each method restore data quality?",
       y = "Mean FRS", x = NULL) +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))

# ---- Plot 2: Component breakdown for a deeper look -----------------------
# Pivot to long format for faceting by component
frs_long <- pivot_longer(
  frs_compare[frs_compare$var != "MEAN" & frs_compare$type != "none", ],
  cols      = c(completeness, plausibility, outlier, consistency),
  names_to  = "component",
  values_to = "score"
)

frs_comp_summary <- aggregate(
  score ~ type + severity + stage + component,
  data = frs_long,
  FUN  = mean
)

ggplot(frs_comp_summary[frs_comp_summary$severity == "severe", ],
       aes(x = stage, y = score, fill = stage)) +
  geom_col() +
  facet_grid(component ~ type) +
  scale_fill_manual(values = c(clean      = "grey70",
                               impaired   = "#e74c3c",
                               ersatzwert = "#3498db",
                               median     = "#f39c12",
                               mice       = "#2ecc71")) +
  labs(title    = "FRS Components: Severe Impairment",
       subtitle = "Completeness should recover; watch plausibility and outliers",
       y = "Score", x = NULL) +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))



```

FRS visualizations per variable

```{r}
# Filter to severe + drop the aggregate MEAN row
frs_var_severe <- frs_compare %>%
  filter(severity == "severe", var != "MEAN")

# Reshape: one column per stage for the heatmap facets
# frs_compare should already have columns: type, severity, var, stage, frs
# (plus the component columns)

ggplot(frs_var_severe,
       aes(x = type, y = var, fill = frs)) +
  geom_tile(colour = "white", linewidth = 0.4) +
  geom_text(aes(label = sprintf("%.3f", frs)), size = 2.5) +
  facet_wrap(~ stage, nrow = 1) +
  scale_fill_gradient2(
    low = "#e74c3c", mid = "#f9e79f", high = "#2ecc71",
    midpoint = 0.95, limits = c(0.85, 1),
    oob = scales::squish,          # clamp anything outside range
    name = "FRS"
  ) +
  labs(
    title    = "Per-Variable FRS: Severe Impairment × Imputation Method",
    subtitle = "Red = unreliable | Green = healthy | Read left→right to see recovery",
    x = "Impairment type", y = NULL
  ) +
  theme_minimal() +
  theme(
    axis.text.x  = element_text(angle = 45, hjust = 1),
    strip.text    = element_text(face = "bold"),
    panel.spacing = unit(1, "lines")
  )
```

```{r}
# Pick whichever impairment type you want to drill into
focus_type <- "implausible"   # swap to "MNAR", "noise", etc.

frs_comp_var <- frs_compare %>%
  filter(severity == "severe", type == focus_type, var != "MEAN") %>%
  pivot_longer(
    cols      = c(completeness, plausibility, outlier, consistency),
    names_to  = "component",
    values_to = "score"
  )

frs_comp_var$stage <- factor(frs_comp_var$stage,
                             levels = c("impaired", "ersatzwert", "median", "mice"))

ggplot(frs_comp_var,
       aes(x = stage, y = score, fill = stage)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = sprintf("%.3f", score)), vjust = -0.3, size = 2.2) +
  facet_grid(component ~ var) +
  coord_cartesian(ylim = c(0.80, 1.02)) +
  scale_fill_manual(values = c(impaired   = "#e74c3c",
                               ersatzwert = "#3498db",
                               median     = "#f39c12",
                               mice       = "#2ecc71")) +
  labs(
    title    = paste0("FRS Component Drill-Down: ", focus_type, " (severe)"),
    subtitle = "Which quality dimension degrades, and which method repairs it?",
    x = NULL, y = "Score"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x     = element_text(angle = 45, hjust = 1),
    strip.text       = element_text(face = "bold", size = 8),
    panel.spacing    = unit(0.5, "lines")
  )
```


------------------------------------------------------------------------

# **Stage 4: Modeling**

## 4a Logistic Regression

### Overview

The logistic regression is the workhorse model for PD estimation in
regulatory credit risk (EBA GL 2017/16). RR uses `glm()` with
`family = binomial()` -- the standard maximum-likelihood logit.

The function below fits the same formula across three stages:

1.  **Clean** -- the baseline, best-case coefficient recovery
2.  **Impaired** -- damage applied, no preparation. How bad does it get?
3.  **Prepared** -- after winsorizing, imputation, and rescaling. Did
    preparation recover anything?

For the missingness conditions (MCAR, MAR, MNAR), the raw impaired
datasets contain NAs. `glm()` handles this via complete-case analysis --
it silently drops any row with a missing value in a predictor. This is
informative: it shows what the model can recover when it simply discards
incomplete observations, and it provides a natural comparison point for
the prepared version where imputation fills those gaps. For noise and
implausible conditions, there are no NAs, so the raw fit runs on the
full distorted data with no cleanup.

Each fit stores:

-   **The fitted model object** (for later prediction, diagnostics,
    etc.)
-   **Estimated coefficients** alongside the true betas, with absolute
    differences -- this is the coefficient recovery check
-   **Number of observations used** -- important for interpreting the
    complete-case fits where glm drops NA rows
-   **Convergence flag** -- glm can occasionally fail to converge on
    heavily impaired data; we want to know when that happens rather than
    silently carrying a bad fit forward

Running it on all conditions -- raw impaired data (impaired_list) and
ersatzwert-imputed prepared data (prep_ersatz_list):

```{r}

glm_results <- fit_logistic_models(
  clean_data    = dat,
  impaired_list = impaired_list,
  prepared_list = prep_ersatz_list,
  formula       = pd_formula,
  true_betas    = result$true_betas
)

# Quick look: coefficient recovery across all conditions
print(glm_results$coef_recovery)

```

### Coefficient recovery: how far did the betas drift?

A first visual diagnostic: for each impairment condition, how close are
the recovered coefficients to the DGP truth? Three stages are shown
side-by-side: the clean baseline (triangle), raw impaired fits (circle),
and prepared fits (square). The question at each panel is: does the
square move back toward the dashed truth line compared to the circle?

Note on sample sizes: for MCAR/MAR/MNAR impaired fits, glm performs
complete-case analysis (dropping rows with any NA). This means the
impaired models for missingness conditions train on fewer observations
than the prepared versions where imputation restores completeness. The
n_obs column in coef_recovery tracks this.

```{r, fig.width=10, fig.height=7}

# ── Coefficient recovery dot plot: three-stage comparison ──────────────

# Readable ordering for the y-axis
condition_order <- c("clean",
                     "mcar_mild", "mcar_severe",
                     "mar_mild",  "mar_severe",
                     "mnar_mild", "mnar_severe",
                     "noise_mild", "noise_severe",
                     "impl_mild", "impl_severe")

glm_results$coef_recovery$condition <- factor(
  glm_results$coef_recovery$condition,
  levels = condition_order
)

glm_results$coef_recovery$stage <- factor(
  glm_results$coef_recovery$stage,
  levels = c("clean", "impaired", "prepared")
)

ggplot(glm_results$coef_recovery,
       aes(x = estimated, y = condition, colour = severity, shape = stage)) +
  geom_point(size = 2.5, position = position_dodge(width = 0.4)) +
  geom_vline(aes(xintercept = true), linetype = "dashed", colour = "grey40") +
  facet_wrap(~ term, scales = "free_x", ncol = 4) +
  scale_colour_manual(values = c(none = "grey30", mild = "#f39c12", severe = "#e74c3c")) +
  scale_shape_manual(values = c(clean = 17, impaired = 16, prepared = 15)) +
  labs(title    = "Coefficient Recovery: Logistic Regression",
       subtitle = "Dashed line = DGP truth | triangle = clean, circle = impaired, square = prepared",
       x = "Estimated coefficient", y = NULL) +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"))

```

## 4b XGBoost

### Overview

XGBoost is the second model in the simulation grid. Where logistic
regression gives us interpretable coefficients to check against the DGP
truth, XGBoost tests whether a flexible, non-parametric learner is more
or less robust to data quality impairment.

Key structural differences from glm that matter here:

-   **No formula interface.** XGBoost needs a numeric matrix
    (`xgb.DMatrix`) and a separate label vector. We extract the same
    predictor columns used in `pd_formula` and convert manually.
-   **Native NA handling.** XGBoost learns optimal split directions for
    missing values during training. This means the raw impaired fits
    (stage = "impaired") use *all* rows, including those with NAs --
    unlike glm, which drops them via complete-case analysis. This is a
    structurally different response to the same damage and part of the
    model comparison story
-   **No coefficients.** XGBoost produces feature importance
    (gain-based) rather than interpretable betas. We store importance
    tables instead of coefficient recovery, and the real model
    comparison happens downstream at the evaluation stage (predicted
    PDs, discrimination, calibration, grading).
-   **Fixed hyperparameters.** Since the research question is about data
    quality, not model tuning, we fix hyperparameters across all
    conditions. Any performance difference is then attributable to the
    data, not to tuning. The defaults below are conservative and
    standard for binary credit risk: shallow trees, moderate learning
    rate, early stopping to avoid overfitting. 

Running it:

```{r}

xgb_results <- fit_xgboost_models(
  clean_data    = dat,
  impaired_list = impaired_list,
  prepared_list = prep_ersatz_list,
  formula       = pd_formula
)

# Quick look at feature importance across conditions
print(xgb_results$importance)

```

### Feature importance: does impairment reshuffle what the model relies on?

In the clean-data fit, XGBoost should allocate gain roughly in line with
the true effect sizes in the DGP. Impairment can distort this in two
ways: a damaged variable loses signal and drops in importance, or
preparation artifacts (e.g., imputed values clustering at the median)
make a variable artificially easy to split on.

The plot below shows gain-based importance by feature, comparing clean,
impaired, and prepared stages across all conditions.

```{r, fig.width=10, fig.height=7}

# ── Feature importance comparison ──────────────────────────────────────

condition_order <- c("clean",
                     "mcar_mild", "mcar_severe",
                     "mar_mild",  "mar_severe",
                     "mnar_mild", "mnar_severe",
                     "noise_mild", "noise_severe",
                     "impl_mild", "impl_severe")

xgb_results$importance$condition <- factor(
  xgb_results$importance$condition,
  levels = condition_order
)

xgb_results$importance$stage <- factor(
  xgb_results$importance$stage,
  levels = c("clean", "impaired", "prepared")
)

ggplot(xgb_results$importance,
       aes(x = Gain, y = condition, colour = severity, shape = stage)) +
  geom_point(size = 2.5, position = position_dodge(width = 0.4)) +
  facet_wrap(~ Feature, scales = "free_x", ncol = 4) +
  scale_colour_manual(values = c(none = "grey30", mild = "#f39c12", severe = "#e74c3c")) +
  scale_shape_manual(values = c(clean = 17, impaired = 16, prepared = 15)) +
  labs(title    = "Feature Importance (Gain): XGBoost",
       subtitle = "How does impairment and preparation shift what the model relies on?",
       x = "Gain", y = NULL) +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"))

```

## 4c PD Grading

### Overview

After fitting models, the next step in a real-world PD framework is to
assign each firm to a risk grade. This maps continuous predicted PDs
onto a discrete rating scale, like how Moody's maps default
probabilities to letter grades (Aaa through C).

We use a 10-grade scale (1 = safest, 10 = riskiest) with PD boundaries
derived from Moody's one-year idealized corporate default rates. The
mapping is:

| Grade | Moody's Equivalent | PD Upper Bound |
|-------|--------------------|----------------|
| 1     | Aaa                | 0.001%         |
| 2     | Aa                 | 0.06%          |
| 3     | A                  | 0.11%          |
| 4     | Baa                | 0.30%          |
| 5     | Ba (upper)         | 0.70%          |
| 6     | Ba (lower)         | 1.50%          |
| 7     | B (upper)          | 3.00%          |
| 8     | B (lower)          | 8.00%          |
| 9     | Caa                | 15.00%         |
| 10    | Ca-C / Default     | 100.00%        |

Given the DGP's ~2% average default rate, most firms should cluster in
grades 5-8, with relatively few in the tails. This is realistic for a
B2B/SME portfolio.

The critical design choice: **bucket boundaries are fixed and model-
independent.** The same PD thresholds are applied to all conditions, all
stages, and both model types. This means a "grade 6" carries the same PD
interpretation everywhere, and any shift in grade distributions across
conditions is attributable to the model + data, not to a moving scale.

This also enables **grade migration analysis**: comparing which grade
each firm receives under clean vs. impaired vs. prepared conditions. A
firm graded 5 on clean data but 7 on impaired data has been
misclassified by the damage -- and whether preparation pulls it back
toward 5 is the key question.

```{r}

# ── PD Grading Function ───────────────────────────────────────────────

assign_pd_grades <- function(models,
                             datasets,
                             formula,
                             model_type = c("glm", "xgboost")) {

  model_type <- match.arg(model_type)
  response   <- all.vars(formula)[1]
  predictors <- all.vars(formula)[-1]

  # Moody's-aligned PD grade boundaries (upper bounds, inclusive)
  grade_breaks <- c(0, 0.00001, 0.0006, 0.0011, 0.003,
                    0.007, 0.015, 0.03, 0.08, 0.15, 1.0)
  grade_labels <- as.character(1:10)

  results <- vector("list", length(models))
  names(results) <- names(models)

  for (key in names(models)) {

    mod <- models[[key]]
    d   <- datasets[[key]]

    # Predict PDs
    if (model_type == "glm") {
      pd_pred <- predict(mod, newdata = d, type = "response")
    } else {
      X <- as.matrix(d[, predictors, drop = FALSE])
      pd_pred <- predict(mod, xgb.DMatrix(data = X))
    }

    # Assign grades using fixed Moody's-aligned boundaries
    grade <- cut(pd_pred,
                 breaks  = grade_breaks,
                 labels  = grade_labels,
                 include.lowest = TRUE,
                 right   = TRUE)

    results[[key]] <- data.frame(
      key     = key,
      firm_id = if ("firm_id" %in% names(d)) d$firm_id else seq_len(nrow(d)),
      year    = if ("year" %in% names(d)) d$year else NA,
      pd_pred = pd_pred,
      grade   = as.integer(as.character(grade)),
      default = d[[response]],
      stringsAsFactors = FALSE
    )
  }

  # Stack everything
  graded <- do.call(rbind, results)
  rownames(graded) <- NULL

  # Parse key into stage + condition
  graded$stage <- ifelse(graded$key == "clean", "clean",
                         ifelse(grepl("^imp_", graded$key), "impaired", "prepared"))
  graded$condition <- ifelse(graded$key == "clean", "clean",
                             sub("^imp_|^prep_", "", graded$key))
  graded$type <- ifelse(graded$condition == "clean", "clean",
                        sub("_mild$|_severe$", "", graded$condition))
  graded$severity <- ifelse(graded$condition == "clean", "none",
                            ifelse(grepl("severe", graded$condition), "severe", "mild"))

  graded
}

```

Running it for both models -- note that we need to pass matched lists of
models and datasets so that each model scores its own training data:

```{r}

# Build matched dataset lists (same structure as the model keys)
all_datasets <- c(
  list(clean = dat),
  setNames(
    lapply(names(impaired_list), function(nm) impaired_list[[nm]]),
    paste0("imp_", names(impaired_list))
  ),
  setNames(
    lapply(names(prep_ersatz_list), function(nm) prep_ersatz_list[[nm]]),
    paste0("prep_", names(prep_ersatz_list))
  )
)

# Grade all models
glm_grades <- assign_pd_grades(
  models     = glm_results$models,
  datasets   = all_datasets,
  formula    = pd_formula,
  model_type = "glm"
)

xgb_grades <- assign_pd_grades(
  models     = xgb_results$models,
  datasets   = all_datasets,
  formula    = pd_formula,
  model_type = "xgboost"
)

```

### Grade distributions: how does impairment shift the rating landscape?

The grade distribution under clean data is the "true" portfolio profile.
Impairment should distort this -- damaged data may push firms into the
wrong grade, either overstating risk (conservative bias) or understating
it (dangerous). The question at each panel is: does preparation restore
the clean-data distribution?

```{r, fig.width=12, fig.height=8}

# ── Grade distribution comparison: GLM ─────────────────────────────────

condition_order <- c("mcar_mild", "mcar_severe",
                     "mar_mild",  "mar_severe",
                     "mnar_mild", "mnar_severe",
                     "noise_mild", "noise_severe",
                     "impl_mild", "impl_severe")

# Get the clean distribution once
clean_dist <- glm_grades %>%
  filter(condition == "clean") %>%
  group_by(grade) %>%
  summarise(n = n(), .groups = "drop") %>%
  mutate(pct = n / sum(n),
         stage = "clean")

# Replicate clean distribution into every impaired condition
clean_repeated <- do.call(rbind, lapply(condition_order, function(cond) {
  clean_dist %>% mutate(condition = cond)
}))

# Compute impaired + prepared distributions (excluding the clean panel)
impaired_dist <- glm_grades %>%
  filter(condition != "clean") %>%
  group_by(condition, stage, grade) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(condition, stage) %>%
  mutate(pct = n / sum(n)) %>%
  ungroup()

# Combine
glm_grade_dist           <- bind_rows(clean_repeated, impaired_dist)
glm_grade_dist$condition <- factor(glm_grade_dist$condition, levels = condition_order)
glm_grade_dist$stage     <- factor(glm_grade_dist$stage, levels = c("clean", "impaired", "prepared"))

ggplot(glm_grade_dist,
       aes(x = factor(grade), y = pct, fill = stage)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  facet_wrap(~ condition, ncol = 2) +
  scale_fill_manual(values = c(clean = "grey40", impaired = "#e74c3c", prepared = "#2ecc71")) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title    = "PD Grade Distribution: Logistic Regression",
       subtitle = "Grey = clean baseline repeated in every panel for comparison",
       x = "PD Grade (1=safest, 10=riskiest)", y = "Share of firms") +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"),
        legend.position = "bottom")


# 

```

```{r, fig.width=12, fig.height=8}

# ── Grade distribution comparison: XGBoost ─────────────────────────────

condition_order <- c("mcar_mild", "mcar_severe",
                     "mar_mild",  "mar_severe",
                     "mnar_mild", "mnar_severe",
                     "noise_mild", "noise_severe",
                     "impl_mild", "impl_severe")

# Get the clean distribution once
clean_dist_xgb <- xgb_grades %>%
  filter(condition == "clean") %>%
  group_by(grade) %>%
  summarise(n = n(), .groups = "drop") %>%
  mutate(pct = n / sum(n),
         stage = "clean")

# Replicate clean distribution into every impaired condition
clean_repeated_xgb <- do.call(rbind, lapply(condition_order, function(cond) {
  clean_dist_xgb %>% mutate(condition = cond)
}))

# Compute impaired + prepared distributions (excluding the clean panel)
impaired_dist_xgb <- xgb_grades %>%
  filter(condition != "clean") %>%
  group_by(condition, stage, grade) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(condition, stage) %>%
  mutate(pct = n / sum(n)) %>%
  ungroup()

# Combine
xgb_grade_dist <- bind_rows(clean_repeated_xgb, impaired_dist_xgb)
xgb_grade_dist$condition <- factor(xgb_grade_dist$condition, levels = condition_order)
xgb_grade_dist$stage     <- factor(xgb_grade_dist$stage, levels = c("clean", "impaired", "prepared"))

ggplot(xgb_grade_dist,
       aes(x = factor(grade), y = pct, fill = stage)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  facet_wrap(~ condition, ncol = 2) +
  scale_fill_manual(values = c(clean = "grey40", impaired = "#e74c3c", prepared = "#2ecc71")) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title    = "PD Grade Distribution: XGBoost",
       subtitle = "Grey = clean baseline repeated in every panel | No NAs — XGBoost handles missing values natively",
       x = "PD Grade (1=safest, 10=riskiest)", y = "Share of firms") +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"),
        legend.position = "bottom")


```



# **Stage 5: Evaluation Metrics**

## Overview

The evaluation stage asks: how much does each impairment condition
degrade model performance, and does preparation recover it? Following
the RR model validation framework, we measure two of the three pillars:
**discrimination** (can the model rank-order firms by risk?) and
**calibration** (do predicted PDs match observed default frequencies?).

The third pillar — stability (temporal drift, population stability) — 
is outside scope. Although the simulation generates a longitudinal panel 
spanning 2005–2025, the research question concerns data quality effects 
on model reliability rather than temporal model degradation. Stability 
testing would require a dedicated development/validation time split, 
which is orthogonal to the Monte Carlo design and left for future work.

**Evaluation design.** Every model -- clean-trained, impaired-trained,
prepared-trained -- predicts on the **same clean dataset**. This is the
critical design choice: the clean data is the true risk surface (known
from the DGP), and scoring every model against it isolates the effect of
training-data damage on model performance. The clean-trained model's
metrics are the best-case ceiling; degradation from that ceiling is what
we report.

Because the research question concerns relative degradation rather than
out-of-sample prediction, all models are evaluated on the full clean
dataset. The clean-trained model's in-sample metrics serve as the upper
bound. The Monte Carlo layer (Stage 6) provides the replication and
sampling variability that a train/test split would otherwise supply.

**Metrics computed:**

-   **AUC** (Area Under the ROC Curve): probability that a randomly
    chosen defaulter receives a higher predicted PD than a randomly
    chosen non-defaulter. Computed via the Mann-Whitney U statistic --
    no external package needed.
-   **Gini coefficient**: `2 × AUC - 1`. Linearly equivalent to AUC but
    the standard reporting metric in regulatory credit risk (EBA GL
    2017/16 uses AR/Gini for discriminatory power assessment).
-   **KS statistic** (Kolmogorov-Smirnov): maximum vertical distance
    between the CDF of predicted PDs for defaulters vs. non-defaulters.
    Captures separation at the single point of maximum divergence, which
    AUC (averaging over the whole curve) can miss.
-   **Brier score**: `mean((predicted - actual)²)`. A proper scoring
    rule that jointly captures discrimination and calibration. Lower is
    better.
-   **O/E ratio** (Observed / Expected): observed default rate divided
    by mean predicted PD. A value of 1.0 indicates perfect aggregate
    calibration; below 1.0 means the model overpredicts default risk;
    above 1.0 means underprediction. This is the simplest regulatory
    calibration check.

```{r}

# ── Evaluate models: discrimination + calibration metrics ──────────────

evaluate_models <- function(models,
                            eval_data,
                            formula,
                            model_type = c("glm", "xgboost")) {

  model_type <- match.arg(model_type)
  response   <- all.vars(formula)[1]
  predictors <- all.vars(formula)[-1]

  actual <- eval_data[[response]]

  # ── Helper: AUC via Mann-Whitney U ──────────────────────────────────
  # P(score_defaulter > score_non_defaulter), with 0.5 for ties.
  # Equivalent to pROC::auc() but avoids the dependency.
  compute_auc <- function(pred, actual) {
    scores_1 <- pred[actual == 1]
    scores_0 <- pred[actual == 0]
    n1 <- length(scores_1)
    n0 <- length(scores_0)
    if (n1 == 0 || n0 == 0) return(NA_real_)
    U <- sum(vapply(scores_1, function(s) {
      sum(s > scores_0) + 0.5 * sum(s == scores_0)
    }, numeric(1)))
    U / (n1 * n0)
  }

  # ── Helper: KS statistic ────────────────────────────────────────────
  # Max |F1(t) - F0(t)| where F1 = CDF of defaulter scores,
  # F0 = CDF of non-defaulter scores.
  compute_ks <- function(pred, actual) {
    scores_1 <- pred[actual == 1]
    scores_0 <- pred[actual == 0]
    if (length(scores_1) == 0 || length(scores_0) == 0) return(NA_real_)
    cdf_1 <- ecdf(scores_1)
    cdf_0 <- ecdf(scores_0)
    all_scores <- sort(unique(pred))
    max(abs(cdf_1(all_scores) - cdf_0(all_scores)))
  }

  # ── Loop over all fitted models ─────────────────────────────────────
  results <- vector("list", length(models))
  names(results) <- names(models)

  for (key in names(models)) {

    mod <- models[[key]]

    # Predict on the SHARED clean evaluation data
    if (model_type == "glm") {
      pd_pred <- predict(mod, newdata = eval_data, type = "response")
    } else {
      X <- as.matrix(eval_data[, predictors, drop = FALSE])
      pd_pred <- predict(mod, xgb.DMatrix(data = X))
    }

    # Discrimination
    auc_val  <- compute_auc(pd_pred, actual)
    gini_val <- 2 * auc_val - 1
    ks_val   <- compute_ks(pd_pred, actual)

    # Calibration
    brier_val <- mean((pd_pred - actual)^2)
    obs_rate  <- mean(actual)
    exp_rate  <- mean(pd_pred)
    oe_ratio  <- obs_rate / exp_rate

    # Parse key into stage + condition
    stage <- ifelse(key == "clean", "clean",
                    ifelse(grepl("^imp_", key), "impaired", "prepared"))
    condition <- ifelse(key == "clean", "clean",
                        sub("^imp_|^prep_", "", key))

    results[[key]] <- data.frame(
      key        = key,
      stage      = stage,
      condition  = condition,
      model_type = model_type,
      auc        = auc_val,
      gini       = gini_val,
      ks         = ks_val,
      brier      = brier_val,
      obs_rate   = obs_rate,
      pred_rate  = exp_rate,
      oe_ratio   = oe_ratio,
      stringsAsFactors = FALSE
    )
  }

  # Stack and parse type/severity
  metrics <- do.call(rbind, results)
  rownames(metrics) <- NULL

  metrics$type <- ifelse(
    metrics$condition == "clean", "clean",
    sub("_mild$|_severe$", "", metrics$condition)
  )
  metrics$severity <- ifelse(
    metrics$condition == "clean", "none",
    ifelse(grepl("severe", metrics$condition), "severe", "mild")
  )

  metrics
}

```

Running it for both models -- every model scores on the same clean data:

```{r}

glm_metrics <- evaluate_models(
  models     = glm_results$models,
  eval_data  = dat,
  formula    = pd_formula,
  model_type = "glm"
)

xgb_metrics <- evaluate_models(
  models     = xgb_results$models,
  eval_data  = dat,
  formula    = pd_formula,
  model_type = "xgboost"
)

# Combine into one table for comparison
all_metrics <- rbind(glm_metrics, xgb_metrics)

print(all_metrics[, c("model_type", "condition", "stage",
                       "auc", "gini", "ks", "brier", "oe_ratio")])

```

### Discrimination: how much ranking power does each model lose?

The primary question: when a model is trained on damaged data, how much
worse does it rank-order firms by risk? AUC and Gini capture this
directly. The clean-trained model sets the ceiling; any drop in the
impaired or prepared rows is degradation caused by the data quality
problem.

The plot below shows Gini (more common in credit risk reporting than
AUC) by condition and stage. The clean baseline appears as a horizontal
reference line in every panel.

```{r, fig.width=10, fig.height=6}

# ── Gini degradation plot ──────────────────────────────────────────────

condition_order <- c("mcar_mild", "mcar_severe",
                     "mar_mild",  "mar_severe",
                     "mnar_mild", "mnar_severe",
                     "noise_mild", "noise_severe",
                     "impl_mild", "impl_severe")

plot_metrics <- all_metrics[all_metrics$condition != "clean", ]
plot_metrics$condition <- factor(plot_metrics$condition, levels = condition_order)
plot_metrics$stage     <- factor(plot_metrics$stage, levels = c("impaired", "prepared"))

# Clean baselines for reference lines (one per model type)
clean_gini <- all_metrics[all_metrics$condition == "clean",
                          c("model_type", "gini")]

ggplot(plot_metrics, aes(x = condition, y = gini,
                         colour = stage, shape = stage)) +
  geom_point(size = 3, position = position_dodge(width = 0.4)) +
  geom_hline(data = clean_gini,
             aes(yintercept = gini),
             linetype = "dashed", colour = "grey40") +
  facet_wrap(~ model_type, ncol = 1) +
  scale_colour_manual(values = c(impaired = "#e74c3c", prepared = "#2ecc71")) +
  scale_shape_manual(values = c(impaired = 16, prepared = 15)) +
  labs(title    = "Discriminatory Power: Gini Coefficient",
       subtitle = "Dashed line = clean-trained baseline | closer to line = better recovery",
       x = NULL, y = "Gini") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        strip.text  = element_text(face = "bold"),
        legend.position = "bottom")

```

GINI / DISCRIMINATION NOTES:

- XGBoost clean baseline (~0.675) meaningfully outperforms GLM (~0.60) due to its ability to capture nonlinearities in the DGP
GLM is surprisingly robust across missingness conditions — complete-case deletion preserves clean relationships on smaller but undistorted data
- Preparation (ersatzwert imputation) 
- Noise is where GLM shows more robustness than XGBoost relative to their own baselines, but XGBoost at its worst still beats GLM's clean ceiling
- Implausible values hit both models hardest on discrimination




### Calibration: do predicted PDs still match reality?

The O/E ratio measures aggregate calibration -- is the average predicted
PD close to the observed default rate? Since every model predicts on the
same clean data (where the true default rate is fixed), any shift in O/E
away from the clean baseline means the model's probability estimates
have been distorted by impairment.

The Brier score combines calibration and discrimination into a single
proper scoring rule. A model that ranks firms well but assigns the wrong
PD levels will have a good AUC but a poor Brier score; the gap between
the two tells you whether the problem is discrimination or calibration.

```{r, fig.width=10, fig.height=6}

# ── Brier score degradation plot ───────────────────────────────────────

clean_brier <- all_metrics[all_metrics$condition == "clean",
                           c("model_type", "brier")]

ggplot(plot_metrics, aes(x = condition, y = brier,
                         colour = stage, shape = stage)) +
  geom_point(size = 3, position = position_dodge(width = 0.4)) +
  geom_hline(data = clean_brier,
             aes(yintercept = brier),
             linetype = "dashed", colour = "grey40") +
  facet_wrap(~ model_type, ncol = 1) +
  scale_colour_manual(values = c(impaired = "#e74c3c", prepared = "#2ecc71")) +
  scale_shape_manual(values = c(impaired = 16, prepared = 15)) +
  labs(title    = "Calibration Quality: Brier Score",
       subtitle = "Dashed line = clean-trained baseline | lower is better",
       x = NULL, y = "Brier Score") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        strip.text  = element_text(face = "bold"),
        legend.position = "bottom")

```


Brier score (discrimination + calibration combined)

- XGBoost again has the better clean baseline
- GLM complete-case analysis holds calibration well until noise severe and implausible conditions, where it jumps sharply
For XGBoost, impaired and prepared are nearly identical across most conditions — degradation is driven by the information loss itself, not the repair strategy
- MAR is the most stable condition for both models, consistent with the mechanism (missingness on crefo is independent of default risk conditional on observables)
- MNAR prepared takes a hit in GLM because ersatzwert imputation pulls the missing interest_cov tail toward the center, underestimating risk for the most distressed firms



```{r, fig.width=10, fig.height=6}

# ── O/E ratio plot ─────────────────────────────────────────────────────

clean_oe <- all_metrics[all_metrics$condition == "clean",
                        c("model_type", "oe_ratio")]

ggplot(plot_metrics, aes(x = condition, y = oe_ratio,
                         colour = stage, shape = stage)) +
  geom_point(size = 3, position = position_dodge(width = 0.4)) +
  geom_hline(yintercept = 1.0, linetype = "solid", colour = "grey70", linewidth = 0.3) +
  geom_hline(data = clean_oe,
             aes(yintercept = oe_ratio),
             linetype = "dashed", colour = "grey40") +
  facet_wrap(~ model_type, ncol = 1) +
  scale_colour_manual(values = c(impaired = "#e74c3c", prepared = "#2ecc71")) +
  scale_shape_manual(values = c(impaired = 16, prepared = 15)) +
  labs(title    = "Aggregate Calibration: Observed / Expected Ratio",
       subtitle = "Solid grey = perfect calibration (1.0) | Dashed = clean baseline",
       x = NULL, y = "O/E Ratio") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        strip.text  = element_text(face = "bold"),
        legend.position = "bottom")

```

Notes O/E ratio (pure calibration):

- How can preparation undermine GLM so much?
- MNAR severe is the standout: GLM impaired underpredicts by ~15% (O/E ≈ 1.15), preparation recovers roughly half of that
Noise and implausible values look stable on O/E despite degraded Gini — this is the dangerous pattern where a supervisor checking only aggregate calibration would miss the problem



### Calibration plot: predicted vs. observed by PD grade

The grade-level calibration plot uses the same Moody's-aligned PD
buckets from Stage 4c. For each grade, we compare the average predicted
PD against the observed default frequency. A well-calibrated model
produces points along the 45-degree line. Impairment can pull points off
the diagonal -- either by mispricing individual grades or by shifting
the entire curve.

We show this for the clean baseline and all prepared conditions (the
ones a practitioner would actually deploy). Impaired conditions are
omitted for readability; the prepared-vs-clean comparison is the one
that answers the research question.

```{r, fig.width=10, fig.height=7}

# ── Calibration plot helper ────────────────────────────────────────────

build_calibration_data <- function(models,
                                   eval_data,
                                   formula,
                                   model_type = c("glm", "xgboost")) {

  model_type <- match.arg(model_type)
  response   <- all.vars(formula)[1]
  predictors <- all.vars(formula)[-1]
  actual     <- eval_data[[response]]

  # Same grade boundaries as assign_pd_grades()
  grade_breaks <- c(0, 0.00001, 0.0006, 0.0011, 0.003,
                    0.007, 0.015, 0.03, 0.08, 0.15, 1.0)
  grade_labels <- as.character(1:10)

  cal_list <- vector("list", length(models))
  names(cal_list) <- names(models)

  for (key in names(models)) {

    mod <- models[[key]]

    if (model_type == "glm") {
      pd_pred <- predict(mod, newdata = eval_data, type = "response")
    } else {
      X <- as.matrix(eval_data[, predictors, drop = FALSE])
      pd_pred <- predict(mod, xgb.DMatrix(data = X))
    }

    grade <- cut(pd_pred,
                 breaks = grade_breaks,
                 labels = grade_labels,
                 include.lowest = TRUE,
                 right = TRUE)

    tmp <- data.frame(pd_pred = pd_pred, actual = actual,
                      grade = as.integer(as.character(grade)))

    agg <- aggregate(cbind(pd_pred, actual) ~ grade, data = tmp, FUN = mean)
    counts <- aggregate(actual ~ grade, data = tmp, FUN = length)
    names(counts)[2] <- "n"
    agg <- merge(agg, counts, by = "grade")

    names(agg)[names(agg) == "pd_pred"] <- "mean_pred_pd"
    names(agg)[names(agg) == "actual"]  <- "obs_default_rate"

    # Parse key
    stage <- ifelse(key == "clean", "clean",
                    ifelse(grepl("^imp_", key), "impaired", "prepared"))
    condition <- ifelse(key == "clean", "clean",
                        sub("^imp_|^prep_", "", key))

    agg$key        <- key
    agg$stage      <- stage
    agg$condition  <- condition
    agg$model_type <- model_type

    cal_list[[key]] <- agg
  }

  cal_df <- do.call(rbind, cal_list)
  rownames(cal_df) <- NULL
  cal_df
}

```

```{r, fig.width=10, fig.height=7}

# ── Build calibration data for both model types ────────────────────────

glm_cal <- build_calibration_data(
  models     = glm_results$models,
  eval_data  = dat,
  formula    = pd_formula,
  model_type = "glm"
)

xgb_cal <- build_calibration_data(
  models     = xgb_results$models,
  eval_data  = dat,
  formula    = pd_formula,
  model_type = "xgboost"
)

all_cal <- rbind(glm_cal, xgb_cal)

# Filter to clean + prepared only (what a practitioner would deploy)
cal_plot_data <- all_cal[all_cal$stage %in% c("clean", "prepared"), ]

# For prepared conditions, parse type + severity for faceting
cal_plot_data$type <- ifelse(
  cal_plot_data$condition == "clean", "clean",
  sub("_mild$|_severe$", "", cal_plot_data$condition)
)
cal_plot_data$severity <- ifelse(
  cal_plot_data$condition == "clean", "none",
  ifelse(grepl("severe", cal_plot_data$condition), "severe", "mild")
)

# Label for faceting: one panel per impairment type
cal_plot_data$facet_label <- ifelse(
  cal_plot_data$condition == "clean", "clean",
  cal_plot_data$condition
)

condition_order <- c("clean",
                     "mcar_mild", "mcar_severe",
                     "mar_mild",  "mar_severe",
                     "mnar_mild", "mnar_severe",
                     "noise_mild", "noise_severe",
                     "impl_mild", "impl_severe")

cal_plot_data$facet_label <- factor(cal_plot_data$facet_label,
                                    levels = condition_order)

ggplot(cal_plot_data,
       aes(x = mean_pred_pd, y = obs_default_rate,
           colour = model_type, shape = stage)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
  geom_point(size = 2.5) +
  facet_wrap(~ facet_label, ncol = 3, scales = "free") +
  scale_colour_manual(values = c(glm = "#3498db", xgboost = "#e67e22")) +
  scale_shape_manual(values = c(clean = 17, prepared = 15)) +
  labs(title    = "Calibration Plot: Predicted PD vs. Observed Default Rate by Grade",
       subtitle = "Dashed line = perfect calibration | Clean (triangle) vs. Prepared (square)",
       x = "Mean Predicted PD (within grade)",
       y = "Observed Default Rate (within grade)") +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"),
        legend.position = "bottom")

```

Calibration plot (grade-level calibration) (re-do)

- Grade-level calibration mostly holds across conditions, confirming that the O/E stability isn't just offsetting errors across grades
- Main vulnerability is in the highest-PD grades, where ersatzwert imputation compresses the tail and underpredicts risk for the riskiest firms



```{r}
vars <- c("log_assets", "debt_to_equity", "interest_cov", "firm_age", "crefo")
sapply(paste0(vars, "_z"), function(v) median(dat[[v]], na.rm = TRUE))
```



### Summary: degradation at a glance

A compact summary showing all five metrics for each condition, with the
clean baseline at the top for reference. The `delta_gini` and
`delta_brier` columns show absolute change from the clean-trained model,
making it easy to scan for which conditions hurt most and whether
preparation recovers the damage.

```{r}

# ── Summary table with deltas from clean ───────────────────────────────

# Get clean baselines
clean_baselines <- all_metrics[all_metrics$condition == "clean",
                               c("model_type", "gini", "brier", "oe_ratio")]
names(clean_baselines) <- c("model_type", "clean_gini", "clean_brier", "clean_oe")

summary_metrics <- merge(all_metrics, clean_baselines, by = "model_type")
summary_metrics$delta_gini  <- summary_metrics$gini  - summary_metrics$clean_gini
summary_metrics$delta_brier <- summary_metrics$brier - summary_metrics$clean_brier
summary_metrics$delta_oe    <- summary_metrics$oe_ratio - summary_metrics$clean_oe

# Order for readability
summary_metrics$condition <- factor(
  summary_metrics$condition,
  levels = c("clean",
             "mcar_mild", "mcar_severe",
             "mar_mild",  "mar_severe",
             "mnar_mild", "mnar_severe",
             "noise_mild", "noise_severe",
             "impl_mild", "impl_severe")
)

summary_metrics <- summary_metrics[order(summary_metrics$model_type,
                                         summary_metrics$condition,
                                         summary_metrics$stage), ]

print(summary_metrics[, c("model_type", "condition", "stage",
                           "gini", "delta_gini",
                           "brier", "delta_brier",
                           "oe_ratio", "delta_oe",
                           "ks")])

```

# **Stage 6: Monte Carlo**
