---
title: "Credit Risk Data Quality Simulation"
output: html_document
date: "2026-03-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

## Part 0: Helper functions and setup --> put this in another script!!

Purpose: Rescale the impaired world with the parameters from the perfect world data
The second function is for the Noise impairment. It rescales the mean only because noise will blow up the variance, don't want to recorrect that
And we don't want any rescaling for implausible values
```{r}
rescale_with_clean_params <- function(x_impaired, scaling_params) {
    (x_impaired - scaling_params["mean"]) / scaling_params["sd"]
  }

# for random noise only
rescale_mean_only <- function(x_impaired, scaling_params) {
  x_impaired - scaling_params["mean"]
}

# implausible values doesn't get rescaled

```


## Part 1: Creating the "Perfect World" with the DGP

# Overview

step 1: creating a function that takes inputs re: how many firms to simulate, terms of the loan, the drift of the SD (random walk time series). The default values are fine. 
Also setting seed for reproduceability 

----------------------------------------------------------

Step 2: generating features (predictors) for the business world


1: root variables

a) firm age: right skewed so using gamma distribution
b) sector: at the moment it's a coin flip between 0 and 1

2: Chained regression dependencies: each one is determined by the one in front of it -- older firms tend to be larger, larger firms tend to have lower leverage

Plus -- adding randomness

These are just tests at the moment

c) log_assets: proxy for firm size, dependent on age and sector
d) debt to equity: higher leverage is more dangerous --> dependent on how big the firm is and how old it is
e) interest_cov: proxy for cash flow 


----------------------------------------------------------

Step 3: we have to standardize features for the true model (DGP)


- compute and save mean and SD of each variable
- use those parameters to standardize

We do it manually rather than just calling scale() because later, when you introduce impairments, you need to apply these exact same parameters to the impaired data — if you let the impaired data rescale itself, the impairment shifts the mean/SD and you'd be confounding two effects at once.


----------------------------------------------------------

Step 4: 
Choose the true coefficients that define the world

This is the "ground truth" -- can be confirmed from RR
All are standardized (change in log-odds per 1 SD increase)

  beta0               <- -3.5   this is the intercept -- should be whatever the default                                  rate is
  beta_log_assets     <- -0.2   larger firms less likely to default
  beta_debt_to_equity <-  0.4   higher leverage = higher risk
  beta_interest_cov   <- -0.3   higher coverage = lower risk
  beta_sector         <-  0.2   manufacturing slightly riskier
  beta_firm_age       <- -0.1   older firms slightly safer


----------------------------------------------------------

Step 5: convert linear predictors to PDs
- the "linpred" part -- weighted sum of the predictors. this is logistic regression
    - computes the log-odds of default for each borrower
- pd_true: then after: the logistic transformation so we get the true PDs and not log-odds

----------------------------------------------------------

Step 6: draw observed defaults. (default) 
For each borrower, draw a default with pd_true value from before
- borrowers can have same PDs but 2 different outcomes (default or not)


and return a list for later
Data has both raw and scaled versions of every variable
Scaling = the mean and SD of every predictor so you can apply consistently to the impaired datasets
Truebetas = ground truth coefficients

----------------------------------------------------------

Then...the end

```{r}


# Creating panel data

# have an inner loop that simulates years / data for one business
# and an outer loop that simulates for multiple businesses

##########################################

simulate_clean_data_panel <- function(N = 5000, seed = 1,
                                      min_term = 3, max_term = 7,
                                      drift_sd_log_assets     = 0.05,
                                      drift_sd_debt_to_equity = 0.03,
                                      drift_sd_interest_cov   = 0.03,
                                      drift_scale             = 0.3,
                                      seasoning_period        = 1) {
  set.seed(seed)
  
  # ============================================================
  # PHASE 1: FIRM ORIGINATION
  # ============================================================
  
  firm_age_orig <- rgamma(N, shape = 3, rate = 0.15)
  sector        <- sample(0:1, N, replace = TRUE)
  
  log_assets_orig     <- 8 + 0.03 * firm_age_orig + 0.3 * sector + rnorm(N, 0, 1) # log bc very right-skewed
  debt_to_equity_orig <- pmax(3 - 0.1 * scale(log_assets_orig) - 0.02 * scale(firm_age_orig) +
                                rgamma(N, shape = 1, rate = 1), 0)
  interest_cov_orig   <- pmax(5 - 0.4 * scale(debt_to_equity_orig) + 0.3 * scale(log_assets_orig) +
                                rnorm(N, 0, 1), 0)
  
  log_assets_mean     <- mean(log_assets_orig);     log_assets_sd     <- sd(log_assets_orig)
  debt_to_equity_mean <- mean(debt_to_equity_orig); debt_to_equity_sd <- sd(debt_to_equity_orig)
  interest_cov_mean   <- mean(interest_cov_orig);   interest_cov_sd   <- sd(interest_cov_orig)
  firm_age_mean       <- mean(firm_age_orig);       firm_age_sd       <- sd(firm_age_orig)
  
  firm_id      <- sprintf("FIRM_%05d", seq_len(N))
  vintage_year <- sample(2005:2025, N, replace = TRUE)
  loan_term    <- sample(min_term:max_term, N, replace = TRUE)
  
  # Firm trajectory type: assigned once at origination, fixed for the firm's lifetime.
  # Determines the direction of the drift mean in the annual random walk.
  # Deteriorating firms face state-dependent worsening (as before).
  # Stable firms follow a pure random walk with zero drift mean.
  # Improving firms experience a gentle drift in the opposite direction --
  # leverage falls slowly, coverage and assets improve.
  # Proportions (0.2 / 0.6 / 0.2) reflect a realistic credit portfolio
  # where most firms are broadly stable and distressed/improving firms are minorities.
  trajectory <- sample(c("deteriorating", "stable", "improving"),
                       N, replace = TRUE, prob = c(0.2, 0.6, 0.2))
  
  # ============================================================
  # PHASE 2: TRUE COEFFICIENTS
  # beta0 lowered from -3.5 to -4.2 to reduce the annual default
  # probability, giving firms runway to deteriorate before defaulting
  # rather than exiting immediately at origination.
  # ============================================================
  
  beta0               <- -4.2
  beta_log_assets     <- -0.2
  beta_debt_to_equity <-  0.4
  beta_interest_cov   <- -0.3
  beta_sector         <-  0.2
  beta_firm_age       <- -0.1
  
  # ============================================================
  # PHASE 3: PANEL EXPANSION
  # ============================================================
  
  rows <- vector("list", N)
  
  for (i in seq_len(N)) {
    
    la  <- log_assets_orig[i]
    dte <- debt_to_equity_orig[i]
    ic  <- interest_cov_orig[i]
    fa  <- firm_age_orig[i]
    
    firm_rows <- vector("list", loan_term[i])
    
    for (t in seq_len(loan_term[i])) {
      
      # --- 3a. State-dependent drift from year 2 onward ---
      if (t > 1) {
        
        # Standardize current values to compute pd_curr as deterioration weight
        la_z_curr  <- (la  - log_assets_mean)     / log_assets_sd
        dte_z_curr <- (dte - debt_to_equity_mean) / debt_to_equity_sd
        ic_z_curr  <- (ic  - interest_cov_mean)   / interest_cov_sd
        fa_z_curr  <- (fa  - firm_age_mean)        / firm_age_sd
        
        pd_curr <- plogis(beta0 +
                            beta_log_assets     * la_z_curr +
                            beta_debt_to_equity * dte_z_curr +
                            beta_interest_cov   * ic_z_curr +
                            beta_sector         * sector[i] +
                            beta_firm_age       * fa_z_curr)
        
        # Drift means are determined by trajectory type.
        # Deteriorating: state-dependent push toward worse financials, scaled by pd_curr.
        # Stable: pure random walk, drift mean = 0 for all variables.
        # Improving: gentle fixed drift in the positive direction, independent of pd_curr.
        #   A fixed (not state-dependent) improvement rate reflects firms that are
        #   genuinely deleveraging or growing, rather than just getting lucky.
        if (trajectory[i] == "deteriorating") {
          drift_mean_dte <- +drift_scale * pd_curr
          drift_mean_ic  <- -drift_scale * pd_curr
          drift_mean_la  <- -drift_scale * 0.5 * pd_curr
        } else if (trajectory[i] == "stable") {
          drift_mean_dte <- 0
          drift_mean_ic  <- 0
          drift_mean_la  <- 0
        } else {  # improving
          drift_mean_dte <- -drift_scale * 0.5   # leverage slowly falls
          drift_mean_ic  <-  drift_scale * 0.5   # coverage slowly rises
          drift_mean_la  <-  drift_scale * 0.25  # assets grow modestly
        }
        
        dte <- pmax(dte + rnorm(1, mean = drift_mean_dte, sd = drift_sd_debt_to_equity), 0)
        ic  <- pmax(ic  + rnorm(1, mean = drift_mean_ic,  sd = drift_sd_interest_cov),   0)
        la  <-      la  + rnorm(1, mean = drift_mean_la,  sd = drift_sd_log_assets)
        fa  <- fa + 1
      }
      
      # --- 3b. Standardize using origination-year scaling parameters ---
      la_z  <- (la  - log_assets_mean)     / log_assets_sd
      dte_z <- (dte - debt_to_equity_mean) / debt_to_equity_sd
      ic_z  <- (ic  - interest_cov_mean)   / interest_cov_sd
      fa_z  <- (fa  - firm_age_mean)        / firm_age_sd
      
      # --- 3c. Compute default probability and draw outcome ---
      linpred <- beta0 +
        beta_log_assets     * la_z  +
        beta_debt_to_equity * dte_z +
        beta_interest_cov   * ic_z  +
        beta_sector         * sector[i] +
        beta_firm_age       * fa_z
      pd_true <- plogis(linpred)
      
      # --- 3d. Seasoning period: no defaults allowed in early years ---
      # Firms must survive at least `seasoning_period` years before they
      # are eligible to default. This reflects real credit portfolio
      # behaviour and ensures the deterioration mechanism has time to act.
      if (t <= seasoning_period) {
        default <- 0L
      } else {
        default <- rbinom(1, size = 1, prob = pd_true)
      }
      
      repaid <- as.integer(!default & t == loan_term[i])
      
      # --- 3e. Store this firm-year row ---
      firm_rows[[t]] <- data.frame(
        firm_id          = firm_id[i],
        vintage_year     = vintage_year[i],
        obs_year         = vintage_year[i] + t - 1,
        years_on_book    = t,
        sector           = sector[i],
        loan_term        = loan_term[i],
        trajectory       = trajectory[i],        # added
        log_assets       = la,
        debt_to_equity   = dte,
        interest_cov     = ic,
        firm_age         = fa,
        log_assets_z     = la_z,
        debt_to_equity_z = dte_z,
        interest_cov_z   = ic_z,
        firm_age_z       = fa_z,
        pd_true          = pd_true,
        default          = default,
        repaid           = repaid
      )
      
      if (default == 1L) {
        firm_rows <- firm_rows[seq_len(t)]
        break
      }
    }
    
    rows[[i]] <- do.call(rbind, firm_rows)
  }
  
  panel <- do.call(rbind, rows)
  
  # ============================================================
  # PHASE 4: REPORTING
  # ============================================================
  
  n_rows      <- nrow(panel)
  n_defaulted <- sum(panel$default)
  n_repaid    <- length(unique(panel$firm_id[panel$repaid == 1]))
  
  cat("Firms simulated:       ", N, "\n")
  cat("Total firm-year rows:  ", n_rows, "\n")
  cat("Default events:        ", n_defaulted,
      sprintf("(%.1f%% of firm-years)\n", 100 * n_defaulted / n_rows))
  cat("Firms repaid:          ", n_repaid, "\n")
  cat("Obs year range:        ", min(panel$obs_year), "to", max(panel$obs_year), "\n")
  
  # --- Deterioration diagnostic ---
  # For firms that defaulted, compare mean pd_true in their first observed
  # year vs their final (default) year. A healthy gap confirms the drift
  # mechanism is doing its job before exit.
  defaulted_firms <- panel[panel$default == 1, "firm_id"]
  defaulter_panel <- panel[panel$firm_id %in% defaulted_firms, ]
  
  pd_trajectory <- do.call(rbind, lapply(
    split(defaulter_panel, defaulter_panel$firm_id),
    function(df) {
      df_sorted <- df[order(df$years_on_book), ]
      data.frame(pd_year1 = df_sorted$pd_true[1],
                 pd_final = df_sorted$pd_true[nrow(df_sorted)])
    }
  ))
  
  cat("\n--- Deterioration diagnostic (defaulting firms only) ---\n")
  cat("Mean pd_true year 1:  ", round(mean(pd_trajectory$pd_year1), 4), "\n")
  cat("Mean pd_true final:   ", round(mean(pd_trajectory$pd_final), 4), "\n")
  cat("Mean years observed:  ",
      round(mean(tapply(defaulter_panel$firm_id, defaulter_panel$firm_id, length)), 2), "\n")
  
  # ============================================================
  # PHASE 5: RETURN
  # ============================================================
  
  list(
    data = panel,
    scaling = list(
      log_assets     = c(mean = log_assets_mean,     sd = log_assets_sd),
      debt_to_equity = c(mean = debt_to_equity_mean, sd = debt_to_equity_sd),
      interest_cov   = c(mean = interest_cov_mean,   sd = interest_cov_sd),
      firm_age       = c(mean = firm_age_mean,        sd = firm_age_sd)
    ),
    true_betas = c(
      intercept       = beta0,
      log_assets      = beta_log_assets,
      debt_to_equity  = beta_debt_to_equity,
      interest_cov    = beta_interest_cov,
      sector          = beta_sector,
      firm_age        = beta_firm_age
    )
  )
}





```

## Simulation

This creates one synthetic dataset with 50k rows

And tests the default rate -- should be around 2-5%


```{r}

######################################################################

result <- simulate_clean_data_panel()

###################################################################

dat    <- result$data


mean(dat$default)  # default rate check - should be near the "truth"
```

## Model fit

Fitting a logistic regression -- this is the baseline model

The second one will fit exactly because the coefficients are also scaled



```{r}
## Model fit
# Fitting a logistic regression -- this is the baseline model
# fitcheck_raw uses raw inputs, fitcheck_scaled uses pre-scaled inputs (should recover true betas)

# Raw inputs -- coefficients not directly comparable to true betas
fitcheck_raw <- glm(default ~ log_assets + debt_to_equity + interest_cov + sector + firm_age,
            data = dat, family = binomial())
summary(fitcheck_raw)

# Pre-scaled inputs -- coefficients should recover true betas, but slightly looser because the trajectory mechanism introduces heterogeneity
#   also because we have repeated observations per firm so no more independece

fitcheck_scaled <- glm(default ~ log_assets_z + debt_to_equity_z + interest_cov_z + 
                      sector + firm_age_z,
            data = dat, family = binomial())
summary(fitcheck_scaled)

# Compare estimated vs true betas -- this is your coefficient recovery check
true_betas <- result$true_betas

recovery <- data.frame(
  true      = true_betas,
  estimated = coef(fitcheck_scaled),
  diff      = coef(fitcheck_scaled) - true_betas
)
print(round(recovery, 4))

```

*** Here: add Igl scores for the clean dataset

```{r}

```



##Part 2: Data Impairment

**Impairment 1: MCAR**

Explanation of what happens in this function:

- first: define function and its three inputs (clean data, scaling parameters, severity with default as "mild" if we don't specify
- switch: mapping mild to 0.08 and severe to .28
- make a copy of the clean dataset to work on
- identify the inancial variables that will get missingness  / apply same impairment logic to each one
- actual mcar mechanism. flip a weighted coin for each of the n firms — it returns 1 (missing) with probability miss_prob and 0 (observed) with probability 1 - miss_prob. The == 1 converts that into a logical vector of TRUE/FALSE, where TRUE means that firm's value will be set to missing. Every firm has exactly the same probability of being missing regardless of anything else about them — that's what makes it MCAR.
- where "missing idx" is true (1), replace with NAs
- for now: mean imputation with the scaling clean mean
- rescale
- metadata: attaching tags to an object without making columns in a dataframe
- return the impaired dataset


```{r}
impair_mcar <- function(dat, scaling, severity = "mild") {
  
  # Realistic placeholder severity levels
  # Mild: ~5-10% missing, Severe: ~25-30% missing
  miss_prob <- switch(severity,
    mild   = 0.08,
    severe = 0.28
  )
  
  dat_impaired <- dat
  n <- nrow(dat)
  
  # Apply MCAR to financial ratio columns only
  # (sector and firm_age are typically always observed)
  financial_vars <- c("log_assets", "debt_to_equity", "interest_cov")
  
  for (var in financial_vars) {
    
    # Introduce missingness completely at random
    missing_idx <- rbinom(n, 1, prob = miss_prob) == 1
    dat_impaired[[var]][missing_idx] <- NA
    
        # Sanity check: print realized missing rate before imputation
    cat(sprintf("MCAR missing rate for %s: %.3f\n", var, mean(missing_idx)))
    
    
    # Placeholder imputation: clean mean (will be swapped out later)
    clean_mean <- scaling[[var]]["mean"]
    dat_impaired[[var]][is.na(dat_impaired[[var]])] <- clean_mean
    
    # Rescale using clean parameters
    dat_impaired[[paste0(var, "_z")]] <- rescale_with_clean_params(
      dat_impaired[[var]],
      scaling[[var]]
    )
  }
  
  # Track what was impaired for later evaluation
  attr(dat_impaired, "impairment") <- list(
    type      = "MCAR",
    severity  = severity,
    miss_prob = miss_prob,
    vars      = financial_vars
  )
  
  dat_impaired
}


```

Running the MCAR function to create "mild" and "severe" impaired data

```{r}

dat_mcar_mild   <- impair_mcar(dat, result$scaling, severity = "mild")
dat_mcar_severe <- impair_mcar(dat, result$scaling, severity = "severe")


```


**Impairment 2: MNAR / MAR**



```{r}

impair_mnar <- function(dat, scaling, severity = "mild") {
  
  # MNAR: missingness in interest_cov depends on interest_cov itself
  # Mechanism: firms with poor interest coverage strategically omit reporting
  # This is MNAR because missingness depends on the unobserved value of the
  # variable itself (Little & Rubin, 2002; Zhang, 2023, p. 117).
  # Logistic regression method used for realistic gradual probability change.
  
  # Note: interest_cov_z (standardised using clean params) is used as the
  # predictor so the intercept/coefficient scale is interpretable and
  # comparable across severity levels.
  
  # Intercepts tuned to match MCAR/MAR severity levels for fair comparison:
  # Mild: ~8% average missing, Severe: ~28% average missing
  mnar_intercept <- switch(severity,
    mild   = -3.5,
    severe = -1.5
  )
  
  dat_impaired <- dat
  n <- nrow(dat)
  
  # Use the clean standardised interest_cov as the predictor
  # Negative coefficient: lower interest coverage = higher missingness prob
  prob_missing <- plogis(mnar_intercept - 1.0 * dat$interest_cov_z)
  
  # Sanity check
  cat("Average MNAR missing rate:", round(mean(prob_missing), 3), "\n")
  
  # Apply missingness
  missing_idx <- rbinom(n, 1, prob = prob_missing) == 1
  dat_impaired$interest_cov[missing_idx] <- NA
  
  # Placeholder imputation: clean mean (will be swapped out later)
  clean_mean <- scaling$interest_cov["mean"]
  dat_impaired$interest_cov[is.na(dat_impaired$interest_cov)] <- clean_mean
  
  # Rescale using clean parameters
  dat_impaired$interest_cov_z <- rescale_with_clean_params(
    dat_impaired$interest_cov,
    scaling$interest_cov
  )
  
  # Tag with metadata
  attr(dat_impaired, "impairment") <- list(
    type           = "MNAR",
    severity       = severity,
    mnar_intercept = mnar_intercept,
    mnar_coef      = -1.0,
    avg_miss_rate  = mean(prob_missing),
    driven_by      = "interest_cov_z",
    vars           = "interest_cov"
  )
  
  dat_impaired
}


```

```{r}

dat_mnar_mild   <- impair_mnar(dat, result$scaling, severity = "mild")
dat_mnar_severe <- impair_mnar(dat, result$scaling, severity = "severe")


```



**Part 2b: MAR**
Can delete later

Note: currently this is MAR, but if we find something like "firms with bad financial ratios don't report" then it is MNAR -- check with thecompany

Due to older firms not reporting. This is subject to change.
MNAR because whether a value is missing depends on an observed variable (here = firm_age)

Explanation of what happens in this function:
- first: taking in data, scaling from clean variables, severity
- (-3.5) / (-1.5) are the log-odds of missingness (intercept) when firm_age =  0
      (tuned until the average missing rate across simulated firms is ~8% or 28%)
- Plogis(): each firm gets its own missing probability based on its age
- Check for the average of all indiv. probabilities (8% / 28%)
- Applying missingness: rbinom flips a weighted coin for each firm using their probability made above
    Result: when something gets set to 1, then it becomes an "NA"
- Rescale with clean parameters (mean and SD), tag with metadata


```{r}

impair_mar <- function(dat, scaling, severity = "mild") {
  
  # MAR: missingness in interest_cov depends on firm_age (fully observed)
  # Mechanism: older firms less likely to report interest coverage
  # This is MAR because firm_age is always observed — the missingness
  # predictor is not the variable with missing data (Little & Rubin, 2002;
  # Zhang, 2023). Logistic regression method used for realistic gradual
  # probability change (Zhang, 2023, p. 116).
  
  # Intercepts tuned to match MCAR severity levels for fair comparison:
  # Mild: ~8% average missing, Severe: ~28% average missing
  mar_intercept <- switch(severity,
    mild   = -3.5,
    severe = -1.5
  )
  
  dat_impaired <- dat
  n <- nrow(dat)
  
  # Missingness probability increases with firm age
  prob_missing <- plogis(mar_intercept + 0.05 * dat$firm_age)
  
  # Sanity check
  cat("Average MAR missing rate:", round(mean(prob_missing), 3), "\n")
  
  # Apply missingness
  missing_idx <- rbinom(n, 1, prob = prob_missing) == 1
  dat_impaired$interest_cov[missing_idx] <- NA
  
  # Placeholder imputation: clean mean (will be swapped out later)
  clean_mean <- scaling$interest_cov["mean"]
  dat_impaired$interest_cov[is.na(dat_impaired$interest_cov)] <- clean_mean
  
  # Rescale using clean parameters
  dat_impaired$interest_cov_z <- rescale_with_clean_params(
    dat_impaired$interest_cov,
    scaling$interest_cov
  )
  
  # Tag with metadata
  attr(dat_impaired, "impairment") <- list(
    type           = "MAR",
    severity       = severity,
    mar_intercept  = mar_intercept,
    avg_miss_rate  = mean(prob_missing),
    driven_by      = "firm_age",
    vars           = "interest_cov"
  )
  
  dat_impaired
}

```

**Impairment 3: Measurement Noise**

What is happening here:
- our function takes in clean data, scaling data, and which severity (defaulted to "mild")
- Switch: maps severity string (above) to a number
- Mild at the moment: will have an extra SD 10% of the clean variable's SD -- severe is 50%. subject to change!!
- Save a new copy and save the sample size in n
- Designating only a few "noise variables" (firm size not getting noise - confirm)
- Cycling through all the variables: pull the clean SD from "scaling", generate N draws from a normal distribution centered around zero. Adding to each variable * multiplier for mild / severe
- Print impairment per variable 
- rescale using clean parameters -- isolating impairment effect from any scale effect. difference in model performance is due to nature of impairment
- metadata tag on the new dataframe



```{r}


impair_noise <- function(dat, scaling, severity = "mild") {
  
  # Measurement noise: Gaussian noise added to continuous financial predictors
  # Severity defined as % of each variable's clean SD
  # mild: SD_noise = 10% of clean SD, severe: SD_noise = 50% of clean SD
  noise_sd_multiplier <- switch(severity,
    mild   = 0.10,
    severe = 0.50
  )
  
  dat_impaired <- dat
  n <- nrow(dat)
  
  # Variables to corrupt — continuous financial predictors only
  # (firm_age excluded: it drives MNAR so corrupting it would confound comparisons)
  # (sector excluded: categorical)
  noise_vars <- c("log_assets", "debt_to_equity", "interest_cov")
  
  for (v in noise_vars) {
    clean_sd   <- scaling[[v]]["sd"]
    noise      <- rnorm(n, mean = 0, sd = noise_sd_multiplier * clean_sd)
    dat_impaired[[v]] <- dat_impaired[[v]] + noise
  }
  
  # Print SD inflation per variable so you can verify severity
  for (v in noise_vars) {
    cat(sprintf("Noise SD for %s: %.4f (%.0f%% of clean SD = %.4f)\n",
      v,
      noise_sd_multiplier * scaling[[v]]["sd"],
      noise_sd_multiplier * 100,
      scaling[[v]]["sd"]
    ))
  }
  
  # Rescale using clean parameters for MEAN ONLY
for (v in noise_vars) {
  z_name <- paste0(v, "_z")
  dat_impaired[[z_name]] <- rescale_mean_only(
    dat_impaired[[v]],
    scaling[[v]]
  )
}
  
  # Tag with metadata
  attr(dat_impaired, "impairment") <- list(
    type               = "noise",
    severity           = severity,
    noise_sd_multiplier = noise_sd_multiplier,
    vars               = noise_vars
  )
  
  dat_impaired
}



```



Creating the actual datasets

```{r}

dat_noise_mild   <- impair_noise(dat, result$scaling, severity = "mild")
dat_noise_severe <- impair_noise(dat, result$scaling, severity = "severe")
```





**Impairment 4: Implausible Values**

What is happening:
- Taking same inputs of clean data, scaling, severity
- Saving new copy, taking number of rows
- Defining what it means to be "implausible": for log_assets and interest_cov this is being 6 SD away
- Debt to equity / firm_age get invalid values (edit later)
- Inject the values: sample picks rows randomly. (floor = for a whole number). The rows get their values replaced with implausible values
- sprintf template:
    It fills in left to right:
      - `%s` ← `v` (e.g. `"log_assets"`)
      - `%d` ← `length(corrupt_idx)` (e.g. `50`)
      - `%.0f` ← `corrupt_rate * 100` (e.g. `5`)
      - `%.2f` ← `implausible_values[[v]]` (e.g. `12.34`)
- rescale (mean and sd (??))



```{r}

impair_implausible <- function(dat, scaling, severity = "mild") {
  
  # Implausible values: replace a random subset of rows with domain-invalid
  # or extreme outlier values, depending on the variable
  # Severity controls the % of rows corrupted
  # PLACEHOLDER rates — to be confirmed with company supervisors
  corrupt_rate <- switch(severity,
    mild   = 0.05,  # 5% of rows corrupted
    severe = 0.15   # 15% of rows corrupted
  )
  
  dat_impaired <- dat
  n <- nrow(dat)
  
  # --- Define implausible value per variable ---
  # PLACEHOLDER definitions — swap these out once company confirms benchmarks
  implausible_values <- list(
    log_assets     = mean(dat$log_assets) + 6 * sd(dat$log_assets),  # extreme high outlier
    debt_to_equity = -1,                                               # domain-invalid: negative
    interest_cov   = mean(dat$interest_cov) + 6 * sd(dat$interest_cov), # extreme high outlier
    firm_age       = -10                                               # domain-invalid: negative age
  )
  
  implausible_vars <- names(implausible_values)
  
  # --- Inject implausible values ---
  for (v in implausible_vars) {
    corrupt_idx <- sample(n, size = floor(corrupt_rate * n), replace = FALSE)
    dat_impaired[[v]][corrupt_idx] <- implausible_values[[v]]
    cat(sprintf("Implausible values injected in %s: %d rows (%.0f%%), value = %.2f\n",
      v,
      length(corrupt_idx),
      corrupt_rate * 100,
      implausible_values[[v]]
    ))
  }
  
  # No rescaling — implausible values are intended to distort the variable's
  # distribution. Rescaling would compress the damage and understate the impairment effect.
  
  # Tag with metadata
  attr(dat_impaired, "impairment") <- list(
    type         = "implausible",
    severity     = severity,
    corrupt_rate = corrupt_rate,
    vars         = implausible_vars
  )
  
  dat_impaired
}


```

Creating the datasets
```{r}

dat_impl_mild   <- impair_implausible(dat, result$scaling, severity = "mild")
dat_impl_severe <- impair_implausible(dat, result$scaling, severity = "severe")


```

**Doublechecking the functions**

```{r}

# ── Sanity Checks: All Impairment Functions ────────────────────────────────

# Assumes the following objects exist in your environment:
# dat                                        (clean data)
# dat_mcar_mild,    dat_mcar_severe
# dat_mnar_mild,    dat_mnar_severe
# dat_noise_mild,   dat_noise_severe
# dat_impl_mild,    dat_impl_severe
# scaling <- attr(dat, "scaling")


scaling <- attr(result, "scaling")


# ── 1. Missing rates (MCAR & MNAR) ────────────────────────────────────────

cat("=== Missing Rates ===\n")
cat("MCAR mild:  ", round(mean(is.na(dat_mcar_mild$interest_cov)), 3),  "\n")
cat("MCAR severe:", round(mean(is.na(dat_mcar_severe$interest_cov)), 3), "\n")
cat("MNAR mild:  ", round(mean(is.na(dat_mnar_mild$interest_cov)), 3),  "\n")
cat("MNAR severe:", round(mean(is.na(dat_mnar_severe$interest_cov)), 3), "\n")

# ── 2. MNAR missingness increases with firm age ────────────────────────────

cat("\n=== MNAR: Missing Rate by Firm Age Quartile ===\n")
age_bins <- cut(dat$firm_age, breaks = 4)

cat("Mild:\n")
print(round(tapply(is.na(dat_mnar_mild$interest_cov), age_bins, mean), 3))

cat("Severe:\n")
print(round(tapply(is.na(dat_mnar_severe$interest_cov), age_bins, mean), 3))

# ── 3. Noise inflated SD ──────────────────────────────────────────────────

cat("\n=== Noise: SD Before and After ===\n")
noise_vars <- c("log_assets", "debt_to_equity", "interest_cov")

for (v in noise_vars) {
  cat(sprintf("%-20s clean SD: %.4f | mild SD: %.4f | severe SD: %.4f\n",
    v,
    sd(dat[[v]]),
    sd(dat_noise_mild[[v]]),
    sd(dat_noise_severe[[v]])
  ))
}

# ── 4. Implausible values are present ─────────────────────────────────────

cat("\n=== Implausible: Summary of Affected Variables ===\n")
impl_vars <- c("log_assets", "debt_to_equity", "interest_cov", "firm_age")

for (v in impl_vars) {
  cat(sprintf("\n%s:\n", v))
  cat("  mild max:  ", round(max(dat_impl_mild[[v]]),  2), "\n")
  cat("  severe max:", round(max(dat_impl_severe[[v]]), 2), "\n")
}

# Domain-invalid checks
cat("\nNegative debt_to_equity — mild: ",
  round(mean(dat_impl_mild$debt_to_equity < 0), 3),  
  "| severe:", 
  round(mean(dat_impl_severe$debt_to_equity < 0), 3), "\n")

cat("Negative firm_age — mild: ",
  round(mean(dat_impl_mild$firm_age < 0), 3),  
  "| severe:", 
  round(mean(dat_impl_severe$firm_age < 0), 3), "\n")

# ── 5. Visual distributions ───────────────────────────────────────────────

par(mfrow = c(3, 3), mar = c(4, 4, 2, 1))

for (v in c("log_assets", "debt_to_equity", "interest_cov")) {
  hist(dat[[v]],            main = paste("Clean -", v),       xlab = v, col = "grey90")
  hist(dat_noise_mild[[v]], main = paste("Noise mild -", v),  xlab = v, col = "lightblue")
  hist(dat_impl_mild[[v]],  main = paste("Implaus mild -", v), xlab = v, col = "lightyellow")
}

# ── 6. Metadata tags ──────────────────────────────────────────────────────

cat("\n=== Metadata Tags ===\n")
for (nm in c("dat_mcar_mild", "dat_mcar_severe",
             "dat_mnar_mild", "dat_mnar_severe",
             "dat_noise_mild", "dat_noise_severe",
             "dat_impl_mild", "dat_impl_severe")) {
  cat(nm, "→ type:", attr(get(nm), "impairment")$type,
      "| severity:", attr(get(nm), "impairment")$severity, "\n")
}

```


