---
title: "PerfectWorld Sim V2 RMD"
output: html_document
date: "2026-03-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("C:/Users/HaightD/OneDrive - Kantar/Documents/Masters/THESIS/CODE")

```

## Explanation of the simple DGP

step 1: creating a function that takes 2 inputs -- how many borrowers / rows to simulate. Default is 5000
Also setting seed for reproduceability 

----------------------------------------------------------

Step 2: generating features (predictors) for the business world


1: root variables

a) firm age: right skewed so using gamma distribution
b) sector: at the moment it's a coin flip between 0 and 1

2: Chained regression dependencies: each one is determined by the one in front of it -- older firms tend to be larger, larger firms tend to have lower leverage

Plus -- adding randomness

c) log_assets: 
d) debt to equity:
e) interest_cov:



----------------------------------------------------------

Step 3: we have to standardize features for the true model (DGP)


- compute and save mean and SD of each variable
- use those parameters to standardize

We do it manually rather than just calling scale() because later, when you introduce impairments, you need to apply these exact same parameters to the impaired data — if you let the impaired data rescale itself, the impairment shifts the mean/SD and you'd be confounding two effects at once.


----------------------------------------------------------

Step 4: 
Choose the true coefficients that define the world

This is the "ground truth" -- can be confirmed from RR
All are standardized (change in log-odds per 1 SD increase)

  beta0               <- -3.5   this is the intercept -- should be whatever the default                                  rate is
  beta_log_assets     <- -0.2   larger firms less likely to default
  beta_debt_to_equity <-  0.4   higher leverage = higher risk
  beta_interest_cov   <- -0.3   higher coverage = lower risk
  beta_sector         <-  0.2   manufacturing slightly riskier
  beta_firm_age       <- -0.1   older firms slightly safer


----------------------------------------------------------

Step 5: convert linear predictors to PDs
- the "linpred" part -- weighted sum of the predictors
    - computes the log-odds of default for each borrower
- pd_true: then after: the logistic transformation so we get the true PDs and not log-odds

----------------------------------------------------------

Step 6: draw observed defaults. (default) 
For each borrower, draw a default with pd_true value from before
- borrowers can have same PDs but 2 different outcomes (default or not)


and return a list for later
Data has both raw and scaled versions of every variable
Scaling = the mean and SD of every predictor so you can apply consistently to the impaired datasets
Truebetas = ground truth coefficients

----------------------------------------------------------

Then...the end

```{r}

simulate_clean_data_simple <- function(N = 50000, seed = 1) {
  set.seed(seed)

  # --- 1. Root variables (generated independently) ---
  firm_age <- rgamma(N, shape = 3, rate = 0.15)   # ~5-40 years, right-skewed
  sector   <- sample(0:1, N, replace = TRUE)        # 0 = services, 1 = manufacturing

  # --- 2. Chained regression dependencies ---
  log_assets     <- 8 + 0.03 * firm_age + 0.3 * sector + rnorm(N, 0, 1)
  debt_to_equity <- pmax(3 - 0.1 * scale(log_assets) - 0.02 * scale(firm_age) +
                         rgamma(N, shape = 1, rate = 1), 0)
  interest_cov   <- pmax(5 - 0.4 * scale(debt_to_equity) + 0.3 * scale(log_assets) +
                         rnorm(N, 0, 1), 0)

  # --- 3. Standardize using clean data parameters (save for later impairment use) ---
  log_assets_mean     <- mean(log_assets);     log_assets_sd     <- sd(log_assets)
  debt_to_equity_mean <- mean(debt_to_equity); debt_to_equity_sd <- sd(debt_to_equity)
  interest_cov_mean   <- mean(interest_cov);   interest_cov_sd   <- sd(interest_cov)
  firm_age_mean       <- mean(firm_age);       firm_age_sd       <- sd(firm_age)

  log_assets_z     <- (log_assets     - log_assets_mean)     / log_assets_sd
  debt_to_equity_z <- (debt_to_equity - debt_to_equity_mean) / debt_to_equity_sd
  interest_cov_z   <- (interest_cov   - interest_cov_mean)   / interest_cov_sd
  firm_age_z       <- (firm_age       - firm_age_mean)       / firm_age_sd

  # --- 4. True coefficients (all in SD units) ---
  beta0               <- -3.5   # this is the default rate we should recover
  beta_log_assets     <- -0.2   # larger firms less likely to default
  beta_debt_to_equity <-  0.4   # higher leverage = higher risk
  beta_interest_cov   <- -0.3   # higher coverage = lower risk
  beta_sector         <-  0.2   # manufacturing slightly riskier
  beta_firm_age       <- -0.1   # older firms slightly safer

  # --- 5. Default outcome ---
  linpred <- beta0 +
    beta_log_assets     * log_assets_z +
    beta_debt_to_equity * debt_to_equity_z +
    beta_interest_cov   * interest_cov_z +
    beta_sector         * sector +
    beta_firm_age       * firm_age_z

  pd_true <- 1 / (1 + exp(-linpred))
  default <- rbinom(N, size = 1, prob = pd_true)

  cat("Default rate:", round(mean(default), 3), "\n")

  # --- 6. Return data and scaling parameters ---
  list(
    data = data.frame(
      default, sector,
      log_assets, debt_to_equity, interest_cov, firm_age,   # raw
      log_assets_z, debt_to_equity_z, interest_cov_z, firm_age_z,  # scaled
      pd_true
    ),
    scaling = list(
      log_assets     = c(mean = log_assets_mean,     sd = log_assets_sd),
      debt_to_equity = c(mean = debt_to_equity_mean, sd = debt_to_equity_sd),
      interest_cov   = c(mean = interest_cov_mean,   sd = interest_cov_sd),
      firm_age       = c(mean = firm_age_mean,        sd = firm_age_sd)
    ),
    true_betas = c(
      intercept       = beta0,
      log_assets      = beta_log_assets,
      debt_to_equity  = beta_debt_to_equity,
      interest_cov    = beta_interest_cov,
      sector          = beta_sector,
      firm_age        = beta_firm_age
    )
  )
}



```

## Simulation

This creates one synthetic dataset with 50k rows

And tests the default rate -- should be around 2-5%


```{r}

dat <- simulate_clean_data_simple(N = 50000, seed = 123)

result <- simulate_clean_data_simple(N = 50000, seed = 123)
dat    <- result$data

mean(dat$default)  # default rate sanity check
```

## Model fit

Fitting a logistic regression -- this is the baseline model

The second one will fit exactly because the coefficients are also scaled



```{r}

## Model fit
# Fitting a logistic regression -- this is the baseline model
# fit1 uses raw inputs, fit2 uses pre-scaled inputs (should recover true betas)

# Raw inputs -- coefficients not directly comparable to true betas
fit1 <- glm(default ~ log_assets + debt_to_equity + interest_cov + sector + firm_age,
            data = dat, family = binomial())
summary(fit1)

# Pre-scaled inputs -- coefficients should recover true betas closely
fit2 <- glm(default ~ log_assets_z + debt_to_equity_z + interest_cov_z + 
                      sector + firm_age_z,
            data = dat, family = binomial())
summary(fit2)

# Compare estimated vs true betas -- this is your coefficient recovery check
true_betas <- result$true_betas

recovery <- data.frame(
  true      = true_betas,
  estimated = coef(fit2),
  diff      = coef(fit2) - true_betas
)
print(round(recovery, 4))

